---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
ggplot2::theme_set(ggplot2::theme_bw())
```

# epiprocess

## TODO: Condense these paragraphs

The [`{epiprocess}`](https://cmu-delphi.github.io/epiprocess/) package works
with epidemiological time series data to provide situational
awareness, processing, and transformations in preparation for modeling, and
version-faithful model backtesting. It contains:

- `epi_df`, a class for working with epidemiological time series data which
behaves like a tibble (and can be manipulated with
[`{dplyr}`](https://dplyr.tidyverse.org/)-esque "verbs") but with some
additional structure;
- `epi_archive`, a class for working with the version history of such time series data;
- sample epidemiological data in these formats;

This package is provided by the Delphi group at Carnegie Mellon University. The
Delphi group provides many tools also hosts the Delphi Epidata API, which provides access to a wide
range of epidemiological data sets, including COVID-19 data, flu data, and more.
This package is designed to work seamlessly with the data in the Delphi Epidata
API, which can be accessed using the `epidatr` package.

It is part of a broader suite of packages that includes
[`{epipredict}`](https://cmu-delphi.github.io/epipredict/),
[`{epidatr}`](https://cmu-delphi.github.io/epidatr/),
[`{rtestim}`](https://dajmcdon.github.io/rtestim/), and
[`{epidatasets}`](https://cmu-delphi.github.io/epidatasets/), for accessing,
analyzing, and forecasting epidemiological time series data. We have expanded
documentation and demonstrations for some of these packages available in an
online "book" format [here](https://cmu-delphi.github.io/delphi-tooling-book/).

## Motivation

[`{epiprocess}`](https://cmu-delphi.github.io/epiprocess/) and
[`{epipredict}`](https://cmu-delphi.github.io/epipredict/) are designed to lower
the barrier to entry and implementation cost for epidemiological time series
analysis and forecasting. Epidemiologists and forecasting groups repeatedly and
separately have had to rush to implement this type of functionality in a much
more ad hoc manner; we are trying to save such effort in the future by providing
well-documented, tested, and general packages that can be called for many common
tasks instead.

## Installation

To install:

```{r, eval=FALSE}
# Stable version
pak::pkg_install("cmu-delphi/epiprocess@main")

# Dev version
pak::pkg_install("cmu-delphi/epiprocess@dev")
```

The package is not yet on CRAN.

## Usage

Once `epiprocess` and `epidatr` are installed, you can use the following code to
get started:

```{r, results=FALSE, warning=FALSE, message=FALSE}
library(epiprocess)
library(epidatr)
library(dplyr)
library(magrittr)
```

Get COVID-19 confirmed cumulative case data from JHU CSSE for California,
Florida, New York, and Texas, from March 1, 2020 to January 31, 2022

```{r cache=TRUE}
df <- pub_covidcast(
  source = "jhu-csse",
  signals = "confirmed_cumulative_num",
  geo_type = "state",
  time_type = "day",
  geo_values = "ca,fl,ny,tx",
  time_values = epirange(20200301, 20220131),
) %>%
  select(geo_value, time_value, cases_cumulative = value)
df
```

Convert the data to an epi_df object and sort by geo_value and time_value. You
can work with the epi_df object like a tibble using dplyr

```{r}
edf <- df %>%
  as_epi_df() %>%
  arrange_canonical() %>%
  group_by(geo_value) %>%
  mutate(cases_daily = cases_cumulative - lag(cases_cumulative, default = 0))
edf
```

Autoplot the confirmed daily cases for each geo_value

```{r}
edf %>%
  autoplot(cases_cumulative)
```

Compute the 7 day moving average of the confirmed daily cases for each geo_value

```{r}
edf %>%
  group_by(geo_value) %>%
  epi_slide_mean(cases_daily, .window_size = 7, na.rm = TRUE)
```

Compute the growth rate of the confirmed cumulative cases for each geo_value

```{r}
edf %>%
  group_by(geo_value) %>%
  mutate(cases_growth = growth_rate(x = time_value, y = cases_cumulative, method = "rel_change", h = 7))
```

Detect outliers in the growth rate of the confirmed cumulative cases for each

```{r}
edf %>%
  group_by(geo_value) %>%
  mutate(outlier_info = detect_outlr(x = time_value, y = cases_daily)) %>%
  ungroup()
```

Add a column to the epi_df object with the daily deaths for each geo_value and
compute the correlations between cases and deaths for each geo_value

```{r cache=TRUE}
df <- pub_covidcast(
  source = "jhu-csse",
  signals = "deaths_incidence_num",
  geo_type = "state",
  time_type = "day",
  geo_values = "ca,fl,ny,tx",
  time_values = epirange(20200301, 20220131),
) %>%
  select(geo_value, time_value, deaths_daily = value) %>%
  as_epi_df() %>%
  arrange_canonical()
edf <- inner_join(edf, df, by = c("geo_value", "time_value"))
edf %>%
  group_by(geo_value) %>%
  epi_slide_mean(deaths_daily, .window_size = 7, na.rm = TRUE) %>%
  epi_cor(cases_daily, deaths_daily)
```
