---
title: "Advanced Plotting with epiprocess"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced Plotting with epiprocess}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
source("_common.R")
```

The `{epiprocess}` package provides advanced visualization capabilities 
for both `epi_df` and `epi_archive` objects.

```{r, message=FALSE, warning=FALSE}
library(epidatr)
library(epiprocess)
library(dplyr)
library(tidyr)
library(ggplot2)
set.seed(123)
```

## Static Plotting with Many Keys

When plotting an `epi_df` with a large number of key combinations (e.g., all US 
states), `{epiprocess}` automatically subsamples the keys to keep the plot readable 
and performant.
You can control this behavior with the `.max_keys` argument (e.g., set it to `Inf`
 to display all keys).

```{r, eval=TRUE}
# Fetching HHS admissions for all states
df <- pub_covidcast(
  source = "hhs",
  signals = "confirmed_admissions_covid_1d",
  geo_type = "state",
  time_type = "day",
  geo_values = "*",
  time_values = epirange(20220901, 20230101)
) %>%
  as_epi_df()

# Static plot with many keys (subsampling will occur)
autoplot(
  df, value
)
```

## Heatmaps

You can visualize `epi_df` objects as heatmaps using `plot_heatmap()`.
This is particularly useful for visualizing values across many locations 
simultaneously.

```{r, eval=TRUE}
# Static heatmap for a single variable
plot_heatmap(df, value)
```

## Interactive Plotting

For interactive exploration of data, you can set `.interactive = TRUE` in 
`autoplot()`.
This returns a `plotly` widget, allowing you to hover over data points or 
isolate certain lines.
```{r, eval=TRUE, fig.height=2}
# Interactive plot for a single variable
autoplot(
  df, value,
  .interactive = TRUE
)

# Interactive plot with facets
autoplot(
  df, value,
  .interactive = TRUE,
  .facet_by = "geo_value",
  .facet_filter = geo_value %in% c("ar", "ca", "ny", "tx", "il", "mi")
)
```

### Interactive Dropdowns

When you have many facets, they can take up a lot of space.
Setting `.facet_to_dropdown = TRUE` along with `.interactive = TRUE` 
collapses the facets into a single plot with a dropdown menu.

```{r, eval=TRUE, fig.height=2}
# Dropdown for geographic locations
autoplot(df, value,
  .interactive = TRUE,
  .facet_by = "geo_value",
  .facet_to_dropdown = TRUE
)
```

You can also combine faceting and dropdowns for more complex data exploration.
For example, dropdown by one key and facet by another:

```{r, eval=requireNamespace("plotly", quietly=TRUE), message=FALSE, warning=FALSE, fig.height=2}
# Synthetic data with multiple keys
df2 <- expand.grid(
  geo_value = c("ca", "ny", "tx"),
  time_value = as.Date("2023-01-01") + 0:30,
  other = c("source_a", "source_b")
) %>%
  dplyr::mutate(cases = rnorm(dplyr::n())) %>%
  as_epi_df(other_keys = "other")

# Dropdown by 'other' key, facet by 'geo_value'
autoplot(df2, cases,
  .facet_by = "geo_value",
  .facet_to_dropdown = TRUE,
  .interactive = TRUE
)
```

## Plotting `epi_archive` Objects

`epi_archive` objects can also be visualized with `autoplot()`.
This allows you to visualize how data appeared at different points in time (versions).

```{r, eval=TRUE}
# Fetch HHS admissions with version history
df_versions <- pub_covidcast(
  source = "hhs",
  signals = "confirmed_admissions_covid_1d",
  geo_type = "state",
  time_type = "day",
  geo_values = "ar,ia,ca,ny,tx,ks,co",
  time_values = epirange(20221001, 20230101),
  issues = epirange(20221001, 20230101)
) %>%
  rename(version = issue) %>%
  as_epi_archive()

# Static epi_archive plot showing all captured versions
autoplot(df_versions, value)

# Static epi_archive plot with filtering
autoplot(
  df_versions, value,
  .facet_filter = geo_value %in% c("ar", "ca")
)
```

Interactive plots for archives also support the dropdown feature, which can 
be useful for selecting a specific location or key.

```{r, eval=TRUE, fig.height=2}
# Archive plot with dropdown for states
autoplot(
  df_versions, value,
  .interactive = TRUE,
  .facet_to_dropdown = TRUE,
  .facet_filter = geo_value %in% c("ar", "ca")
)
```
