---
title: Work with archive objects and data revisions
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Work with archive objects and data revisions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

In addition to the `epi_df` data structure, which we have been working with all
along in these vignettes, the `epiprocess` package has a companion structure
called `epi_archive`. In comparison to an `epi_df` object, which can be seen as
storing a single snapshot of a data set with the most up-to-date signal values
as of some given time, an `epi_archive` object stores the full version history
of a data set. Many signals of interest for epidemiological tracking are subject
to revision (some more than others), and paying attention to data revisions can
be important for all sorts of downstream data analysis and modeling tasks.

This vignette walks through working with `epi_archive` objects and demonstrates
some of their key functionality. We'll work with a signal on the percentage of
doctor's visits with CLI (COVID-like illness) computed from medical insurance
claims, available through the [COVIDcast
API](https://cmu-delphi.github.io/delphi-epidata/api/covidcast.html). This
signal is subject to very heavy and regular revision; you can read more about it
on its [API documentation
page](https://cmu-delphi.github.io/delphi-epidata/api/covidcast-signals/doctor-visits.html).

```{r, message = FALSE, warning = FALSE, eval=FALSE}
library(delphi.epidata)
library(epiprocess)
library(data.table)
library(dplyr)

dv <- covidcast(
  data_source = "doctor-visits",
  signals = "smoothed_adj_cli",
  time_type = "day",
  geo_type = "state",
  time_values = epirange(20200601, 20211201),
  geo_values = "ca,fl,ny,tx",
  issues = epirange(20200601, 20211201)
) %>% fetch_tbl()

```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(delphi.epidata)
library(epiprocess)
library(data.table)
library(dplyr)
```

## Getting data into `epi_archive` format

An <code><a href="../reference/epi_archive.html">epi_archive</a></code> object
can be constructed from a data frame, data table, or tibble, provided that it
has (at least) the following columns:

* `geo_value`: the geographic value associated with each row of measurements.
* `time_value`: the time value associated with each row of measurements.
* `version`: the time value specifying the version for each row of measurements.
  For example, if in a given row the `version` is January 15, 2022 and
  `time_value` is January 14, 2022, then this row contains the measurements of
  the data for January 14, 2022 that were available one day later.

As we can see from the above, the data frame returned by
`delphi.epidata::covidcast()` has the columns required for the `epi_archive`
format, with `issue` playing the role of `version`. We can now use
`as_epi_archive()` to bring it into `epi_archive` format. For removal of
redundant version updates in `as_epi_archive` using compactify, please refer
to the compactify vignette.

```{r, eval=FALSE}
x <- dv %>%
  select(geo_value, time_value, version = issue, percent_cli = value) %>%
  as_epi_archive(compactify=TRUE)

class(x)
print(x)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
x <- archive_cases_dv_subset$DT %>%
  select(geo_value, time_value, version , percent_cli)  %>%
  as_epi_archive(compactify=TRUE)

class(x)
print(x)
```

An `epi_archive` is special kind of class called an R6 class. Its primary field
is a data table `DT`, which is of class `data.table` (from the `data.table`
package), and has columns `geo_value`, `time_value`, `version`, as well as any
number of additional columns.

```{r}
class(x$DT)
head(x$DT)
```

The variables `geo_value`, `time_value`, `version` serve as **key variables**
for the data table, as well as any other specified in the metadata (described
below). There can only be a single row per unique combination of key variables,
and therefore the key variables are critical for figuring out how to generate a
snapshot of data from the archive, as of a given version (also described below).
   
```{r, error=TRUE}
key(x$DT)
```
 
In general, the last version of each observation is carried forward (LOCF) to
fill in data between recorded versions. **A word of caution:** R6 objects,
unlike most other objects in R, have reference semantics. An important
consequence of this is that objects are not copied when modified.
   
```{r}
original_value <- x$DT$percent_cli[1]
y <- x # This DOES NOT make a copy of x
y$DT$percent_cli[1] = 0
head(y$DT)
head(x$DT) 
x$DT$percent_cli[1] <- original_value
```

To make a copy, we can use the `clone()` method for an R6 class, as in `y <-
x$clone()`. You can read more about reference semantics in Hadley Wickham's
[Advanced R](https://adv-r.hadley.nz/r6.html#r6-semantics) book.

## Some details on metadata

The following pieces of metadata are included as fields in an `epi_archive`
object: 

* `geo_type`: the type for the geo values.
* `time_type`: the type for the time values.
* `additional_metadata`: list of additional metadata for the data archive.

Metadata for an `epi_archive` object `x` can be accessed (and altered) directly,
as in `x$geo_type` or `x$time_type`, etc. Just like `as_epi_df()`, the function
`as_epi_archive()` attempts to guess metadata fields when an `epi_archive`
object is instantiated, if they are not explicitly specified in the function
call (as it did in the case above).

Note that `compactify` is **NOT** metadata and is an argument passed when creating
the dataset, without being stored in the end:

```{r,message=FALSE}
# `dt` here is taken from the tests
as_epi_archive(archive_cases_dv_subset$DT,compactify=TRUE)$geo_type # "state"
as_epi_archive(archive_cases_dv_subset$DT,compactify=TRUE)$compactify # NULL
```

## Producing snapshots in `epi_df` form

A key method of an `epi_archive` class is `as_of()`, which generates a snapshot
of the archive in `epi_df` format. This represents the most up-to-date values of
the signal variables as of a given version. This can be accessed via `x$as_of()`
for an `epi_archive` object `x`, but the package also provides a simple wrapper 
function `epix_as_of()` since this is likely a more familiar interface for users
not familiar with R6 (or object-oriented programming).

```{r}
x_snapshot <- epix_as_of(x, max_version = as.Date("2021-06-01"))
class(x_snapshot)
head(x_snapshot)
max(x_snapshot$time_value)
attributes(x_snapshot)$metadata$as_of
```

We can see that the max time value in the `epi_df` object `x_snapshot` that was 
generated from the archive is May 29, 2021, even though the specified version
date was June 1, 2021. From this we can infer that the doctor's visits signal
was 2 days latent on June 1. Also, we can see that the metadata in the `epi_df`
object has the version date recorded in the `as_of` field.

By default, using the maximum of the `version` column in the underlying data table in an
`epi_archive` object itself generates a snapshot of the latest values of signal
variables in the entire archive. The `epix_as_of()` function issues a warning in
this case, since updates to the current version may still come in at a later 
point in time, due to various reasons, such as synchronization issues.

```{r}
x_latest <- epix_as_of(x, max_version = max(x$DT$version))
```

Below, we pull several snapshots from the archive, spaced one month apart. We
overlay the corresponding signal curves as colored lines, with the version dates
marked by dotted vertical lines, and draw the latest curve in black (from the 
latest snapshot `x_latest` that the archive can provide).

```{r, fig.width = 8, fig.height = 7}
library(purrr)
library(ggplot2)
theme_set(theme_bw())

self_max = max(x$DT$version)
versions = seq(as.Date("2020-06-01"), self_max - 1, by = "1 month")
snapshots <- map_dfr(versions, function(v) { 
  epix_as_of(x, max_version = v) %>% mutate(version = v)
}) %>%
  bind_rows(x_latest %>% mutate(version = self_max)) %>%
  mutate(latest = version == self_max)

ggplot(snapshots %>% filter(!latest),
            aes(x = time_value, y = percent_cli)) +  
  geom_line(aes(color = factor(version))) + 
  geom_vline(aes(color = factor(version), xintercept = version), lty = 2) +
  facet_wrap(~ geo_value, scales = "free_y", ncol = 1) +
  scale_x_date(minor_breaks = "month", date_labels = "%b %y") +
  labs(x = "Date", y = "% of doctor's visits with CLI") + 
  theme(legend.position = "none") +
  geom_line(data = snapshots %>% filter(latest),
               aes(x = time_value, y = percent_cli), 
               inherit.aes = FALSE, color = "black")
```

We can see some interesting and highly nontrivial revision behavior: at some
points in time the provisional data snapshots grossly underestimate the latest
curve (look in particular at Florida close to the end of 2021), and at others
they overestimate it (both states towards the beginning of 2021), though not 
quite as dramatically. Modeling the revision process, which is often called
*backfill modeling*, is an important statistical problem in it of itself.

<!-- todo: refer to some project/code? perhaps we should think about writing a
 function in `epiprocess` or even a separate package? -->

## Merging `epi_archive` objects 

Now we demonstrate how to merge two `epi_archive` objects together, e.g., so
that grabbing data from multiple sources as of a particular version can be
performed with a single `as_of` call. The `epi_archive` class provides a method
`merge()` precisely for this purpose. The wrapper function is called
`epix_merge()`; this wrapper avoids mutating its inputs, while `x$merge` will
mutate `x`. Below we merge the working `epi_archive` of versioned percentage CLI
from outpatient visits to another one of versioned COVID-19 case reporting data,
which we fetch the from the [COVIDcast
API](https://cmu-delphi.github.io/delphi-epidata/api/covidcast.html/), on the
rate scale (counts per 100,000 people in the population).

When merging archives, unless the archives have identical data release patterns,
`NA`s can be introduced in the non-key variables for a few reasons:
- to represent the "value" of an observation before its initial release (when we
  need to pair it with additional observations from the other archive that have
  been released)
- to represent the "value" of an observation that has no recorded versions at
  all (in the same sort of situation)
- if requested via `sync="na"`, to represent potential update data that we do
  not yet have access to (e.g., due to encountering issues while attempting to
  download the currently available version data for one of the archives, but not
  the other).

```{r, message = FALSE, warning = FALSE,eval=FALSE}
y <- covidcast(
  data_source = "jhu-csse",
  signals = "confirmed_7dav_incidence_prop",
  time_type = "day",
  geo_type = "state",
  time_values = epirange(20200601, 20211201),
  geo_values = "ca,fl,ny,tx",
  issues = epirange(20200601, 20211201)
) %>%
  fetch_tbl() %>%
  select(geo_value, time_value, version = issue, case_rate_7d_av = value) %>%
  as_epi_archive(compactify=TRUE)

x$epix_merge(y)
print(x)
head(x$DT)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
x <- archive_cases_dv_subset
print(x)
head(x$DT)
```

Importantly, see that `x$merge` mutated `x` to hold the result of the merge. We
could also have used `xy = epix_merge(x,y)` to avoid mutating `x`. See the
documentation for either for more detailed descriptions of what mutation,
pointer aliasing, and pointer reseating is possible.

## Sliding version-aware computations
    
Lastly, we demonstrate another key method of the `epi_archive` class, which is
the `slide()` method. It works just like `epi_slide()` does for an `epi_df`
object, but with one key difference: it performs version-aware computations.
That is, for the computation at any given reference time t, it only uses **data
that would have been available as of t**. The wrapper function is called
`epix_slide()`; again, this is just for convenience/familiarity---and its 
interface is purposely designed mirror that of `epi_slide()` for `epi_df`
objects.

For the demonstration, we'll revisit the forecasting example from the [slide 
vignette](https://cmu-delphi.github.io/epiprocess/articles/slide.html), and now
we'll build a forecaster that uses properly-versioned data (that would have been
available in real-time) to forecast future COVID-19 case rates from current and
past COVID-19 case rates, as well as current and past values of the outpatient 
CLI signal from medical claims. We'll extend the `prob_ar()` function from the
slide vignette to accomodate exogenous variables in the autoregressive model,
which is often referred to as an ARX model. 

```{r}
prob_arx <- function(x, y, lags = c(0, 7, 14), ahead = 7, min_train_window = 20,
                     lower_level = 0.05, upper_level = 0.95, symmetrize = TRUE, 
                     intercept = FALSE, nonneg = TRUE) {   
  # Return NA if insufficient training data
  if (length(y) < min_train_window + max(lags) + ahead) {
    return(data.frame(point = NA, lower = NA, upper = NA))
  }
  
  # Useful transformations
  if (!missing(x)) x <- data.frame(x, y)
  else x <- data.frame(y)
  if (!is.list(lags)) lags <- list(lags)
  lags = rep(lags, length.out = ncol(x))
  
  # Build features and response for the AR model, and then fit it
  dat <- do.call(
    data.frame, 
    unlist( # Below we loop through and build the lagged features
      purrr::map(1:ncol(x), function(i) { 
        purrr::map(lags[[i]], function(j) lag(x[,i], n = j))
      }),
      recursive = FALSE))
  names(dat) = paste0("x", 1:ncol(dat))
  if (intercept) dat$x0 = rep(1, nrow(dat))
  dat$y <- lead(y, n = ahead) 
  obj <- lm(y ~ . + 0, data = dat)
  
  # Use LOCF to fill NAs in the latest feature values, make a prediction
  setDT(dat) 
  setnafill(dat, type = "locf")
  point <- predict(obj, newdata = tail(dat, 1))
  
  # Compute a band 
  r <- residuals(obj)
  s <- ifelse(symmetrize, -1, NA) # Should the residuals be symmetrized?
  q <- quantile(c(r, s * r), probs = c(lower_level, upper_level), na.rm = TRUE)
  lower <- point + q[1]
  upper <- point + q[2]
  
  # Clip at zero if we need to, then return
  if (nonneg) { 
    point = max(point, 0) 
    lower = max(lower, 0) 
    upper = max(upper, 0) 
  }
  return(data.frame(point = point, lower = lower, upper = upper))
}
```

Next we slide this forecaster over the working `epi_archive` object, in order to
forecast COVID-19 case rates 7 days into the future.

```{r}
fc_time_values <- seq(as.Date("2020-08-01"), 
                      as.Date("2021-12-01"), 
                      by = "1 month")

z <- epix_slide(x, fc = prob_arx(x = percent_cli, y = case_rate_7d_av), n = 120, 
                ref_time_values = fc_time_values, group_by = geo_value)

head(z, 10)
```

We get back a tibble `z` with the grouping variables (here geo value), the time
values, and three columns `fc_point`, `fc_lower`, and `fc_upper` produced by the
slide computation that correspond to the point forecast, and the lower and upper
endpoints of the 95\% prediction band, respectively. (If instead we had set
`as_list_col = TRUE` in the call to `epix_slide()`, then we would have gotten a
list column `fc`, where each element of `fc` is a data frame with named columns
`point`, `lower`, and `upper`.)

On the whole, `epix_slide()` works similarly to `epix_slide()`, though there are
a few notable differences, even apart from the version-aware aspect. You can
read the documentation for `epix_slide()` for details.

We finish off by comparing version-aware and -unaware forecasts at various 
points in time and forecast horizons. The former comes from applying 
`epix_slide()` to the `epi_archive` object `x`, and the latter from applying
`epi_slide()` to the latest snapshot of the data `x_latest`.

```{r, message = FALSE, warning = FALSE, fig.width = 9, fig.height = 6}
x_latest <- epix_as_of(x, max_version = max(x$DT$version))

# Simple function to produce forecasts k weeks ahead
k_week_ahead <- function(x, ahead = 7, as_of = TRUE) {
  if (as_of) {
    x %>%
      epix_slide(fc = prob_arx(percent_cli, case_rate_7d_av, ahead = ahead), n = 120,
                 ref_time_values = fc_time_values, group_by = geo_value) %>%
      mutate(target_date = time_value + ahead, as_of = TRUE) 
  }
  else {
    x_latest %>% 
      group_by(geo_value) %>%
      epi_slide(fc = prob_arx(percent_cli, case_rate_7d_av, ahead = ahead), n = 120,
                ref_time_values = fc_time_values) %>%
      mutate(target_date = time_value + ahead, as_of = FALSE) 
  }
}

# Generate the forecasts, and bind them together
fc <- bind_rows(k_week_ahead(x, ahead = 7, as_of = TRUE),
                k_week_ahead(x, ahead = 14, as_of = TRUE),
                k_week_ahead(x, ahead = 21, as_of = TRUE),
                k_week_ahead(x, ahead = 28, as_of = TRUE),
                k_week_ahead(x, ahead = 7, as_of = FALSE),
                k_week_ahead(x, ahead = 14, as_of = FALSE),
                k_week_ahead(x, ahead = 21, as_of = FALSE),
                k_week_ahead(x, ahead = 28, as_of = FALSE))

# Plot them, on top of latest COVID-19 case rates 
ggplot(fc, aes(x = target_date, group = time_value, fill = as_of)) +
  geom_ribbon(aes(ymin = fc_lower, ymax = fc_upper), alpha = 0.4) +
  geom_line(data = x_latest, aes(x = time_value, y = case_rate_7d_av), 
               inherit.aes = FALSE, color = "gray50") +
  geom_line(aes(y = fc_point)) + geom_point(aes(y = fc_point), size = 0.5) + 
  geom_vline(aes(xintercept = time_value), linetype = 2, alpha = 0.5) +
  facet_grid(vars(geo_value), vars(as_of), scales = "free") +
  scale_x_date(minor_breaks = "month", date_labels = "%b %y") +
  labs(x = "Date", y = "Reported COVID-19 case rates") + 
  theme(legend.position = "none") 
```

Each row displays the forecasts for a different location (CA, FL, NY, and TX), and each
column corresponds to whether properly-versioned data is used (`FALSE` means no,
and `TRUE` means yes). We can see that the properly-versioned forecaster is, at
some points in time, more problematic; for example, it massively overpredicts
the peak in both locations in winter wave of 2020. However, performance is
pretty poor across the board here, whether or not properly-versioned data is
used. Similar to what we saw in the [slide
vignette](https://cmu-delphi.github.io/epiprocess/articles/archive.html), the
ARX forecasts can too volatile, overconfident, or both.

Some of the volatility can be attenuated here by training an ARX model jointly
over locations; the [advanced sliding
vignette](https://cmu-delphi.github.io/epiprocess/articles/advanced.html) gives
a demonstration of how to do this. But really, the
[`epipredict`](https://cmu-delphi.github.io/epipredict/) package, which builds
off the data structures and functionality in the current package, is the place
to look for more robust forecasting methodology. The forecasters that appear in
the vignettes in the current package are only meant to demo the slide
functionality with some of the most basic forecasting methodology possible.

## Attribution
This document contains dataset that is a modified part of the [COVID-19 Data Repository by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University](https://github.com/CSSEGISandData/COVID-19) as [republished in the COVIDcast Epidata API](https://cmu-delphi.github.io/delphi-epidata/api/covidcast-signals/jhu-csse.html). This data set is licensed under the terms of the [Creative Commons Attribution 4.0 International license](https://creativecommons.org/licenses/by/4.0/) by the Johns Hopkins University on behalf of its Center for Systems Science in Engineering. Copyright Johns Hopkins University 2020.

The `percent_cli` data is a modified part of the [COVIDcast Epidata API Doctor Visits data](https://cmu-delphi.github.io/delphi-epidata/api/covidcast-signals/doctor-visits.html). This dataset is licensed under the terms of the [Creative Commons Attribution 4.0 International license](https://creativecommons.org/licenses/by/4.0/). Copyright Delphi Research Group at Carnegie Mellon University 2020.


