---
title: Work with archive objects and data revisions
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Work with archive objects and data revisions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

In addition to the `epi_df` data structure, which we have been working with all
along in these vignettes, the `epiprocess` package has a companion structure
called `epi_archive`. In comparison to an `epi_df` object, which can be seen as
storing a single snapshot of a data set with the most up-to-date signal values
as of some given time, an `epi_archive` object stores the full version history
of a data set. Many signals of interest for epidemiological tracking are subject
to revision (some more than others), and paying attention to data revisions can
be important for all sorts of downstream data analysis and modeling tasks.

This vignette walks through working with `epi_archive` objects and demonstrates
some of their key functionality. We'll work with a signal on the percentage of
doctor's visits with CLI (COVID-like illness) computed from medical insurance
claims, available through the [COVIDcast
API](https://cmu-delphi.github.io/delphi-epidata/api/covidcast.html/). This
signal is subject to very heavy and regular revision; you can read more about it
on its [API documentation
page](https://cmu-delphi.github.io/delphi-epidata/api/covidcast-signals/doctor-visits.html).

```{r, message = FALSE}
library(covidcast)
library(epiprocess)
library(data.table)
library(dplyr)

dv <- covidcast_signal("doctor-visits", 
                       "smoothed_adj_cli",
                       start_day = "2020-06-01", 
                       end_day = "2021-12-01",
                       issues = c("2020-06-01", "2021-12-01"),
                       geo_type = "state", 
                       geo_values = c("ca", "fl"))

colnames(dv)
```

## Getting data into `epi_archive` format

An <code><a href="../reference/epi_archive.html">epi_archive</a></code> object
can be constructed from a data frame, data table, or tibble, provided that it
has (at least) the following columns:

* `geo_value`: the geographic value associated with each row of measurements.
* `time_value`: the time value associated with each row of measurements.
* `version`: the time value specifying the version for each row of measurements.
  For example, if in a given row the `version` is January 15, 2022 and
  `time_value` is January 14, 2022, then this row contains the measurements of
  the data for January 14, 2022 that were available one day later.

As we can see from the above, the data frame returned by
`covidcast::covidcast_signal()` has the columns required for the `epi_archive`
format, with `issue` playing the role of `version`. We can now use
`as_epi_archive()` to bring it into `epi_archive` format.

```{r}
x <- dv %>%
  select(geo_value, time_value, version = issue, percent_cli = value) %>%
  as_epi_archive()

class(x)
print(x)
```

An `epi_archive` is special kind of class called an R6 class. Its primary field
is a data table `DT`, which is of class `data.table` (from the `data.table`
package), and has columns `geo_value`, `time_value`, `version`, as well as any
number of additional columns.

```{r}
class(x$DT)
head(x$DT)
```

The variables `geo_value`, `time_value`, `version` serve as **key variables**
for the data table, as well as any other specified in the metadata (described
below). There can only be a single row per unique combination of key variables,
and therefore the key variables are critical for figuring out how to generate a
snapshot of data from the archive, as of a given version (also described below).
   
```{r}
key(x$DT)
```
 
In general, last-observation-carried-forward (LOCF) is used to data in between
recorded versions.  **A word of caution:** R6 objects, unlike most other objects
in R, have reference semantics. An important consequence of this is that objects
are not copied when modified.
   
```{r}
original_value <- x$DT$percent_cli[1]
y <- x # This DOES NOT make a copy of x
y$DT$percent_cli[1] = 0
head(y$DT)
head(x$DT) 
x$DT$percent_cli[1] <- original_value
```   
   
To make a copy, we can use the `clone()` method for an R6 class, as in `y <-
x$clone()`. You can read more about reference semantics in Hadley Wickham's
[Advanced R](https://adv-r.hadley.nz/r6.html#r6-semantics) book.
 
## Some details on metadata

The following pieces of metadata are included as fields in an `epi_archive`
object: 

* `geo_type`: the type for the geo values.
* `time_type`: the type for the time values.
* `additional_metadata`: list of additional metadata for the data archive.

Metadata for an `epi_archive` object `x` can be accessed (and altered) directly,
as in `x$geo_type` or `x$time_type`, etc. Just like `as_epi_df()`, the function
`as_epi_archive()` attempts to guess metadata fields when an `epi_archive`
object is instantiated, if they are not explicitly specified in the function
call (as it did in the case above).

## Producing snapshots in `epi_df` form

A key method of an `epi_archive` class is `as_of()`, which generates a snapshot
of the archive in `epi_df` format. This represents the most up-to-date values of
the signal variables as of a given version.

```{r}
y <- x$as_of(max_version = as.Date("2021-06-01"))
class(y)
head(y)
max(y$time_value)
attributes(y)$metadata$as_of
```

We can see that the max time value in the `epi_df` object `y` that has been
generated from the archive is May 29, 2021, even though the specified version
date was June 1, 2021. From this we can infer that the doctor's visits signal
was 2 days latent on June 1. Also, we can see that the metadata in the `epi_df`
object has the version date recorded in the `as_of` field.

Using the maximum of the `version` column in the underlying data table in an
`epi_archive` object itself generates a snapshot of the latest values of signal
variables in the entire archive. The `as_of()` function issues a warning in this
case, since updates to the current version may still come in at a later point in
time, due to various reasons like synchronization issues.

```{r}
z <- x$as_of(max_version = max(x$DT$version))
```

Below, we pull several snapshots from the archive, spaced one month apart. We
overlay the corresponding signal curves as colored lines, with the version dates
marked by dotted vertical lines, and draw the "finalized" curve in black, from
the latest snapshot `z` that the archive can provide.

```{r fig.width = 8, fig.height = 7}
library(purrr)
library(ggplot2)
theme_set(theme_bw())

self_max = max(x$DT$version)
versions = seq(as.Date("2020-06-01"), self_max - 1, by = "1 month")
snapshots <- map_dfr(versions, function(v) { 
  x$as_of(max_version = v) %>% mutate(version = v)
}) %>%
  bind_rows(z %>% mutate(version = self_max)) %>%
  mutate(final = version == self_max)

p <- ggplot(snapshots %>% filter(!final),
            aes(x = time_value, y = percent_cli)) +  
  geom_line(aes(color = factor(version))) + 
  geom_vline(aes(color = factor(version), xintercept = version), lty = 2) +
  facet_wrap(~ geo_value, scales = "free_y", ncol = 1) +
  scale_x_date(minor_breaks = "month", date_labels = "%b %y") +
  labs(x = "Date", y = "% of doctor's visits with CLI") + 
  theme(legend.position = "none")

gginnards::append_layers(
  p, geom_line(data = snapshots %>% filter(final),
               aes(x = time_value, y = percent_cli), 
               inherit.aes = FALSE, color = "black"), pos = "top")
```

We can see some interesting and highly nontrivial revision behavior: at some
points in time the provisional data snapshots grossly underestimate the
finalized curve (look in particular at Florida close to the end of 2021), and at
others they overestimate it (both states towards the beginning of 2021), though
not quite as dramatically. Modeling the revision process, which is often called
*backfill* is an important statistical problem in it of itself.

*todo: refer to some project/code? perhaps we should think about writing a
 function in `epiprocess` or even a separate package?*

# Sliding version-aware computations
    
Next we demonstrate another key method of `epi_archive`, which is the `slide()`
method. It works just like `epi_slide()` does for an `epi_df` object, but with
one key difference: it performs version-aware computations. That is, for the 
computation at any given reference time t, it uses only **data that would have 
been available as of t**.

To demonstrate this, we're going to ...

```{r eval = FALSE}
y <- covidcast_signal("jhu-csse", 
                      "confirmed_7dav_incidence_prop",
                      start_day = "2020-06-01", 
                      end_day = "2021-12-01",
                      issues = c("2020-06-01", "2021-12-01"),
                      geo_type = "state",
                      geo_values = c("ca", "fl")) %>%
  select(geo_value, time_value, version = issue, case_rates = value) %>%
  as_epi_archive()

# Do a full join at the data.table level
id <- c("geo_value", "time_value", "version")
unique_ids <- unique(c(x$DT[,..id], y$DT[,..id]))


DT <- merge.data.table(x$DT, y$DT, all = TRUE)

DT[, percent_cli = nafill(percent_cli, type = "locf"), by = .(geo_value, )]


setkeyv(DT, cols = setdiff(key(DT), "version"))
setnafill(DT, type = "locf", cols = names(DT), key(DT)

xx %>%
  table.express::group_by(t) %>%
  setnafill(type = "locf", # Fill according to LOCF
            cols = c("var", "var2"))

xx[, var := nafill(var, type = "locf"), by  = t]

cols = c("var", "var2")
key_col = "t"
xx[, ..cols := nafill(..cols, type = "locf"), by = t]


x$as_of(self_max) %>% 
  ggplot(aes(x = time_value)) +
  geom_line(aes(y = percent_cli), col = "red") + 
  geom_line(aes(y = case_rates / 1000), col = "blue") +
  facet_wrap(~ geo_value, scales = "free_y", ncol = 1) +
  scale_x_date(minor_breaks = "month", date_labels = "%b %y") +
  labs(x = "Date", y = "")
```

[about training on finalized versus honest data...]

This makes a bigger deal for some signals than others; for reported
COVID-19 cases, data revisions are most often made in the form of big spikes or
drops; for other signals (like those based on medical insurance claims), the 
revision process can be more regular and incremental. 
