---
title: Work with archive objects and data revisions
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Work with archive objects and data revisions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

In addition to the `epi_df` data structure, the `epiprocess` package has a
companion structure called `epi_archive`. In comparison to an `epi_df` object,
which can be seen as storing a single snapshot of a data set with the most
up-to-date signal values as of some given time, an `epi_archive` object stores
the full version history of a data set. Many signals of interest for
epidemiological tracking are subject to revision (some more than others) and
paying attention to data revisions can be important for all sorts of downstream
data analysis and modeling tasks.

This vignette walks through working with `epi_archive()` objects and demonstrates
some of their key functionality. We'll work with a signal on the percentage of
doctor's visits with CLI (COVID-like illness) computed from medical insurance
claims, available through the [COVIDcast
API](https://cmu-delphi.github.io/delphi-epidata/api/covidcast.html). This
signal is subject to very heavy and regular revision; you can read more about it
on its [API documentation
page](https://cmu-delphi.github.io/delphi-epidata/api/covidcast-signals/doctor-visits.html).

```{r, message = FALSE, warning = FALSE, eval=FALSE}
library(epidatr)
library(epiprocess)
library(data.table)
library(dplyr)
library(purrr)
library(ggplot2)

dv <- pub_covidcast(
  source = "doctor-visits",
  signals = "smoothed_adj_cli",
  geo_type = "state",
  time_type = "day",
  geo_values = "ca,fl,ny,tx",
  time_values = epirange(20200601, 20211201),
  issues = epirange(20200601, 20211201)
)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(epidatr)
library(epiprocess)
library(data.table)
library(dplyr)
library(purrr)
library(ggplot2)
```

## Getting data into `epi_archive` format

An `epi_archive()` object can be constructed from a data frame, data table, or
tibble, provided that it has (at least) the following columns:

* `geo_value`: the geographic value associated with each row of measurements.
* `time_value`: the time value associated with each row of measurements.
* `version`: the time value specifying the version for each row of measurements.
  For example, if in a given row the `version` is January 15, 2022 and
  `time_value` is January 14, 2022, then this row contains the measurements of
  the data for January 14, 2022 that were available one day later.

As we can see from the above, the data frame returned by
`epidatr::pub_covidcast()` has the columns required for the `epi_archive`
format, with `issue` playing the role of `version`. We can now use
`as_epi_archive()` to bring it into `epi_archive` format. For removal of
redundant version updates in `as_epi_archive` using compactify, please refer to
the [compactify vignette](articles/compactify.html).

```{r, eval=FALSE}
x <- dv %>%
  select(geo_value, time_value, version = issue, percent_cli = value) %>%
  as_epi_archive(compactify = TRUE)

class(x)
print(x)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
x <- archive_cases_dv_subset$DT %>%
  select(geo_value, time_value, version, percent_cli) %>%
  as_epi_archive(compactify = TRUE)

class(x)
print(x)
```

An `epi_archive` is consists of a primary field `DT`, which is a data table
(from the `data.table` package) that has the columns `geo_value`, `time_value`,
`version` (and possibly additional ones), and other metadata fields, such as
`geo_type`.

```{r}
class(x$DT)
head(x$DT)
```

The variables `geo_value`, `time_value`, `version` serve as **key variables**
for the data table, as well as any other specified in the metadata (described
below). There can only be a single row per unique combination of key variables,
and therefore the key variables are critical for figuring out how to generate a
snapshot of data from the archive, as of a given version (also described below).

```{r, error=TRUE}
key(x$DT)
```

In general, the last version of each observation is carried forward (LOCF) to
fill in data between recorded versions.

## Some details on metadata

The following pieces of metadata are included as fields in an `epi_archive`
object:

* `geo_type`: the type for the geo values.

Metadata for an `epi_archive` object `x` can be accessed (and altered) directly,
as in `x$geo_type`, etc. Just like `as_epi_df()`, the function
`as_epi_archive()` attempts to guess metadata fields when an `epi_archive`
object is instantiated, if they are not explicitly specified in the function
call (as it did in the case above).

## Summarizing Revision Behavior
There are many ways to examine the ways that signals change across different revisions.
The simplest that is included directly in epiprocess is `revision_summary()`, which computes simple summary statistics for each key (by default, `(geo_value,time_value)` pairs), such as the lag to the first value (latency). In addition to the per key summary, it also returns an overall summary:
```{r}
revision_details <- revision_summary(x, print_inform = TRUE)
```

So as was mentioned at the top, this is clearly a data set where basically everything has some amount of revisions, only 0.37% have no revision at all, and 0.92 have fewer than 3.
Over 94% change by more than 10%.
On the other hand, most are within plus or minus 20% within 5-9 days, so the revisions converge relatively quickly, even if the revisions continue for longer.

To do more detailed analysis than is possible with the above printing, we have `revision_details`:
```{r}
revision_details %>%
  group_by(geo_value) %>%
  summarise(
    n_rev = mean(n_revisions),
    min_lag = min(min_lag),
    max_lag = max(max_lag),
    spread = mean(spread),
    rel_spread = mean(rel_spread),
    time_near_latest = mean(time_near_latest)
  )
```
Most of the states have similar stats on most of these features, except for Florida, which takes nearly double the amount of time to get close to the right value, with California not too far behind.

## Producing snapshots in `epi_df` form

A key method of an `epi_archive` class is `epix_as_of()`, which generates a snapshot
of the archive in `epi_df` format. This represents the most up-to-date values of
the signal variables as of a given version.

```{r}
x_snapshot <- epix_as_of(x, max_version = as.Date("2021-06-01"))
class(x_snapshot)
head(x_snapshot)
max(x_snapshot$time_value)
attributes(x_snapshot)$metadata$as_of
```

We can see that the max time value in the `epi_df` object `x_snapshot` that was
generated from the archive is May 29, 2021, even though the specified version
date was June 1, 2021. From this we can infer that the doctor's visits signal
was 2 days latent on June 1. Also, we can see that the metadata in the `epi_df`
object has the version date recorded in the `as_of` field.

Below, we pull several snapshots from the archive, spaced one month apart. We
overlay the corresponding signal curves as colored lines, with the version dates
marked by dotted vertical lines, and draw the latest curve in black (from the
latest snapshot `x_latest` that the archive can provide).

```{r, fig.width = 8, fig.height = 7}
theme_set(theme_bw())

self_max <- max(x$DT$version)
versions <- seq(as.Date("2020-06-01"), self_max - 1, by = "1 month")
snapshots <- map_dfr(versions, function(v) {
  epix_as_of(x, max_version = v) %>% mutate(version = v)
}) %>%
  bind_rows(
    x_latest %>% mutate(version = self_max)
  ) %>%
  mutate(latest = version == self_max)

ggplot(
  snapshots %>% filter(!latest),
  aes(x = time_value, y = percent_cli)
) +
  geom_line(aes(color = factor(version)), na.rm = TRUE) +
  geom_vline(aes(color = factor(version), xintercept = version), lty = 2) +
  facet_wrap(~geo_value, scales = "free_y", ncol = 1) +
  scale_x_date(minor_breaks = "month", date_labels = "%b %y") +
  labs(x = "Date", y = "% of doctor's visits with CLI") +
  theme(legend.position = "none") +
  geom_line(
    data = snapshots %>% filter(latest),
    aes(x = time_value, y = percent_cli),
    inherit.aes = FALSE, color = "black", na.rm = TRUE
  )
```

We can see some interesting and highly nontrivial revision behavior: at some
points in time the provisional data snapshots grossly underestimate the latest
curve (look in particular at Florida close to the end of 2021), and at others
they overestimate it (both states towards the beginning of 2021), though not
quite as dramatically. Modeling the revision process, which is often called
*backfill modeling*, is an important statistical problem in it of itself.

## Merging `epi_archive` objects

Now we demonstrate how to merge two `epi_archive` objects together, e.g., so
that grabbing data from multiple sources as of a particular version can be
performed with a single `epix_as_of` call. The function `epix_merge()` is made
for this purpose. Below we merge the working `epi_archive` of versioned
percentage CLI from outpatient visits to another one of versioned COVID-19 case
reporting data, which we fetch the from the [COVIDcast
API](https://cmu-delphi.github.io/delphi-epidata/api/covidcast.html/), on the
rate scale (counts per 100,000 people in the population).

When merging archives, unless the archives have identical data release patterns,
`NA`s can be introduced in the non-key variables for a few reasons:
- to represent the "value" of an observation before its initial release (when we
  need to pair it with additional observations from the other archive that have
  been released)
- to represent the "value" of an observation that has no recorded versions at
  all (in the same sort of situation)
- if requested via `sync="na"`, to represent potential update data that we do
  not yet have access to (e.g., due to encountering issues while attempting to
  download the currently available version data for one of the archives, but not
  the other).

```{r, message = FALSE, warning = FALSE,eval=FALSE}
y <- pub_covidcast(
  source = "jhu-csse",
  signals = "confirmed_7dav_incidence_prop",
  geo_type = "state",
  time_type = "day",
  geo_values = "ca,fl,ny,tx",
  time_values = epirange(20200601, 20211201),
  issues = epirange(20200601, 20211201)
) %>%
  select(geo_value, time_value, version = issue, case_rate_7d_av = value) %>%
  as_epi_archive(compactify = TRUE)

x <- epix_merge(x, y, sync = "locf", compactify = TRUE)
print(x)
head(x$DT)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
x <- archive_cases_dv_subset
print(x)
head(x$DT)
```

## Sliding version-aware computations

Lastly, we demonstrate another key method for archives, which is the
`epix_slide()`. It works just like `epi_slide()` does for an `epi_df` object,
but with one key difference: it performs version-aware computations. That is,
for the computation at any given reference time t, it only uses **data that
would have been available as of t**.

For the demonstration, we'll revisit the forecasting example from the [slide
vignette](https://cmu-delphi.github.io/epiprocess/articles/slide.html), and now
we'll build a forecaster that uses properly-versioned data (that would have been
available in real-time) to forecast future COVID-19 case rates from current and
past COVID-19 case rates, as well as current and past values of the outpatient
CLI signal from medical claims. We'll extend the `prob_ar()` function from the
slide vignette to accomodate exogenous variables in the autoregressive model,
which is often referred to as an ARX model.

```{r}
prob_arx <- function(x, y, lags = c(0, 7, 14), ahead = 7, min_train_window = 20,
                     lower_level = 0.05, upper_level = 0.95, symmetrize = TRUE,
                     intercept = FALSE, nonneg = TRUE) {
  # Return NA if insufficient training data
  if (length(y) < min_train_window + max(lags) + ahead) {
    return(data.frame(point = NA, lower = NA, upper = NA))
  }

  # Useful transformations
  if (!missing(x)) {
    x <- data.frame(x, y)
  } else {
    x <- data.frame(y)
  }
  if (!is.list(lags)) lags <- list(lags)
  lags <- rep(lags, length.out = ncol(x))

  # Build features and response for the AR model, and then fit it
  dat <- do.call(
    data.frame,
    unlist( # Below we loop through and build the lagged features
      purrr::map(seq_len(ncol(x)), function(i) {
        purrr::map(lags[[i]], function(j) lag(x[, i], n = j))
      }),
      recursive = FALSE
    )
  )
  names(dat) <- paste0("x", seq_len(ncol(dat)))
  if (intercept) dat$x0 <- rep(1, nrow(dat))
  dat$y <- lead(y, n = ahead)
  obj <- lm(y ~ . + 0, data = dat)

  # Use LOCF to fill NAs in the latest feature values, make a prediction
  setDT(dat)
  setnafill(dat, type = "locf")
  point <- predict(obj, newdata = tail(dat, 1))

  # Compute a band
  r <- residuals(obj)
  s <- ifelse(symmetrize, -1, NA) # Should the residuals be symmetrized?
  q <- quantile(c(r, s * r), probs = c(lower_level, upper_level), na.rm = TRUE)
  lower <- point + q[1]
  upper <- point + q[2]

  # Clip at zero if we need to, then return
  if (nonneg) {
    point <- max(point, 0)
    lower <- max(lower, 0)
    upper <- max(upper, 0)
  }
  return(data.frame(point = point, lower = lower, upper = upper))
}
```

Next we slide this forecaster over the working `epi_archive` object, in order to
forecast COVID-19 case rates 7 days into the future.

```{r}
fc_time_values <- seq(as.Date("2020-08-01"),
  as.Date("2021-11-30"),
  by = "1 month"
)

z <- x %>%
  group_by(geo_value) %>%
  epix_slide(
    fc = prob_arx(x = percent_cli, y = case_rate_7d_av), .before = 119,
    .ref_time_values = fc_time_values
  ) %>%
  ungroup()

head(z, 10)
```



We get back a tibble `z` with the grouping variables (here geo value), the
(reference) time values, and a ["packed"][tidyr::pack] data frame column `fc`
containing `fc$point`, `fc$lower`, and `fc$upper` that correspond to the point
forecast, and the lower and upper endpoints of the 95\% prediction band,
respectively. (We could also have used `, prob_ar(cases_7dav)` to get three
separate columns `point`, `lower`, and `upper`, or used `fc =
list(prob_ar(cases_7dav))` to make an `fc` column with a ["nested"][tidyr::nest]
format (list of data frames) instead.)

On the whole, `epix_slide()` works similarly to `epix_slide()`, though there are
a few notable differences, even apart from the version-aware aspect. You can
read the documentation for `epix_slide()` for details.

We finish off by comparing version-aware and -unaware forecasts at various
points in time and forecast horizons. The former comes from using
`epix_slide()` with the `epi_archive` object `x`, and the latter from applying
`epi_slide()` to the latest snapshot of the data `x_latest`.

```{r, message = FALSE, warning = FALSE, fig.width = 9, fig.height = 6}
x_latest <- epix_as_of(x, x$versions_end)

# Simple function to produce forecasts k weeks ahead
k_week_ahead <- function(x, ahead = 7, as_of = TRUE) {
  if (as_of) {
    x %>%
      group_by(.data$geo_value) %>%
      epix_slide(
        fc = prob_arx(.data$percent_cli, .data$case_rate_7d_av, ahead = ahead), .before = 119,
        .ref_time_values = fc_time_values
      ) %>%
      mutate(target_date = .data$time_value + ahead, as_of = TRUE) %>%
      ungroup()
  } else {
    x_latest %>%
      group_by(.data$geo_value) %>%
      epi_slide(
        fc = prob_arx(.data$percent_cli, .data$case_rate_7d_av, ahead = ahead), .window_size = 120,
        .ref_time_values = fc_time_values
      ) %>%
      mutate(target_date = .data$time_value + ahead, as_of = FALSE) %>%
      ungroup()
  }
}

# Generate the forecasts, and bind them together
fc <- bind_rows(
  k_week_ahead(x, ahead = 7, as_of = TRUE),
  k_week_ahead(x, ahead = 14, as_of = TRUE),
  k_week_ahead(x, ahead = 21, as_of = TRUE),
  k_week_ahead(x, ahead = 28, as_of = TRUE),
  k_week_ahead(x, ahead = 7, as_of = FALSE),
  k_week_ahead(x, ahead = 14, as_of = FALSE),
  k_week_ahead(x, ahead = 21, as_of = FALSE),
  k_week_ahead(x, ahead = 28, as_of = FALSE)
)

# Plot them, on top of latest COVID-19 case rates
ggplot(fc, aes(x = target_date, group = time_value, fill = as_of)) +
  geom_ribbon(aes(ymin = fc$lower, ymax = fc$upper), alpha = 0.4) +
  geom_line(
    data = x_latest, aes(x = time_value, y = case_rate_7d_av),
    inherit.aes = FALSE, color = "gray50"
  ) +
  geom_line(aes(y = fc$point)) +
  geom_point(aes(y = fc$point), size = 0.5) +
  geom_vline(aes(xintercept = time_value), linetype = 2, alpha = 0.5) +
  facet_grid(vars(geo_value), vars(as_of), scales = "free") +
  scale_x_date(minor_breaks = "month", date_labels = "%b %y") +
  labs(x = "Date", y = "Reported COVID-19 case rates") +
  theme(legend.position = "none")
```

Each row displays the forecasts for a different location (CA, FL, NY, and TX), and each
column corresponds to whether properly-versioned data is used (`FALSE` means no,
and `TRUE` means yes). We can see that the properly-versioned forecaster is, at
some points in time, more problematic; for example, it massively overpredicts
the peak in both locations in winter wave of 2020. However, performance is
pretty poor across the board here, whether or not properly-versioned data is
used. Similar to what we saw in the [slide
vignette](https://cmu-delphi.github.io/epiprocess/articles/archive.html), the
ARX forecasts can too volatile, overconfident, or both.

Some of the volatility can be attenuated here by training an ARX model jointly
over locations; the [advanced sliding
vignette](https://cmu-delphi.github.io/epiprocess/articles/advanced.html) gives
a demonstration of how to do this. But really, the
[`epipredict`](https://cmu-delphi.github.io/epipredict/) package, which builds
off the data structures and functionality in the current package, is the place
to look for more robust forecasting methodology. The forecasters that appear in
the vignettes in the current package are only meant to demo the slide
functionality with some of the most basic forecasting methodology possible.

## Attribution
This document contains a dataset that is a modified part of the [COVID-19 Data Repository by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University](https://github.com/CSSEGISandData/COVID-19) as [republished in the COVIDcast Epidata API](https://cmu-delphi.github.io/delphi-epidata/api/covidcast-signals/jhu-csse.html). This data set is licensed under the terms of the [Creative Commons Attribution 4.0 International license](https://creativecommons.org/licenses/by/4.0/) by the Johns Hopkins University on behalf of its Center for Systems Science in Engineering. Copyright Johns Hopkins University 2020.

The `percent_cli` data is a modified part of the [COVIDcast Epidata API Doctor Visits data](https://cmu-delphi.github.io/delphi-epidata/api/covidcast-signals/doctor-visits.html). This dataset is licensed under the terms of the [Creative Commons Attribution 4.0 International license](https://creativecommons.org/licenses/by/4.0/). Copyright Delphi Research Group at Carnegie Mellon University 2020.


