---
title: Constructing epi_archive objects
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Constructing epi_archive objects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
source("_common.R")
```

The `epi_archive` data structure provided by `epiprocess` provides convenient
ways to work with data sets that are subject to revision (a common occurrence in
the public health space, as situational awareness improves). In comparison to an
`epi_df` object, which can viewed as a data snapshot at a point in time, an
`epi_archive` object stores the full version history of a data set. Paying
attention to data revisions can be important for all sorts of downstream data
analysis and modeling tasks.

In this vignette we will:

- construct an `epi_archive` object from a data frame,
- discuss the required columns and structure.

## Getting data into `epi_archive` format

We will work with a signal on the percentage of doctor's visits with CLI
(COVID-like illness) computed from medical insurance claims, available through
the [COVIDcast
API](https://cmu-delphi.github.io/delphi-epidata/api/covidcast.html). This
signal is subject to very heavy and regular revision; you can read more about it
on its [API documentation
page](https://cmu-delphi.github.io/delphi-epidata/api/covidcast-signals/doctor-visits.html).

The data is included in the [`epidatasets` package](https://cmu-delphi.github.io/epidatasets/),
which is loaded along with `epiprocess`, and can be accessed with:

```{r, message = FALSE, warning = FALSE}
library(epiprocess)
library(data.table)
library(dplyr)
library(purrr)
library(ggplot2)

# This fetches the raw data backing the archive_cases_dv_subset object.
dv <- archive_cases_dv_subset$DT %>%
  as_tibble()
```

The data can also be fetched from the Delphi Epidata API with the following query:

```{r, message = FALSE, warning = FALSE, eval = FALSE}
library(epidatr)

dv <- pub_covidcast(
  source = "doctor-visits",
  signals = "smoothed_adj_cli",
  geo_type = "state",
  time_type = "day",
  geo_values = "ca,fl,ny,tx",
  time_values = epirange(20200601, 20211201),
  issues = epirange(20200601, 20211201)
) %>%
  rename(version = issue, percent_cli = value)
```

An `epi_archive()` object can be constructed from a data frame, data table, or
tibble, provided that it has (at least) the following columns:

* `geo_value`: the geographic value associated with each row of measurements.
* `time_value`: the time value associated with each row of measurements.
* `version`: the time value specifying the version for each row of measurements.
  For example, if in a given row the `version` is January 15, 2022 and
  `time_value` is January 14, 2022, then this row contains the measurements of
  the data for January 14, 2022 that were available one day later.

As we can see from the above, the data frame returned by
`epidatr::pub_covidcast()` has the columns required for the `epi_archive`
format, with `issue` playing the role of `version`. We can now use
`as_epi_archive()` to bring it into `epi_archive` format.

```{r}
dv_archive <- dv %>%
  select(geo_value, time_value, version, percent_cli) %>%
  as_epi_archive()
dv_archive
```

See the `epi_archive()` documentation for more information about its internal
structure.

For more information on using `epi_archive` objects, see `vignette("epi_archive_usage")`.

## Data attribution

This document contains a dataset that is a modified part of the [COVID-19 Data
Repository by the Center for Systems Science and Engineering (CSSE) at Johns
Hopkins University](https://github.com/CSSEGISandData/COVID-19) as [republished
in the COVIDcast Epidata
API](https://cmu-delphi.github.io/delphi-epidata/api/covidcast-signals/jhu-csse.html).
This data set is licensed under the terms of the [Creative Commons Attribution
4.0 International license](https://creativecommons.org/licenses/by/4.0/) by the
Johns Hopkins University on behalf of its Center for Systems Science in
Engineering. Copyright Johns Hopkins University 2020.

The `percent_cli` data is a modified part of the [COVIDcast Epidata API Doctor
Visits
data](https://cmu-delphi.github.io/delphi-epidata/api/covidcast-signals/doctor-visits.html).
This dataset is licensed under the terms of the [Creative Commons Attribution
4.0 International license](https://creativecommons.org/licenses/by/4.0/).
Copyright Delphi Research Group at Carnegie Mellon University 2020.
