---
title: Working with epi_archive objects and data revisions
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with epi_archive objects and data revisions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
source("_common.R")
```

The `epi_archive` data structure provided by `epiprocess` provides convenient
ways to work with data sets that are subject to revision (a common occurrence in
the public health space, as situational awareness improves). In comparison to an
`epi_df` object, which can viewed as a data snapshot at a point in time, an
`epi_archive` object stores the full version history of a data set. Paying
attention to data revisions can be important for all sorts of downstream data
analysis and modeling tasks.

In this vignette we will:

- construct an `epi_archive` object from a data frame,
- summarize revision behavior in the archive,
- produce snapshots of the data in `epi_df` form,
- merge `epi_archive` objects together,
- provide a link to a backtesting forecasting models vignette.

## Getting data into `epi_archive` format

We will work with a signal on the percentage of doctor's visits with CLI
(COVID-like illness) computed from medical insurance claims, available through
the [COVIDcast
API](https://cmu-delphi.github.io/delphi-epidata/api/covidcast.html). This
signal is subject to very heavy and regular revision; you can read more about it
on its [API documentation
page](https://cmu-delphi.github.io/delphi-epidata/api/covidcast-signals/doctor-visits.html).

The data is included in the [`epidatasets` package](https://cmu-delphi.github.io/epidatasets/),
which is loaded along with `epiprocess`, and can be accessed with:

```{r, message = FALSE, warning = FALSE}
library(epiprocess)
library(data.table)
library(dplyr)
library(purrr)
library(ggplot2)

# This fetches the raw data backing the archive_cases_dv_subset object.
dv <- archive_cases_dv_subset$DT %>%
  as_tibble()
```

The data can also be fetched from the Delphi Epidata API with the following query:

```{r, message = FALSE, warning = FALSE, eval = FALSE}
library(epidatr)

dv <- pub_covidcast(
  source = "doctor-visits",
  signals = "smoothed_adj_cli",
  geo_type = "state",
  time_type = "day",
  geo_values = "ca,fl,ny,tx",
  time_values = epirange(20200601, 20211201),
  issues = epirange(20200601, 20211201)
) %>%
  rename(version = issue, percent_cli = value)
```

An `epi_archive()` object can be constructed from a data frame, data table, or
tibble, provided that it has (at least) the following columns:

* `geo_value`: the geographic value associated with each row of measurements.
* `time_value`: the time value associated with each row of measurements.
* `version`: the time value specifying the version for each row of measurements.
  For example, if in a given row the `version` is January 15, 2022 and
  `time_value` is January 14, 2022, then this row contains the measurements of
  the data for January 14, 2022 that were available one day later.

As we can see from the above, the data frame returned by
`epidatr::pub_covidcast()` has the columns required for the `epi_archive`
format, with `issue` playing the role of `version`. We can now use
`as_epi_archive()` to bring it into `epi_archive` format.

```{r}
dv_archive <- dv %>%
  select(geo_value, time_value, version, percent_cli) %>%
  as_epi_archive()
dv_archive
```

See the `epi_archive()` documentation for more information about its internal
structure.

## Producing snapshots in `epi_df` form

A key method of an `epi_archive` class is `epix_as_of()`, which generates a
snapshot of the archive in `epi_df` format. This represents the most up-to-date
values of the signal variables as of a given version.

```{r}
edf <- epix_as_of(dv_archive, as.Date("2021-06-01"))
print(edf)
print(max(edf$time_value))
```

Note that that the max time value in the `epi_df` object is May 29, 2021, even
though the specified version date was June 1, 2021 (note that the `as_of` field
printed above helps us see the date of the snapshot). From this we can infer
that the doctor's visits signal was 2 days latent on June 1.

Now, let's investigate how much this data was revised. We will plot the most
up-to-date version of the time series in black (`edf_latest` below) and then
overlay several revisions from the archive, spaced one month apart, as colored
lines (`snapshots` below). We will also mark the version dates with dotted
vertical lines.

```{r}
edf_latest <- epix_as_of(dv_archive, dv_archive$versions_end)
max_version <- max(dv_archive$DT$version)
versions <- seq(as.Date("2020-06-01"), max_version - 1, by = "1 month")
monthly_snapshots <- map(versions, function(v) {
  epix_as_of(dv_archive, v) %>% mutate(version = v)
}) %>%
  bind_rows(
    edf_latest %>% mutate(version = max_version)
  ) %>%
  mutate(latest = version == max_version)

ggplot(
  monthly_snapshots %>% filter(!latest),
  aes(x = time_value, y = percent_cli)
) +
  geom_line(aes(color = factor(version)), na.rm = TRUE) +
  geom_vline(aes(color = factor(version), xintercept = version), lty = 2) +
  facet_wrap(~geo_value, scales = "free_y", ncol = 1) +
  scale_x_date(minor_breaks = "month", date_labels = "%b %y") +
  labs(x = "Date", y = "% of doctor's visits with CLI") +
  theme(legend.position = "none") +
  geom_line(
    data = monthly_snapshots %>% filter(latest),
    aes(x = time_value, y = percent_cli),
    inherit.aes = FALSE, color = "black", na.rm = TRUE
  )
```

We can see some interesting and highly nontrivial revision behavior: at some
points in time the provisional data snapshots grossly underestimate the latest
curve (look in particular at Florida close to the end of 2021), and at others
they overestimate it (both states towards the beginning of 2021), though not
quite as dramatically. Modeling the revision process, which is often called
*backfill modeling*, is an important statistical problem in it of itself.

## Summarizing revision behavior

There are many ways to examine how signals change across revisions. We provide
the convenient analysis wrapper `revision_summary()`, which computes simple
summary statistics for each key (by default, `(geo_value,time_value)` pairs). In
addition to the per key summary, it also returns an overall summary. Here is an
a sample of the output:

```{r}
revision_details <- revision_summary(dv_archive)
revision_details
```

We can see from the output that, as mentioned above, this data set has a lot of
revisions: there are no keys that have no revision at all and 34% of the keys
change by 10% or more when revised.

To do more detailed analysis than is possible with the above printing, we can
inspect the returned `revision_details` tibble. Here we collect a number of
statistics for each state:

```{r}
revision_details$revision_behavior %>%
  group_by(geo_value) %>%
  summarize(
    n_rev = mean(n_revisions),
    min_lag = min(min_lag),
    max_lag = max(max_lag),
    spread = mean(spread),
    rel_spread = mean(rel_spread),
    lag_near_latest = mean(lag_near_latest)
  )
```

Most of the states have similar stats on most of these features, except for the
`time_near_latest` stat, which is the amount of time that it takes for the
revisions to converge to within 20% of the final value and stay there. It is the
highest for CA and the lowest for TX.

## Merging `epi_archive` objects

A common operation on datasets is merging (or joining) them together, such as
when we grab data from multiple sources for joint analysis or modeling. Merging
two `epi_archive` objects together is a bit tricky however, since we need to handle
datasets that might get revised at different times. The function `epix_merge()`
is made to smooth this out. Below we merge the working `epi_archive` of versioned
percentage CLI from outpatient visits to another one of versioned COVID-19 case
reporting data, which we fetch the from the [COVIDcast
API](https://cmu-delphi.github.io/delphi-epidata/api/covidcast.html/), on the
rate scale (counts per 100,000 people in the population).

```{r, message = FALSE, warning = FALSE, eval=FALSE}
library(epidatr)

y <- pub_covidcast(
  source = "jhu-csse",
  signals = "confirmed_7dav_incidence_prop",
  geo_type = "state",
  time_type = "day",
  geo_values = "ca,fl,ny,tx",
  time_values = epirange(20200601, 20211201),
  issues = epirange(20200601, 20211201)
) %>%
  select(geo_value, time_value, version = issue, case_rate_7d_av = value) %>%
  as_epi_archive()

dv_cases_archive <- epix_merge(dv_archive, y, sync = "locf")
print(dv_cases_archive)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
dv_cases_archive <- archive_cases_dv_subset
print(dv_cases_archive)
```

Note that we have used the `sync = "locf"` argument to specify that we want to
synchronize the two datasets on their disjoint revisions by using the last
observation carried forward (LOCF). For more information, see `epix_merge()`.

## Backtesting forecasting models

One of the most common use cases of `epiprocess::epi_archive()` object is for
accurate model backtesting. See `vignette("backtesting", package="epipredict")`
for an in-depth demo, using a pre-built forecaster in that package.

## Attribution

This document contains a dataset that is a modified part of the [COVID-19 Data
Repository by the Center for Systems Science and Engineering (CSSE) at Johns
Hopkins University](https://github.com/CSSEGISandData/COVID-19) as [republished
in the COVIDcast Epidata
API](https://cmu-delphi.github.io/delphi-epidata/api/covidcast-signals/jhu-csse.html).
This data set is licensed under the terms of the [Creative Commons Attribution
4.0 International license](https://creativecommons.org/licenses/by/4.0/) by the
Johns Hopkins University on behalf of its Center for Systems Science in
Engineering. Copyright Johns Hopkins University 2020.

The `percent_cli` data is a modified part of the [COVIDcast Epidata API Doctor
Visits
data](https://cmu-delphi.github.io/delphi-epidata/api/covidcast-signals/doctor-visits.html).
This dataset is licensed under the terms of the [Creative Commons Attribution
4.0 International license](https://creativecommons.org/licenses/by/4.0/).
Copyright Delphi Research Group at Carnegie Mellon University 2020.
