---
title: Using epi_archive version history
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using epi_archive version history}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
source("_common.R")
```

Once you have data in the `epi_archive` format (as discussed in
`vignette("epi_archive_construction")`), you can use its version history for a
variety of diagnostic and modeling tasks. This vignette demonstrates how to
effectively use an `epi_archive` to analyze revision behavior, generate
historical data snapshots, and merge multiple archives while maintaining
version consistency.

In particular, this vignette we will:

- summarize revision behavior in an archive,
- produce snapshots of the data in `epi_df` form,
- merge `epi_archive` objects together,
- discuss backtesting forecasting models.

## Should I worry about reporting lag and revision behaviors?

Before diving into complex analysis, it is often useful to understand how much
and how often your data is being revised. In this context, `revision_summary()`
computes simple summary statistics for each key (by default, 
`(geo_value,time_value)` pairs).

```{r, message = FALSE, warning = FALSE}
library(epiprocess)
library(dplyr)
library(purrr)
library(ggplot2)

# Load example archive
dv_archive <- archive_cases_dv_subset$DT %>%
    as_epi_archive()

revision_details <- revision_summary(dv_archive, percent_cli)
revision_details
```

We can see from the output that this data set has a lot of revisions: there are
no keys that have no revision at all and 34% of the keys have a spread of more
than 2.2 in their values when revised.

To do more detailed analysis, we can inspect the returned `revision_details`
tibble. Here we collect a number of statistics for each state:

```{r}
revision_details$revision_behavior %>%
    group_by(geo_value) %>%
    summarize(
        n_rev = mean(n_revisions),
        min_lag = min(min_lag),
        max_lag = max(max_lag),
        spread = mean(spread),
        rel_spread = mean(rel_spread),
        lag_near_latest = mean(lag_near_latest)
    )
```

## Producing snapshots in `epi_df` form

A key method of an `epi_archive` class is `epix_as_of()`, which generates a
snapshot of the archive in `epi_df` format. This represents the most up-to-date
values of the signal variables as of a given version.

```{r}
edf <- epix_as_of(dv_archive, as.Date("2021-06-01"))
print(edf)
print(max(edf$time_value))
```

Note that that the max time value in the `epi_df` object is May 29, 2021, even
though the specified version date was June 1, 2021 (note that the `as_of` field
printed above helps us see the date of the snapshot). From this we can infer
that the doctor's visits signal was 2 days latent on June 1.

Now, let's investigate how much this data was revised. We will plot the most
up-to-date version of the time series in black (`edf_latest` below) and then
overlay several revisions from the archive, spaced one month apart, as colored
lines (`snapshots` below). We will also mark the version dates with dotted
vertical lines.

```{r}
edf_latest <- epix_as_of(dv_archive, dv_archive$versions_end)
max_version <- max(dv_archive$DT$version)
versions <- seq(as.Date("2020-06-01"), max_version - 1, by = "1 month")
monthly_snapshots <- map(versions, function(v) {
    epix_as_of(dv_archive, v) %>% mutate(version = v)
}) %>%
    bind_rows(
        edf_latest %>% mutate(version = max_version)
    ) %>%
    mutate(latest = version == max_version)

ggplot(
    monthly_snapshots %>% filter(!latest),
    aes(x = time_value, y = percent_cli)
) +
    geom_line(aes(color = factor(version)), na.rm = TRUE) +
    geom_vline(aes(color = factor(version), xintercept = version), lty = 2) +
    facet_wrap(~geo_value, scales = "free_y", ncol = 1) +
    scale_x_date(minor_breaks = "month", date_labels = "%b %y") +
    labs(x = "Date", y = "% of doctor's visits with CLI") +
    theme(legend.position = "none") +
    geom_line(
        data = monthly_snapshots %>% filter(latest),
        aes(x = time_value, y = percent_cli),
        inherit.aes = FALSE, color = "black", na.rm = TRUE
    )
```

We can see some interesting and highly nontrivial revision behavior: at some
points in time the provisional data snapshots grossly underestimate the latest
curve (look in particular at Florida close to the end of 2021), and at others
they overestimate it (both states towards the beginning of 2021), though not
quite as dramatically. Modeling the revision process, which is often called
*backfill modeling*, is an important statistical problem in it of itself.

## Summarizing revision behavior

There are many ways to examine how signals change across revisions. We provide
the convenient analysis wrapper `revision_summary()`, which computes simple
summary statistics for each key (by default, `(geo_value,time_value)` pairs). In
addition to the per key summary, it also returns an overall summary. Here is an
a sample of the output:

```{r}
revision_details <- revision_summary(dv_archive, percent_cli)
revision_details
```

We can see from the output that, as mentioned above, this data set has a lot of
revisions: there are no keys that have no revision at all and 34% of the keys
have a spread of more than 2.2 in their values when revised.

To do more detailed analysis than is possible with the above printing, we can
inspect the returned `revision_details` tibble. Here we collect a number of
statistics for each state:

```{r}
revision_details$revision_behavior %>%
    group_by(geo_value) %>%
    summarize(
        n_rev = mean(n_revisions),
        min_lag = min(min_lag),
        max_lag = max(max_lag),
        spread = mean(spread),
        rel_spread = mean(rel_spread),
        lag_near_latest = mean(lag_near_latest)
    )
```

Most of the states have similar stats on most of these features, except for the
`time_near_latest` stat, which is the amount of time that it takes for the
revisions to converge to within 20% of the final value and stay there. It is the
highest for CA and the lowest for TX.

## Merging `epi_archive` objects

A common operation on datasets is merging (or joining) them together, such as
when we grab data from multiple sources for joint analysis or modeling. Merging
two `epi_archive` objects together is a bit tricky however, since we need to handle
datasets that might get revised at different times. The function `epix_merge()`
is made to smooth this out. Below we merge the working `epi_archive` of versioned
percentage CLI from outpatient visits to another one of versioned COVID-19 case
reporting data, which we fetch the from the [COVIDcast
API](https://cmu-delphi.github.io/delphi-epidata/api/covidcast.html/), on the
rate scale (counts per 100,000 people in the population).

```{r, message = FALSE, warning = FALSE, eval=FALSE}
library(epidatr)

y <- pub_covidcast(
    source = "jhu-csse",
    signals = "confirmed_7dav_incidence_prop",
    geo_type = "state",
    time_type = "day",
    geo_values = "ca,fl,ny,tx",
    time_values = epirange(20200601, 20211201),
    issues = epirange(20200601, 20211201)
) %>%
    select(geo_value, time_value, version = issue, case_rate_7d_av = value) %>%
    as_epi_archive()

dv_cases_archive <- epix_merge(dv_archive, y, sync = "locf")
print(dv_cases_archive)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
dv_cases_archive <- archive_cases_dv_subset
print(dv_cases_archive)
```

Note that we have used the `sync = "locf"` argument to specify that we want to
synchronize the two datasets on their disjoint revisions by using the last
observation carried forward (LOCF). For more information, see `epix_merge()`.

## Backtesting forecasting models

One of the most common use cases of `epiprocess::epi_archive()` object is for
accurate model backtesting. See [`vignette("backtesting",
package="epipredict")`](https://cmu-delphi.github.io/epipredict/articles/backtesting.html)
for an in-depth demo, using a pre-built forecaster in that package.

## Data attribution

This document contains a dataset that is a modified part of the [COVID-19 Data
Repository by the Center for Systems Science and Engineering (CSSE) at Johns
Hopkins University](https://github.com/CSSEGISandData/COVID-19) as [republished
in the COVIDcast Epidata
API](https://cmu-delphi.github.io/delphi-epidata/api/covidcast-signals/jhu-csse.html).
This data set is licensed under the terms of the [Creative Commons Attribution
4.0 International license](https://creativecommons.org/licenses/by/4.0/) by the
Johns Hopkins University on behalf of its Center for Systems Science in
Engineering. Copyright Johns Hopkins University 2020.

The `percent_cli` data is a modified part of the [COVIDcast Epidata API Doctor
Visits
data](https://cmu-delphi.github.io/delphi-epidata/api/covidcast-signals/doctor-visits.html).
This dataset is licensed under the terms of the [Creative Commons Attribution
4.0 International license](https://creativecommons.org/licenses/by/4.0/).
Copyright Delphi Research Group at Carnegie Mellon University 2020.
