---
title: Working with epi_df objects and time series data
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with epi_df objects and time series data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
source("_common.R")
```

The `epi_df` data structure provided by `epiprocess` provides convenient ways to
perform common processing tasks. In this vignette, we will:

- construct an `epi_df` from a data frame
- perform rolling time-window computations using `epi_slide()`
- perform group-level aggregation using `sum_groups_epi_df()`
- detect and fill time gaps using `complete.epi_df()` and `{tsibble}`
- perform geographic aggregation (not yet implemented)

## Getting data into `epi_df` format

As in `vignette("epiprocess")`, we will fetch daily reported COVID-19 cases from
CA, FL, NY, and TX (note: here we're using new, not cumulative cases) using the
[`epidatr`](https://github.com/cmu-delphi/epidatr) package, and then convert
this to `epi_df` format.

```{r, message = FALSE, warning = FALSE}
library(epiprocess)
library(dplyr)
```

The data is included in this package (via the [`epidatasets`
package](https://cmu-delphi.github.io/epidatasets/)) and can be loaded with:

```{r}
edf <- cases_deaths_subset %>%
  select(geo_value, time_value, cases) %>%
  arrange(geo_value, time_value)
```

The data can also be fetched from the Delphi Epidata API with the following query:

```{r, message = FALSE, eval = FALSE}
library(epidatr)

d <- as.Date("2024-03-20")

edf <- pub_covidcast(
  source = "jhu-csse",
  signals = "confirmed_incidence_num",
  geo_type = "state",
  time_type = "day",
  geo_values = "ca,fl,ny,tx,ga,pa",
  time_values = epirange(20200301, 20211231),
  as_of = d
) %>%
  select(geo_value, time_value, cases = value) %>%
  arrange(geo_value, time_value) %>%
  as_epi_df(as_of = d)
```

The data has 2,684 rows and 3 columns.

## Rolling computations using `epi_slide`

A very common operation in time series processing is aggregating the values of
the time series by applying some function on a rolling time window of data
points. The key tool that allows this is `epi_slide()`. The function always
first makes sure to group the data by the grouping variables of the `epi_df`
object, which includes the `geo_value` and possibly `other_keys` columns. It
then applies the rolling slide computation inside each group.

The `epi_slide()` function has three ways to specify the computation to be
performed:

- by using a tidy evaluation approach
- by passing a formula
- by passing a function

### Slide the tidy way

Usually, the most convenient way to setup a computation in `epi_slide()` is to
pass in an expression for tidy evaluation. In this case, we can simply define
the name of the new column directly as part of the expression, setting it equal
to a computation in which we can access any columns of `.x` by name, just as we
would in a call to, say, `dplyr::mutate()`. For example:

```{r}
slide_output <- edf %>%
  epi_slide(cases_7sd = sd(cases, na.rm = TRUE), .window_size = 7)
```

As a simple sanity check, we visualize the 7-day trailing averages computed on
top of the original counts:

```{r, message = FALSE, warning = FALSE}
library(ggplot2)

ggplot(slide_output, aes(x = time_value)) +
  geom_col(aes(y = cases, fill = geo_value), alpha = 0.5, show.legend = FALSE) +
  geom_line(aes(y = cases_7sd, col = geo_value), show.legend = FALSE) +
  facet_wrap(~geo_value, scales = "free_y") +
  scale_x_date(minor_breaks = "month", date_labels = "%b %y") +
  labs(x = "Date", y = "Reported COVID-19 cases")
```

As we can see from the Texas plot, the state moved to weekly reporting of
COVID-19 cases in summer of 2021.

Note that without `epi_slide()`, the computation is much less convenient. For
instance, a rough equivalent of the above computation would be the following,
which is easy to get wrong:

```{r}
edf %>%
  complete(geo_value, time_value = seq.Date(min(time_value), max(time_value), by = "day")) %>%
  arrange_canonical() %>%
  group_by(geo_value) %>%
  mutate(cases_7sd = slider::slide_dbl(cases, .f = sd, na.rm = TRUE, .before = 7, .after = 0))
```

Furthermore `epi_slide()` allows for selecting `.ref_time_value`, which the
latter recipe does not support.

### Slide with a function

We can also pass a function to the second argument in `epi_slide()`. In this
case, the passed function `.f` must have the form `function(x, g, t, ...)`,
where

- `x` is an epi_df with the same column names as the input epi_df
- `g` is a one-row tibble containing the values of the grouping variables
  for the associated group, for instance `g$geo_value`
- `t` is the ref_time_value for the current window
- `...` are additional arguments

The same computation as above can be done with a function:

```{r}
edf %>%
  epi_slide(.f = function(x, g, t) sd(x$cases, na.rm = TRUE), .window_size = 7)
```

### `epi_slide()` with a formula

The same computation as above can be done with a formula, where all references
to the columns must be made with the prefix `.x$...`, for instance:

```{r}
edf %>%
  epi_slide(~ sd(.x$cases, na.rm = TRUE), .window_size = 7)
```

Note that the name of the column defaults to `slide_value` in the unnamed
formula or function case. This can be adjusted with `.new_col_name`.

### Rolling computations with multiple column outputs

If your formula (or function) returns a data.frame, then the columns of the
data.frame will be unpacked into the resulting `epi_df` (in the sense of
`tidyr::unpack()`). For example, the following computes the 7-day trailing
average of daily cases as well as the the 7-day trailing standard deviation of
daily cases:

```{r}
edf %>%
  epi_slide(
    ~ data.frame(cases_mean = mean(.x$cases, na.rm = TRUE), cases_sd = sd(.x$cases, na.rm = TRUE)),
    .window_size = 7
  )
```

### Optimized rolling mean and sums

For the two most common sliding operations, we offer two optimized versions:
`epi_slide_mean()` and `epi_slide_sum()`. These are much faster than
`epi_slide()`, so we recommend using them when you are only interested in the
mean or sum of a column. The following computes the 7-day trailing mean of daily
cases:

```{r}
edf %>%
  group_by(geo_value) %>%
  epi_slide_mean("cases", .window_size = 7, na.rm = TRUE)
edf %>%
  group_by(geo_value) %>%
  epi_slide_sum("cases", .window_size = 7, na.rm = TRUE)
```

### Running a forecaster on a sliding window of data

The natural next step is to use the sliding window to forecast future values.
However to do this correctly, we should make sure that our data is historically
accurate. The data structure we use for that is the `epi_archive` and the
analogous slide function is `epix_slide()`. To read further along this train of
thought, see `vignette("epi_archive")`.

## Adding more keys to an `epi_df` and aggregating groups with `sum_groups_epi_df`

An `epi_df` object can have more key columns than just `geo_value` and
`time_value`. For example, if we have demographic attributes like age group, we
can add this as a key column. We can then aggregate the data by these key
columns using `sum_groups_epi_df()`. Let's use influenza hospitalization rate
data from the CDC system FluSurv as an example. We can get it from the [Delphi
Epidata API](https://cmu-delphi.github.io/delphi-epidata/api/flusurv.html)

```{r}
library(epidatr)
flu_data <- pub_flusurv(
  locations = "ca",
  epiweeks = epirange(201801, 202001),
) %>%
  select(location, epiweek, issue, rate_age_0, rate_age_1, rate_age_2, rate_age_3, rate_age_4) %>%
  tidyr::pivot_longer(cols = starts_with("rate_age_"), names_to = "age_group", values_to = "rate")
flu_data
```

We can now convert this data to an `epi_df` object and set the `age_group`
column as an additional group key:

```{r}
flu_data <- flu_data %>% as_epi_df(other_keys = "age_group", as_of = as.Date("2024-03-20"))
flu_data
```

Note that the `epi_df` object now has an additional key column `age_group`. This
means that there should only be one row for each combination of `geo_value`,
`time_value`, and `age_group` in the dataset (this is enforced at construction
time).

Now we can aggregate the data by `age_group`, if we want to compute the total:

```{r}
group_cols <- key_colnames(exclude = "age_group")
flu_data %>%
  sum_groups_epi_df("rate", group_cols = group_cols)
```

## Detecting and filling time gaps with `complete.epi_df`

Sometimes you may have missing data in your time series. This can be due to
actual missing data, or it can be due to the fact that the data is only reported
on certain days. In the latter case, it is often useful to fill in the missing
data with explicit zeros. This can be done with the `complete.epi_df()`
function.

First, let's create a data set with some missing data. We will reuse the dataset
`edf` from above, but modify it slightly.

```{r}
edf_missing <- edf %>%
  filter(geo_value %in% c("ca", "tx")) %>%
  group_by(geo_value) %>%
  slice(1:3, 5:6)

edf_missing %>%
  print(n = 10)
```

Now let's fill in the missing data with explicit zeros:

```{r}
edf_missing %>%
  complete(
    time_value = seq.Date(min(time_value), max(time_value), by = "day"),
    fill = list(cases = 0)
  ) %>%
  print(n = 12)
```

### Detecting and filling time gaps with `tsibble`

We can also use the `tsibble` package to detect and fill time gaps. We'll work
with county-level reported COVID-19 cases in MA and VT.

The data is included in this package (via the [`epidatasets`
package](https://cmu-delphi.github.io/epidatasets/)) and can be loaded with:

```{r, warning = FALSE, message = FALSE}
library(epiprocess)
library(dplyr)
library(readr)

x <- covid_incidence_county_subset
```

The data can also be fetched from the Delphi Epidata API with the following query:

```{r, message = FALSE, eval = FALSE, warning = FALSE}
library(epidatr)

d <- as.Date("2024-03-20")

# Get mapping between FIPS codes and county&state names:
y <- read_csv("https://github.com/cmu-delphi/covidcast/raw/c89e4d295550ba1540d64d2cc991badf63ad04e5/Python-packages/covidcast-py/covidcast/geo_mappings/county_census.csv", # nolint: line_length_linter
  col_types = c(
    FIPS = col_character(),
    CTYNAME = col_character(),
    STNAME = col_character()
  )
) %>%
  filter(STNAME %in% c("Massachusetts", "Vermont"), STNAME != CTYNAME) %>%
  select(geo_value = FIPS, county_name = CTYNAME, state_name = STNAME)

# Fetch only counties from Massachusetts and Vermont, then append names columns as well
x <- pub_covidcast(
  source = "jhu-csse",
  signals = "confirmed_incidence_num",
  geo_type = "county",
  time_type = "day",
  geo_values = paste(y$geo_value, collapse = ","),
  time_values = epirange(20200601, 20211231),
  as_of = d
) %>%
  select(geo_value, time_value, cases = value) %>%
  inner_join(y, by = "geo_value", relationship = "many-to-one", unmatched = c("error", "drop")) %>%
  as_epi_df(as_of = d)
```

The data contains 16,212 rows and 5 columns.

## Converting to `tsibble` format

For manipulating and wrangling time series data, the
[`tsibble`](https://tsibble.tidyverts.org/index.html) already provides a host of
useful tools. A tsibble object (formerly, of class `tbl_ts`) is basically a
tibble (data frame) but with two specially-marked columns: an **index** column
representing the time variable (defining an order from past to present), and a
**key** column identifying a unique observational unit for each time point. In
fact, the key can be made up of any number of columns, not just a single one.

In an `epi_df` object, the index variable is `time_value`, and the key variable
is typically `geo_value` (though this need not always be the case: for example,
if we have an age group variable as another column, then this could serve as a
second key variable). The `epiprocess` package thus provides an implementation
of `as_tsibble()` for `epi_df` objects, which sets these variables according to
these defaults.

```{r, message = FALSE}
library(tsibble)

xt <- as_tsibble(x)
head(xt)
key(xt)
index(xt)
interval(xt)
```

We can also set the key variable(s) directly in a call to `as_tsibble()`.
Similar to SQL keys, if the key does not uniquely identify each time point (that
is, the key and index together do not not uniquely identify each row), then
`as_tsibble()` throws an error:

```{r, error = TRUE}
head(as_tsibble(x, key = "county_name"))
```

As we can see, there are duplicate county names between Massachusetts and
Vermont, which caused the error.

```{r, message = FALSE}
head(duplicates(x, key = "county_name"))
```

Keying by both county name and state name, however, does work:

```{r, message = FALSE}
head(as_tsibble(x, key = c("county_name", "state_name")))
```

One of the major advantages of the `tsibble` package is its ability to handle
**implicit gaps** in time series data. In other words, it can infer what time
scale we're interested in (say, daily data), and detect apparent gaps (say, when
values are reported on January 1 and 3 but not January 2). We can subsequently
use functionality to make these missing entries explicit, which will generally
help avoid bugs in further downstream data processing tasks.

Let's first remove certain dates from our data set to create gaps:

```{r}
state_naming <- read_csv("https://github.com/cmu-delphi/covidcast/raw/c89e4d295550ba1540d64d2cc991badf63ad04e5/Python-packages/covidcast-py/covidcast/geo_mappings/state_census.csv", # nolint: line_length_linter
  col_types = c(NAME = col_character(), ABBR = col_character())
) %>%
  transmute(state_name = NAME, abbr = tolower(ABBR)) %>%
  as_tibble()

# First make geo value more readable for tables, plots, etc.
x <- x %>%
  inner_join(state_naming, by = "state_name", relationship = "many-to-one", unmatched = c("error", "drop")) %>%
  mutate(geo_value = paste(substr(county_name, 1, nchar(county_name) - 7), state_name, sep = ", ")) %>%
  select(geo_value, time_value, cases)

xt <- as_tsibble(x) %>% filter(cases >= 3)
```

The functions `has_gaps()`, `scan_gaps()`, `count_gaps()` in the `tsibble`
package each provide useful summaries, in slightly different formats.

```{r}
head(has_gaps(xt))
head(scan_gaps(xt))
head(count_gaps(xt))
```

We can also visualize the patterns of missingness:

```{r, message = FALSE, warning = FALSE}
library(ggplot2)

ggplot(
  count_gaps(xt),
  aes(
    x = reorder(geo_value, desc(geo_value)),
    color = geo_value
  )
) +
  geom_linerange(aes(ymin = .from, ymax = .to)) +
  geom_point(aes(y = .from)) +
  geom_point(aes(y = .to)) +
  coord_flip() +
  labs(x = "County", y = "Date") +
  theme(legend.position = "none")
```

Using the `fill_gaps()` function from `tsibble`, we can replace all gaps by an
explicit value. The default is `NA`, but in the current case, where missingness
is not at random but rather represents a small value that was censored (only a
hypothetical with COVID-19 reports, but certainly a real phenomenon that occurs
in other signals), it is better to replace it by zero, which is what we do here.
(Other approaches, such as LOCF: last observation carried forward in time, could
be accomplished by first filling with `NA` values and then following up with a
second call to `tidyr::fill()`.)

```{r}
fill_gaps(xt, cases = 0) %>%
  head()
```

Note that the time series for Addison, VT only starts on August 27, 2020, even
though the original (uncensored) data set itself was drawn from a period that
went back to June 6, 2020. By setting `.full = TRUE`, we can at zero-fill over
the entire span of the observed (censored) data.

```{r}
xt_filled <- fill_gaps(xt, cases = 0, .full = TRUE)

head(xt_filled)
```

Explicit imputation for missingness (zero-filling in our case) can be important
for protecting against bugs in all sorts of downstream tasks. For example, even
something as simple as a 7-day trailing average is complicated by missingness.
The function `epi_slide()` looks for all rows within a window of 7 days anchored
on the right at the reference time point (when `.window_size = 7`).
But when some days in a given week are missing because they were censored
because they had small case counts, taking an average of the observed case
counts can be misleading and is unintentionally biased upwards. Meanwhile,
running `epi_slide()` on the zero-filled data brings these trailing averages
(appropriately) downwards, as we can see inspecting Plymouth, MA around July 1,
2021.

```{r}
xt %>%
  as_epi_df(as_of = as.Date("2024-03-20")) %>%
  group_by(geo_value) %>%
  epi_slide(cases_7dav = mean(cases), .window_size = 7) %>%
  ungroup() %>%
  filter(
    geo_value == "Plymouth, MA",
    abs(time_value - as.Date("2021-07-01")) <= 3
  ) %>%
  print(n = 7)

xt_filled %>%
  as_epi_df(as_of = as.Date("2024-03-20")) %>%
  group_by(geo_value) %>%
  epi_slide(cases_7dav = mean(cases), .window_size = 7) %>%
  ungroup() %>%
  filter(
    geo_value == "Plymouth, MA",
    abs(time_value - as.Date("2021-07-01")) <= 3
  ) %>%
  print(n = 7)
```

## Geographic aggregation

We do not yet provide tools for geographic aggregation in `epiprocess`. However,
we have some Python geocoding utilities available. Reach out to us if this is
functionality you would like to see us add to `epiprocess`.

## Attribution

The `percent_cli` data is a modified part of the [COVIDcast Epidata API Doctor
Visits
data](https://cmu-delphi.github.io/delphi-epidata/api/covidcast-signals/doctor-visits.html).
This dataset is licensed under the terms of the [Creative Commons Attribution
4.0 International license](https://creativecommons.org/licenses/by/4.0/).
Copyright Delphi Research Group at Carnegie Mellon University 2020.

This document contains a dataset that is a modified part of the [COVID-19 Data
Repository by the Center for Systems Science and Engineering (CSSE) at Johns
Hopkins University](https://github.com/CSSEGISandData/COVID-19) as [republished
in the COVIDcast Epidata
API](https://cmu-delphi.github.io/delphi-epidata/api/covidcast-signals/jhu-csse.html).
This data set is licensed under the terms of the [Creative Commons Attribution
4.0 International license](https://creativecommons.org/licenses/by/4.0/) by the
Johns Hopkins University on behalf of its Center for Systems Science in
Engineering. Copyright Johns Hopkins University 2020.

[From the COVIDcast Epidata
API](https://cmu-delphi.github.io/delphi-epidata/api/covidcast-signals/jhu-csse.html):
These signals are taken directly from the JHU CSSE [COVID-19 GitHub
repository](https://github.com/CSSEGISandData/COVID-19) without changes.

