---
title: Get started with `epiprocess`
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with epiprocess}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This package introduces a common data structure for epidemiological data sets
measured over space and time, and offers associated utilities to perform basic
signal processing tasks. 

## Installing

This package is not on CRAN yet, so it can be installed using the
[`devtools`](https://cran.r-project.org/package=devtools) package:

```{r, eval = FALSE}
devtools::install_github("cmu-delphi/epiprocess", ref = "main")	
```

Building the vignettes, such as this getting started guide, takes a significant
amount of time. They are not included in the package by default. If you want to
include vignettes, then use this modified command:

```{r, eval = FALSE}
devtools::install_github("cmu-delphi/epiprocess", ref = "main",
                         build_vignettes = TRUE, dependencies = TRUE)
```

## Getting data into `epi_df` format

We'll start by showing how to get data into `epi_df` format, which is just a
tibble with a bit of special structure, and is the format assumed by all of the
functions in the `epiprocess` package. An `epi_df` object has (at least) the
following columns:

* `geo_value`: the geographic value associated with each measurement.
* `time_value`: the time value associated with each measurement.

Other columns can be considered as measured variables, which we also broadly
refer to as signal variables. The documentation for `epi_df()` gives more
details about this data format.

A data frame or tibble that has `geo_value` and `time_value` columns can be
converted into an `epi_df` object, using the function `as_epi_df()`. As an
example, we'll look at daily cumulative COVID-19 cases for 4 states (CA, FL, NY,
and TX) in the U.S., over a year spanning mid 2020 to mid 2021, using the
[`covidcast`](https://cmu-delphi.github.io/covidcast/covidcastR/) package to
fetch this data from the [COVIDcast
API](https://cmu-delphi.github.io/delphi-epidata/api/covidcast.html/).

```{r, message = FALSE}
library(covidcast)

cases <- covidcast_signal(data_source = "jhu-csse", 
                          signal = "confirmed_cumulative_num",
                          start_day = "2020-06-01", 
                          end_day = "2021-5-31",
                          geo_type = "state", 
                          geo_values = c("ca", "fl", "ny", "tx"))

class(cases)
colnames(cases)
```

As we can see, a data frame returned `covidcast::covidcast_signal()` has the
columns required for an `epi_df` object (along with many others). A call to
`as_epi_df()`, with further specification of some relevant metadata, brings the
data frame into `epi_df` format.

```{r, message = FALSE}
library(epiprocess)
library(dplyr)

x <- as_epi_df(cases, 
               geo_type = "state",
               time_type = "day",
               as_of = max(cases$issue)) %>%
  rename("cases_cum" = value) %>%
  select(geo_value, time_value, cases_cum)

class(x)
summary(x)
head(x)
attributes(x)$metadata
```

## Some info about required metadata

In general, an `epi_df` object must have following fields in its metadata:
 
* `geo_type`: the type for the geo values.
* `time_type`: the type for the time values.
* `as_of`: the time value at which the given data were available.

The last field here, `as_of`, is one of the most unique aspects of an `epi_df`
object. In brief, we can think of an `epi_df` object as a single snapshot of a
data set that contains the most up-to-date values of some signals of interest,
as of the time specified `as_of`. For example, if `as_of` were January 31, 2022,
then the `epi_df` object has the most up-to-date version of the data available
as of January 31, 2022. The `epiprocess` package also provides a companion data
structure called `epi_archive`, which stores the full version history of a given
data set. See the [data versioning
vignette](https://cmu-delphi.github.io/epiprocess/articles/archive.html) for
more.

If either of the `geo_type` or `time_type` arguments are missing in a call to
`as_epi_df()`, then this function will try to infer them from the passed
object. Same with the `as_of` argument, though this is often harder to infer.
See the documentation for `as_epi_df()` more details.

```{r}
x <- as_epi_df(cases) %>%
  rename("cases_cum" = value) %>%
  select(geo_value, time_value, cases_cum)

attributes(x)$metadata
```

## Working with `epi_df` objects downstream

Data in `epi_df` format should be easy to work with downstream; in the other
vignettes, we'll walk through some basic signal processing tasks using functions
provided in the `epiprocess` package; of course, we can also write custom code
for other downstream uses, like plotting, which is pretty easy to do `ggplot2`. 

```{r, message = FALSE, warning = FALSE}
library(ggplot2)
theme_set(theme_bw())

ggplot(x, aes(x = time_value, y = cases_cum, color = geo_value)) + 
  geom_line() +
  scale_x_date(minor_breaks = "month", date_labels = "%b %y") +
  labs(x = "Date", y = "Cumulative COVID-19 cases", color = "State")
```

As a last example, we'll look at data on daily new (not cumulative) SARS cases 
in Canada in 2003, from the [outbreaks](https://github.com/reconverse/outbreaks)
package.

```{r}
x <- outbreaks::sars_canada_2003 %>%
  mutate(geo_value = "ca") %>%
  rename(time_value = date) %>%
  select(geo_value, time_value, starts_with("cases")) %>%
  as_epi_df(geo_type = "nation")

head(x)

library(tidyr)
x <- x %>% 
  pivot_longer(starts_with("cases"), names_to = "type") %>%
  mutate(type = substring(type, 7))

yrange <- range(x %>% group_by(time_value) %>% 
                  summarize(value = sum(value)) %>% pull(value))

ggplot(x, aes(x = time_value, y = value)) +
  geom_col(aes(y = value, fill = type)) +
  scale_x_date(minor_breaks = "month", date_labels = "%b %y") +
  scale_y_continuous(breaks = yrange[1]:yrange[2]) +
  labs(x = "Date", y = "SARS cases in Canada", fill = "Type")
```