---
title: Get started with epiprocess
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with epiprocess}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r, include = FALSE}
source("_common.R")
```

## Overview

This vignette provides a brief introduction to the `{epiprocess}` package. We will
do the following:

- Get data into `epi_df()` format and plot the data
- Perform basic signal processing tasks (lagged differences, rolling average,
  cumulative sum, etc.)
- Detect outliers in the data and apply corrections
- Calculate the growth rate of the data
- Get data into `epi_archive()` format and perform similar signal processing
  tasks

## Getting data into `epi_df` format

We'll start by getting data into `epi_df()` format, which is just a tibble with
a bit of special structure. As an example, we will get COVID-19 confirmed
cumulative case data from JHU CSSE for California, Florida, New York, and Texas,
from March 1, 2020 to January 31, 2022. We have
included this example data in the `epidatasets::covid_confirmed_cumulative_num`
object, which we prepared by downloading the data using
`epidatr::pub_covidcast()`.

```{r, results=FALSE, warning=FALSE, message=FALSE}
library(epidatasets)
library(epidatr)
library(epiprocess)
library(dplyr)
library(tidyr)
library(withr)

covid_confirmed_cumulative_num
class(covid_confirmed_cumulative_num)
colnames(covid_confirmed_cumulative_num)
```

The same data can be downloaded with `{epidatr}` as follows:

```{r eval=FALSE}
covid_confirmed_cumulative_num <- pub_covidcast(
  source = "jhu-csse",
  signals = "confirmed_cumulative_num",
  geo_type = "state",
  time_type = "day",
  geo_values = "ca,fl,ny,tx",
  time_values = epirange(20200301, 20220131),
)
```

The tibble returned has the columns required for an `epi_df` object, `geo_value`
and `time_value`, so we can convert it directly to an `epi_df` object using
`as_epi_df()`.

```{r, message = FALSE}
edf <- covid_confirmed_cumulative_num %>%
  select(geo_value, time_value, cases_cumulative = value) %>%
  as_epi_df() %>%
  group_by(geo_value) %>%
  mutate(cases_daily = cases_cumulative - lag(cases_cumulative, default = 0))
edf
```

In brief, we can think of an `epi_df` object as snapshot of an epidemiological
data set as it was at a particular point in time (recorded in the `as_of`
attribute). We can easily plot the data using the `autoplot()` method (which is
a convenience wrapper to `ggplot2`).

```{r, message = FALSE, warning = FALSE}
edf %>%
  autoplot(cases_cumulative)
```

We can compute the 7 day moving average of the confirmed daily cases for each
geo_value by using the `epi_slide_mean()` function. For a more in-depth guide to
sliding, see `vignette("epi_df")`.

```{r}
edf %>%
  group_by(geo_value) %>%
  epi_slide_mean(cases_daily, .window_size = 7, na.rm = TRUE)
```

We can compute the growth rate of the confirmed cumulative cases for each
geo_value. For a more in-depth guide to growth rates, see `vignette("growth_rate")`.

```{r}
edf %>%
  group_by(geo_value) %>%
  mutate(cases_growth = growth_rate(x = time_value, y = cases_cumulative, method = "rel_change", h = 7))
```

Detect outliers in daily reported cases for each geo_value. For a more in-depth
guide to outlier detection, see `vignette("outliers")`.

```{r message=FALSE}
edf %>%
  group_by(geo_value) %>%
  mutate(outlier_info = detect_outlr(x = time_value, y = cases_daily)) %>%
  ungroup()
```

Add a column to the epi_df object with the daily deaths for each geo_value and
compute the correlations between cases and deaths for each geo_value. For a more
in-depth guide to correlations, see `vignette("correlation")`.

```{r}
df <- pub_covidcast(
  source = "jhu-csse",
  signals = "deaths_incidence_num",
  geo_type = "state",
  time_type = "day",
  geo_values = "ca,fl,ny,tx",
  time_values = epirange(20200301, 20220131),
) %>%
  select(geo_value, time_value, deaths_daily = value) %>%
  as_epi_df() %>%
  arrange_canonical()
edf <- inner_join(edf, df, by = c("geo_value", "time_value"))
edf %>%
  group_by(geo_value) %>%
  epi_slide_mean(deaths_daily, .window_size = 7, na.rm = TRUE) %>%
  epi_cor(cases_daily, deaths_daily)
```

Note that if an epi_df object loses its `geo_value` or `time_value` columns, it
will decay to a regular tibble.

```{r}
edf %>% select(-time_value)
```

## Getting data into `epi_archive` format

We can also get data into `epi_archive()` format, which can be thought of as an
aggregation of many `epi_df` snapshots. We can perform similar signal processing
tasks on `epi_archive` objects as we did on `epi_df` objects, though the
interface is a bit different.

```{r, message = FALSE, warning = FALSE, eval=FALSE}
library(epidatr)
library(epiprocess)
library(data.table)
library(dplyr)
library(purrr)
library(ggplot2)

dv <- pub_covidcast(
  source = "doctor-visits",
  signals = "smoothed_adj_cli",
  geo_type = "state",
  time_type = "day",
  geo_values = "ca,fl,ny,tx",
  time_values = epirange(20200601, 20211201),
  issues = epirange(20200601, 20211201)
) %>%
  select(geo_value, time_value, issue, percent_cli = value) %>%
  as_epi_archive(compactify = TRUE)
dv
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(epidatr)
library(epiprocess)
library(data.table)
library(dplyr)
library(purrr)
library(ggplot2)
dv <- archive_cases_dv_subset$DT %>%
  select(-case_rate_7d_av) %>%
  tidyr::drop_na() %>%
  as_epi_archive(compactify = TRUE)
dv
```

See `vignette("epi_archive")` for a more in-depth guide to `epi_archive` objects.

## Data attribution

This document contains a dataset that is a modified part of the [COVID-19 Data
Repository by the Center for Systems Science and Engineering (CSSE) at Johns
Hopkins University](https://github.com/CSSEGISandData/COVID-19) as [republished
in the COVIDcast Epidata
API](https://cmu-delphi.github.io/delphi-epidata/api/covidcast-signals/jhu-csse.html).
This data set is licensed under the terms of the [Creative Commons Attribution
4.0 International license](https://creativecommons.org/licenses/by/4.0/) by the
Johns Hopkins University on behalf of its Center for Systems Science in
Engineering. Copyright Johns Hopkins University 2020.

[From the COVIDcast Epidata
API](https://cmu-delphi.github.io/delphi-epidata/api/covidcast-signals/jhu-csse.html):
These signals are taken directly from the JHU CSSE [COVID-19 GitHub
repository](https://github.com/CSSEGISandData/COVID-19) without changes.
