---
title: Get started with `epiprocess`
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with epiprocess}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

This package introduces a common data structure for epidemiological data sets
measured over space and time, and offers associated utilities to perform basic
signal processing tasks. 

## Installing

This package is not on CRAN yet, so it can be installed using the
[`devtools`](https://devtools.r-lib.org) package:

```{r, eval = FALSE}
devtools::install_github("cmu-delphi/epiprocess", ref = "main")	
```

Building the vignettes, such as this getting started guide, takes a significant
amount of time. They are not included in the package by default. If you want to
include vignettes, then use this modified command:

```{r, eval = FALSE}
devtools::install_github("cmu-delphi/epiprocess", ref = "main",
                         build_vignettes = TRUE, dependencies = TRUE)
```

## Getting data into `epi_df` format

We'll start by showing how to get data into 
<code><a href="../reference/epi_df.html">epi_df</a></code> format, which is just
a tibble with a bit of special structure, and is the format assumed by all of
the functions in the `epiprocess` package. An `epi_df` object has (at least) the
following columns:

* `geo_value`: the geographic value associated with each row of measurements.
* `time_value`: the time value associated with each row of measurements.

It can have any number of other columns which can serve as measured variables,
which we also broadly refer to as signal variables. The documentation for
 gives more details about this data format.

A data frame or tibble that has `geo_value` and `time_value` columns can be
converted into an `epi_df` object, using the function `as_epi_df()`. As an
example, we'll work with daily cumulative COVID-19 cases from four U.S. states:
CA, FL, NY, and TX, over time span from mid 2020 to early 2022, and we'll use
the [`epidatr`](https://github.com/cmu-delphi/epidatr) package
to fetch this data from the [COVIDcast
API](https://cmu-delphi.github.io/delphi-epidata/api/covidcast.html).

```{r, message = FALSE}
library(epidatr)
library(epiprocess)
library(dplyr)
library(withr)

cases <- covidcast(
  data_source = "jhu-csse",
  signals = "confirmed_cumulative_num",
  time_type = "day",
  geo_type = "state",
  time_values = epirange(20200301, 20220131),
  geo_values = "ca,fl,ny,tx"
) %>% fetch()

colnames(cases)
```

As we can see, a data frame returned by `epidatr::covidcast()` has the
columns required for an `epi_df` object (along with many others). We can use
`as_epi_df()`, with specification of some relevant metadata, to bring the data
frame into `epi_df` format.

```{r, message = FALSE}
x <- as_epi_df(cases, 
               geo_type = "state",
               time_type = "day",
               as_of = max(cases$issue)) %>%
  select(geo_value, time_value, total_cases = value)

class(x)
summary(x)
head(x)
attributes(x)$metadata
```

## Some details on metadata

In general, an `epi_df` object has the following fields in its metadata:
 
* `geo_type`: the type for the geo values.
* `time_type`: the type for the time values.
* `as_of`: the time value at which the given data were available.

Metadata for an `epi_df` object `x` can be accessed (and altered) via
`attributes(x)$metadata`. The first two fields here, `geo_type` and `time_type`,
are not currently used by any downstream functions in the `epiprocess` package,
and serve only as useful bits of information to convey about the data set at
hand. The last field here, `as_of`, is one of the most unique aspects of an
`epi_df` object.

In brief, we can think of an `epi_df` object as a single snapshot of a data set
that contains the most up-to-date values of some signals of interest, as of the
time specified `as_of`. For example, if `as_of` is January 31, 2022, then the
`epi_df` object has the most up-to-date version of the data available as of
January 31, 2022. The `epiprocess` package also provides a companion data
structure called `epi_archive`, which stores the full version history of a given
data set. See the [archive
vignette](https://cmu-delphi.github.io/epiprocess/articles/archive.html) for
more.

If any of the `geo_type`, `time_type`, or `as_of` arguments are missing in a 
call to `as_epi_df()`, then this function will try to infer them from the passed
object. Usually, `geo_type` and `time_type` can be inferred from the `geo_value`
and `time_value` columns, respectively, but inferring the `as_of` field is not 
as easy. See the documentation for `as_epi_df()` more details.

```{r}
x <- as_epi_df(cases) %>%
  select(geo_value, time_value, total_cases = value)

attributes(x)$metadata
```

## Using additional key columns in `epi_df`
In the following examples we will show how to create an `epi_df` with additional keys.

### Converting a `tsibble` that has county code as an extra key
```{r}
ex1 <- tibble(
  geo_value = rep(c("ca", "fl", "pa"), each = 3),
  county_code = c("06059","06061","06067",
             "12111","12113","12117",
             "42101","42103","42105"),
  time_value = rep(seq(as.Date("2020-06-01"), as.Date("2020-06-03"),
                       by = "day"), length.out = length(geo_value)),
  value = 1:length(geo_value) + 0.01 * withr::with_rng_version("3.0.0", withr::with_seed(42, length(geo_value)))
  ) %>% 
  as_tsibble(index = time_value, key = c(geo_value, county_code))

ex1 <- as_epi_df(x = ex1, geo_type = "state", time_type = "day", as_of = "2020-06-03")
```

The metadata now includes `county_code` as an extra key.
```{r}
attr(ex1,"metadata")
```


### Dealing with misspecified column names 

`epi_df` requires there to be columns `geo_value` and `time_value`, if they do not exist then `as_epi_df()` throws an error.
```{r, error = TRUE}
data.frame(
  state = rep(c("ca", "fl", "pa"), each = 3), # misnamed
  pol = rep(c("blue", "swing", "swing"), each = 3), # extra key
  reported_date = rep(seq(as.Date("2020-06-01"), as.Date("2020-06-03"),
                       by = "day"), length.out = length(geo_value)), # misnamed
  value = 1:length(geo_value) + 0.01 * withr::with_rng_version("3.0.0", withr::with_seed(42, length(geo_value)))
) %>% as_epi_df() 
```

The columns can be renamed to match `epi_df` format. In the example below, notice there is also an additional key `pol`. 
```{r}
ex2 <- tibble(
  state = rep(c("ca", "fl", "pa"), each = 3), # misnamed
  pol = rep(c("blue", "swing", "swing"), each = 3), # extra key
  reported_date = rep(seq(as.Date("2020-06-01"), as.Date("2020-06-03"),
                          by = "day"), length.out = length(state)), # misnamed
  value = 1:length(state) + 0.01 * withr::with_rng_version("3.0.0", withr::with_seed(42, length(state)))
) %>% data.frame()

head(ex2) 

ex2 <- ex2 %>% rename(geo_value = state, time_value = reported_date) %>%
  as_epi_df(geo_type = "state", as_of = "2020-06-03", 
            additional_metadata = list(other_keys = "pol"))

attr(ex2,"metadata")
```


### Adding additional keys to an `epi_df` object

In the above examples, all the keys are added to objects that are not `epi_df` objects. We illustrate how to add keys to an `epi_df` object. 

We use a toy data set included in `epiprocess` prepared using the `covidcast` library and are filtering to a single state for simplicity.

```{r}
ex3 <- jhu_csse_county_level_subset %>%
  filter(time_value > "2021-12-01", state_name == "Massachusetts") %>%
  slice_tail(n = 6) 
  
attr(ex3,"metadata") # geo_type is county currently
```

Now we add `state` (MA) and `pol` as new columns to the data and as new keys to the metadata. Reminder that lower case state name abbreviations are what we would expect if this were a `geo_value` column. 

```{r}
ex3 <- ex3 %>% 
  as_tibble() %>% # needed to add the additional metadata
  mutate(
    state = rep(tolower("MA"),6),
    pol = rep(c("blue", "swing", "swing"), each = 2)) %>%
  as_epi_df(additional_metadata = list(other_keys = c("state", "pol")))

attr(ex3,"metadata")
```

Note that the two additional keys we added, `state` and `pol`, are specified as a character vector in the `other_keys` component of the `additional_metadata` list. They must be specified in this manner so that downstream actions on the `epi_df`, like model fitting and prediction, can recognize and use these keys.

Currently `other_keys` metadata in `epi_df` doesn't impact `epi_slide()`, contrary to `other_keys` in `as_epi_archive` which affects how the update data is interpreted.

## Working with `epi_df` objects downstream

Data in `epi_df` format should be easy to work with downstream, since it is a
very standard tabular data format; in the other vignettes, we'll walk through
some basic signal processing tasks using functions provided in the `epiprocess`
package. Of course, we can also write custom code for other downstream uses,
like plotting, which is pretty easy to do `ggplot2`.

```{r, message = FALSE, warning = FALSE}
library(ggplot2)
theme_set(theme_bw())

ggplot(x, aes(x = time_value, y = total_cases, color = geo_value)) + 
  geom_line() +
  scale_x_date(minor_breaks = "month", date_labels = "%b %y") +
  labs(x = "Date", y = "Cumulative COVID-19 cases", color = "State")
```

As a last couple examples, we'll look at some more data sets just to show how 
we might get them into `epi_df` format. Data on daily new (not cumulative) SARS 
cases in Canada in 2003, from the 
[outbreaks](https://github.com/reconverse/outbreaks) package:

```{r}
x <- outbreaks::sars_canada_2003 %>%
  mutate(geo_value = "ca") %>%
  select(geo_value, time_value = date, starts_with("cases")) %>%
  as_epi_df(geo_type = "nation")

head(x)

library(tidyr)
x <- x %>% 
  pivot_longer(starts_with("cases"), names_to = "type") %>%
  mutate(type = substring(type, 7))

yrange <- range(x %>% group_by(time_value) %>% 
                  summarize(value = sum(value)) %>% pull(value))

ggplot(x, aes(x = time_value, y = value)) +
  geom_col(aes(fill = type)) +
  scale_x_date(minor_breaks = "month", date_labels = "%b %y") +
  scale_y_continuous(breaks = yrange[1]:yrange[2]) +
  labs(x = "Date", y = "SARS cases in Canada", fill = "Type")
```

Data on new cases of Ebola in Sierra Leone in 2014, from the same package:

```{r, message = FALSE, fig.width = 9, fig.height = 6}
x <- outbreaks::ebola_sierraleone_2014 %>%
  mutate(
    cases = ifelse(status == "confirmed", 1, 0),
    province = case_when(
      district %in% c("Kailahun", "Kenema", "Kono") ~ "Eastern",
      district %in% c("Bombali", "Kambia", "Koinadugu",
                      "Port Loko", "Tonkolili") ~ "Northern",
      district %in% c("Bo", "Bonthe", "Moyamba", "Pujehun") ~ "Sourthern",
      district %in% c("Western Rural", "Western Urban") ~ "Western")) %>% 
  select(geo_value = province,
         time_value = date_of_onset,
         cases) %>% filter(cases==1) %>%
  group_by(geo_value, time_value) %>% 
  summarise(cases = sum(cases)) %>%
  as_epi_df(geo_type="province")

ggplot(x, aes(x = time_value, y = cases)) + 
  geom_col(aes(fill = geo_value), show.legend = FALSE) + 
  facet_wrap(~ geo_value, scales = "free_y") +
  scale_x_date(minor_breaks = "month", date_labels = "%b %y") +
  labs(x = "Date", y = "Confirmed cases of Ebola in Sierra Leone") 
```



## Attribution
This document contains a dataset that is a modified part of the [COVID-19 Data Repository by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University](https://github.com/CSSEGISandData/COVID-19) as [republished in the COVIDcast Epidata API](https://cmu-delphi.github.io/delphi-epidata/api/covidcast-signals/jhu-csse.html). This data set is licensed under the terms of the [Creative Commons Attribution 4.0 International license](https://creativecommons.org/licenses/by/4.0/) by the Johns Hopkins University on behalf of its Center for Systems Science in Engineering. Copyright Johns Hopkins University 2020.

[From the COVIDcast Epidata API](https://cmu-delphi.github.io/delphi-epidata/api/covidcast-signals/jhu-csse.html): 
 These signals are taken directly from the JHU CSSE [COVID-19 GitHub repository](https://github.com/CSSEGISandData/COVID-19) without changes. 

