---
title: Get started with `epiprocess`
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with `epiprocess`}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%"
)
ggplot2::theme_set(ggplot2::theme_bw())
```

## Overview

## TODO: Actually write the sections below (pieces are already in README.Rmd)

This vignette provides a brief introduction to the `epiprocess` package. We will
do the following:

- Get data into `epi_df` format
- Plot the data
- Perform basic signal processing tasks (rolling average, cumulative sum, etc.)
- Detect outliers in the data and apply corrections
- Calculate the growth rate of the data
- Get data into `epi_archive` format
- Analyze the revision history of the data
- Demonstrate the use of `epi_archive` for backtesting

## Getting data into `epi_df` format

We'll start by getting data into `epi_df` format, which is just a tibble with a
bit of special structure. See `as_epi_df()` for more details. As an example, we
will get COVID-19 confirmed cumulative case data from JHU CSSE for California,
Florida, New York, and Texas, from March 1, 2020 to January 31, 2022.

```{r, results=FALSE, warning=FALSE, message=FALSE}
library(epiprocess)
library(epidatr)
library(dplyr)
library(magrittr)
library(tidyr)
```

```{r cache=TRUE}
df <- pub_covidcast(
  source = "jhu-csse",
  signals = "confirmed_cumulative_num",
  geo_type = "state",
  time_type = "day",
  geo_values = "ca,fl,ny,tx",
  time_values = epirange(20200301, 20220131),
) %>%
  select(geo_value, time_value, cases_cumulative = value)
df
```

The tibble returned by `epidatr::pub_covidcast()` has the columns required
for an `epi_df` object, `geo_value` and `time_value`, so we can convert it
directly to an `epi_df` object using `as_epi_df()`.

```{r, message = FALSE}
edf <- as_epi_df(df)
edf
```

In brief, we can think of an `epi_df` object as snapshot of an epidemiological data set
as it was at a particular point in time (recorded in the `as_of` attribute). We can easily plot the
data using the `autoplot()` method (which is a convenience wrapper to `ggplot2`).

```{r, message = FALSE, warning = FALSE}
edf %>%
  autoplot(cases_cumulative)
```

<details>

As another example, here we will import the daily new (not cumulative) SARS
cases in Canada in 2003, from the
[outbreaks](https://github.com/reconverse/outbreaks) package:

```{r}
edf <- outbreaks::sars_canada_2003 %>%
  mutate(geo_value = "ca") %>%
  select(geo_value, time_value = date, starts_with("cases")) %>%
  pivot_longer(starts_with("cases"), names_to = "type") %>%
  mutate(type = substring(type, 7)) %>%
  as_epi_df(other_keys = "type")

head(edf)

edf %>%
  autoplot()
```

Get confirmed cases of Ebola in Sierra Leone from 2014 to 2015 by province and
date of onset, prepared from line list data from the same package:

```{r, fig.width = 9, fig.height = 6}
edf <- outbreaks::ebola_sierraleone_2014 %>%
  select(district, date_of_onset, status) %>%
  mutate(province = case_when(
    district %in% c("Kailahun", "Kenema", "Kono") ~
      "Eastern",
    district %in% c(
      "Bombali", "Kambia", "Koinadugu", "Port Loko",
      "Tonkolili"
    ) ~
      "Northern",
    district %in% c("Bo", "Bonthe", "Moyamba", "Pujehun") ~
      "Sourthern",
    district %in% c("Western Rural", "Western Urban") ~
      "Western"
  )) %>%
  group_by(geo_value = province, time_value = date_of_onset) %>%
  summarise(cases = sum(status == "confirmed"), .groups = "drop") %>%
  complete(geo_value,
    time_value = full_seq(time_value, period = 1),
    fill = list(cases = 0)
  ) %>%
  as_epi_df()

head(edf)

edf %>%
  autoplot()
```

</details>

## Attribution
This document contains a dataset that is a modified part of the [COVID-19 Data Repository by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University](https://github.com/CSSEGISandData/COVID-19) as [republished in the COVIDcast Epidata API](https://cmu-delphi.github.io/delphi-epidata/api/covidcast-signals/jhu-csse.html). This data set is licensed under the terms of the [Creative Commons Attribution 4.0 International license](https://creativecommons.org/licenses/by/4.0/) by the Johns Hopkins University on behalf of its Center for Systems Science in Engineering. Copyright Johns Hopkins University 2020.

[From the COVIDcast Epidata API](https://cmu-delphi.github.io/delphi-epidata/api/covidcast-signals/jhu-csse.html):
 These signals are taken directly from the JHU CSSE [COVID-19 GitHub repository](https://github.com/CSSEGISandData/COVID-19) without changes.
