---
title: Get started with `epitools`
description: An introductory tutorial with examples.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with epitools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This package introduces a common data structure for epidemiological data sets
measured over space and time, and offers associated utilities to perform basic
signal processing tasks. 

## Installing

This package is not on CRAN yet, so it can be installed using the
[`devtools`](https://cran.r-project.org/package=devtools) package:

```{r, eval = FALSE}
devtools::install_github("cmu-delphi/epitools", ref = "main")	
```

Building the vignettes, such as this getting started guide, takes a significant
amount of time. They are not included in the package by default. If you want to
include vignettes, then use this modified command:

```{r, eval = FALSE}
devtools::install_github("cmu-delphi/epitools", ref = "main",
                         build_vignettes = TRUE, dependencies = TRUE)
```

## Getting data into `epi_signal` format

We'll start by showing how to get data into `epi_signal` format, which is just a
tibble (data frame) with a bit of special structure, and is the format assumed 
by all of the functions in the `epitools` package. An `epi_signal` object has 
(at least) the following columns:  

* `geo_value`: the geographic value associated with each measurement.
* `time_value`: the time value associated with each measurement.

Other columns can be considered as measured variables, which we also broadly
refer to as signal variables. To learn more about the `epi_signal` format, you 
can read the documentation for `as.epi_signal()`.

A data frame or tibble that has `geo_value` and `time_value` columns can be 
converted into an `epi_signal` object, using the function `as.epi_signal()`. As 
an example, we'll look at daily cumulative COVID-19 cases for 4 states (CA, FL, 
NY, and TX) in the U.S., over a year spanning mid 2020 to mid 2021, using the
[`covidcast`](https://cmu-delphi.github.io/covidcast/covidcastR/) 
package to fetch this data from the [COVIDcast
API](https://cmu-delphi.github.io/delphi-epidata/api/covidcast.html/). 

```{r, message = FALSE}
library(covidcast)

case_data <- covidcast_signal(data_source = "jhu-csse", 
                              signal = "confirmed_cumulative_num",
                              start_day = "2020-06-01", 
                              end_day = "2021-5-31",
                              geo_type = "state", 
                              geo_values = c("ca", "fl", "ny", "tx"))

class(case_data)
colnames(case_data)
```

As we can see, a data frame returned `covidcast::covidcast_signal()` has the 
columns required for an `epi_signal` object (along with many others). A call to
`as.epi_signal()`, with further specification of some relevant metadata, brings 
the data frame into `epi_signal` format.

```{r, message = FALSE}
library(epitools)
library(dplyr)

x <- as.epi_signal(case_data, 
                   geo_type = "state",
                   time_type = "day",
                   issue = max(case_data$issue)) %>%
  rename("cases_cum" = value) %>%
  select(geo_value, time_value, cases_cum)

class(x)
summary(x)
head(x)
attributes(x)$metadata
```

## Some info about required metadata

In general, an `epi_signal` object must have following fields in its metadata:
 
* `geo_type`: the type for the geo values.
* `time_type`: the type for the time values.
* `issue`: the time value at which the given data set was issued. 

In a typical case, the `issue` field represents the maximum of the issues of 
individual signal values measured in the data set; hence we would also say that 
the data set is comprised of all signal values observed "as of" the given issue
in the metadata.

If any of the `geo_type`, `time_type`, and `issue` arguments are missing in a 
call to `as.epi_signal()`, then this function will try to infer them from the 
passed object (The documentation for `as.epi_signal()` gives more details.) 

```{r}
x <- as.epi_signal(case_data) %>%
  rename("cases_cum" = value) %>%
  select(geo_value, time_value, cases_cum)

attributes(x)$metadata
```

## Working with `epi_signal` objects downstream

Data in `epi_signal` format should be easy to work with downstream; in the other
vignettes, we'll walk through some basic signal processing tasks using functions 
provided in the `epitools` package; of course, we can also write custom code for
for other downstream uses, like plotting, which is pretty easy to do `ggplot2`.

```{r, message = FALSE, warning = FALSE}
library(ggplot2)
theme_set(theme_bw())

ggplot(x, aes(x = time_value, y = cases_cum, color = geo_value)) + 
  geom_line() +
  scale_x_date(minor_breaks = "month", date_labels = "%b %y") +
  labs(x = "Date", y = "Cumulative COVID-19 cases", color = "State")
```

As a last example, we'll look at data on daily new (not cumulative) SARS cases 
in Canada in 2003, from the [outbreaks](https://github.com/reconverse/outbreaks)
package.

```{r}
x <- outbreaks::sars_canada_2003 %>%
  mutate(geo_value = "ca") %>%
  rename(time_value = date) %>%
  select(geo_value, time_value, starts_with("cases")) %>%
  as.epi_signal(geo_type = "nation")

head(x)

library(tidyr)
x <- x %>% 
  pivot_longer(starts_with("cases"), names_to = "type") %>%
  mutate(type = substring(type, 7))

yrange <- range(x %>% group_by(time_value) %>% 
                  summarize(value = sum(value)) %>% pull(value))

ggplot(x, aes(x = time_value, y = value)) +
  geom_col(aes(y = value, fill = type)) +
  scale_x_date(minor_breaks = "month", date_labels = "%b %y") +
  scale_y_continuous(breaks = yrange[1]:yrange[2]) +
  labs(x = "Date", y = "SARS cases in Canada", fill = "Type")
```