---
title: Estimate growth rates in signals
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Estimate growth rates in signals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

A basic way of assessing growth in a signal is to look at its relative change
over two neighboring time windows. The `epiprocess` package provides a function
`growth_rate()` to compute such relative changes, as well as more sophisticated 
estimates the growth rate of a signal. We investigate this functionality in the
current vignette, applied to state-level daily reported COVID-19 cases from GA
and PA, smoothed using a 7-day trailing average.

```{r, message = FALSE, warning = FALSE}
library(epidatr)
library(epiprocess)
library(dplyr)
library(tidyr)
```

The data is fetched with the following query:
```{r, message = FALSE, eval=F}
x <- covidcast(
  data_source = "jhu-csse",
  signals = "confirmed_7dav_incidence_num",
  time_type = "day",
  geo_type = "state",
  time_values = epirange(20200601, 20211231),
  geo_values = "ga,pa"
) %>%
  fetch() %>%
  select(geo_value, time_value, cases = value) %>%
  arrange(geo_value, time_value) %>%
  as_epi_df()
```

The data has 1,158 rows and 3 columns.


```{r, echo=FALSE}
data(jhu_csse_daily_subset)
x <- jhu_csse_daily_subset %>%
  select(geo_value, time_value, cases = cases_7d_av) %>%
  filter(geo_value %in% c("pa","ga") & time_value >= "2020-06-01") %>%
  arrange(geo_value, time_value) %>%
  as_epi_df()
```

## Growth rate basics

The growth rate of a function $f$ defined over a continuously-valued parameter
$t$ is defined as $f'(t)/f(t)$, where $f'(t)$ is the derivative of $f$ at
$t$. To estimate the growth rate of a signal in discrete-time (which can be
thought of as evaluations or discretizations of an underlying function in
continuous-time), we can estimate the derivative and divide by the signal value
itself (or possibly a smoothed version of the signal value). 

The `growth_rate()` function takes a sequence of underlying design points `x`
and corresponding sequence `y` of signal values, and allows us to choose from
the following methods for estimating the growth rate at a given reference point
`x0`, by setting the `method` argument: 

* "rel_change": uses $(\bar B/\bar A - 1) / h$, where $\bar B$ is the average of
  `y` over the second half of a sliding window of bandwidth `h` centered at the
  reference point `x0`, and $\bar A$ the average over the first half. This can 
  be seen as using a first-difference approximation to the derivative.
* "linear_reg": uses the slope from a linear regression of `y` on `x` over a
  sliding window centered at the reference point `x0`, divided by the fitted
  value from this linear regression at `x0`.
* "smooth_spline": uses the estimated derivative at `x0` from a smoothing spline
  fit to `x` and `y`, via `stats::smooth.spline()`, divided by the fitted value
  of the spline at `x0`.
* "trend_filter": uses the estimated derivative at `x0` from polynomial trend
  filtering (a discrete spline) fit to `x` and `y`, via
  `genlasso::trendfilter()`, divided by the fitted value of the discrete spline
  at `x0`.

The default in `growth_rate()` is `x0 = x`, so that it returns an estimate of
the growth rate at each underlying design point.  

## Relative change

The default method is "rel_change", which is the simplest way to estimate growth
rates. The default bandwidth is `h = 7`, which for daily data, considers the
relative change in a signal over adjacent weeks. We can wrap `growth_rate()` in
a call to `dplyr::mutate()` to append a new column to our `epi_df` object with
the computed growth rates.

```{r}
x <- x %>% 
  group_by(geo_value) %>% 
  mutate(cases_gr1 = growth_rate(time_value, cases))

head(x, 10)
```

We can visualize these growth rate estimates by plotting the signal values and
highlighting the periods in time for which the relative change is above 1% (in
red) and below -1% (in blue), faceting by geo value. 

```{r, message = FALSE, warning = FALSE, fig.width = 9, fig.height = 4}
library(ggplot2)
theme_set(theme_bw())

upper = 0.01
lower = -0.01

ggplot(x, aes(x = time_value, y = cases)) + 
  geom_tile(data = x %>% filter(cases_gr1 >= upper),
            aes(x = time_value, y = 0, width = 7, height = Inf), 
            fill = 2, alpha = 0.08) + 
  geom_tile(data = x %>% filter(cases_gr1 <= lower),
            aes(x = time_value, y = 0, width = 7, height = Inf), 
            fill = 4, alpha = 0.08) +  
  geom_line() +   
  facet_wrap(vars(geo_value), scales = "free_y") +
  scale_x_date(minor_breaks = "month", date_labels = "%b %y") +
  labs(x = "Date", y = "Reported COVID-19 cases")
```

As a more direct visualization, we plot the estimated growth rates themselves,
overlaying the curves for the two states on one plot. 

```{r, message = FALSE, warning = FALSE}
ggplot(x, aes(x = time_value, y = cases_gr1)) + 
  geom_line(aes(col = geo_value)) + 
  geom_hline(yintercept = upper, linetype = 2, col = 2) +
  geom_hline(yintercept = lower, linetype = 2, col = 4) +
  scale_color_manual(values = c(3,6)) +
  scale_x_date(minor_breaks = "month", date_labels = "%b %y") +
  labs(x = "Date", y = "Growth rate", col = "State") 
```

We can see that the estimated growth rates from the relative change method are
somewhat volatile, and there appears to be some bias towards towards the right
boundary of the time span---look at the estimated growth rate for Georgia in
late December 2021, which takes a potentially suspicious dip. In general,
estimation of derivatives will be difficult near the boundary, but relative
changes can suffer from particularly noticeable boundary bias because they are
based on a difference in averages over two halves of a local window, and with
this simplistic approach, one of these halves will be truncated near a boundary.

## Linear regression

The second simplest method available is "linear_reg", whose default bandwidth is
again `h = 7`. Compared to "rel_change", it appears to behave similarly overall,
but thankfully avoids some of the troublesome spikes:

```{r, message = FALSE, warning = FALSE, fig.width = 9, fig.height = 7}
x <- x %>% 
  group_by(geo_value) %>% 
  mutate(cases_gr2 = growth_rate(time_value, cases, method = "linear_reg"))

x %>%
  pivot_longer(cols = starts_with("cases_gr"),
               names_to = "method", 
               values_to = "gr") %>%
  mutate(method = recode(method,
                         cases_gr1 = "rel_change",
                         cases_gr2 = "linear_reg")) %>%
  ggplot(aes(x = time_value, y = gr)) + 
  geom_line(aes(col = method)) + 
  scale_color_manual(values = c(2,4)) +
  facet_wrap(vars(geo_value), scales = "free_y", ncol = 1) +
  scale_x_date(minor_breaks = "month", date_labels = "%b %y") +
  labs(x = "Date", y = "Growth rate", col = "Method") 
```

## Nonparametric estimation

We can also use a nonparametric method to estimate the derivative, through
"smooth_spline" or "trend_filter". The latter is going to be generally more
computationally expensive, but it is also able to adapt better to the local
level of smoothness. (The apparent efficiency is actually compounded by the
particular implementations and default settings for these methods:
"trend_filter" is based on a full solution path algorithm provided in the
`genlasso` package, and performs cross-validation by default in order to pick
the level of regularization; read the documentation for `growth_rate()` more
details.) 

```{r, message = FALSE, warning = FALSE, fig.width = 9, fig.height = 7}
x <- x %>% 
  group_by(geo_value) %>% 
  mutate(cases_gr3 = growth_rate(time_value, cases, method = "smooth_spline"),
         cases_gr4 = growth_rate(time_value, cases, method = "trend_filter"))

x %>%
  select(geo_value, time_value, cases_gr3, cases_gr4) %>%
  pivot_longer(cols = starts_with("cases_gr"),
               names_to = "method", 
               values_to = "gr") %>%
  mutate(method = recode(method,
                         cases_gr3 = "smooth_spline",
                         cases_gr4 = "trend_filter")) %>%
  ggplot(aes(x = time_value, y = gr)) + 
  geom_line(aes(col = method)) + 
  scale_color_manual(values = c(3,6)) +
  facet_wrap(vars(geo_value), scales = "free_y", ncol = 1) +
  scale_x_date(minor_breaks = "month", date_labels = "%b %y") +
  labs(x = "Date", y = "Growth rate", col = "Method") 
```

In this particular example, the trend filtering estimates of growth rate appear 
to be much more stable than those from the smoothing spline, and also much more 
stable than the estimates from local relative changes and linear regressions.

The smoothing spline growth rate estimates are based on the default settings in
`stats::smooth.spline()`, and appear severely under-regularized here. Any of the
arguments to `stats::smooth.spline()` can be customized by passing them as
additional arguments `...` in the call to `growth_rate()`; similarly, we can
also use additional arguments to customize the settings in the underlying trend
filtering functions `genlasso::trendfilter()`, `genlasso::cv.trendfilter()`, and
the documentation for `growth_rate()` gives the full details.

## Log scale estimation

In general, and alternative view for the growth rate of a function $f$ is given
by defining $g(t) = \log(f(t))$, and then observing that $g'(t) = f'(t)/f(t)$.
Therefore, any method that estimates the derivative can be simply applied to the
log of the signal of interest, and in this light, each method above 
("rel_change", "linear_reg", "smooth_spline", and "trend_filter") has a log
scale analog, which can be used by setting the argument `log_scale = TRUE` in
the call to `growth_rate()`.

```{r, message = FALSE, warning = FALSE, fig.width = 9, fig.height = 7}
x <- x %>% 
  group_by(geo_value) %>% 
  mutate(cases_gr5 = growth_rate(time_value, cases, method = "rel_change", 
                                 log_scale = TRUE),
         cases_gr6 = growth_rate(time_value, cases, method = "linear_reg", 
                                 log_scale = TRUE),
         cases_gr7 = growth_rate(time_value, cases, method = "smooth_spline", 
                                 log_scale = TRUE),
         cases_gr8 = growth_rate(time_value, cases, method = "trend_filter", 
                                 log_scale = TRUE))

x %>%
  select(geo_value, time_value, cases_gr5, cases_gr6) %>%
  pivot_longer(cols = starts_with("cases_gr"),
               names_to = "method", 
               values_to = "gr") %>%
  mutate(method = recode(method,
                         cases_gr5 = "rel_change_log",
                         cases_gr6 = "linear_reg_log")) %>%
  ggplot(aes(x = time_value, y = gr)) + 
  geom_line(aes(col = method)) + 
  scale_color_manual(values = c(2,4)) +
  facet_wrap(vars(geo_value), scales = "free_y", ncol = 1) +
  scale_x_date(minor_breaks = "month", date_labels = "%b %y") +
  labs(x = "Date", y = "Growth rate", col = "Method") 

x %>%
  select(geo_value, time_value, cases_gr7, cases_gr8) %>%
  pivot_longer(cols = starts_with("cases_gr"),
               names_to = "method", 
               values_to = "gr") %>%
  mutate(method = recode(method,
                         cases_gr7 = "smooth_spline_log",
                         cases_gr8 = "trend_filter_log")) %>%
  ggplot(aes(x = time_value, y = gr)) + 
  geom_line(aes(col = method)) + 
  scale_color_manual(values = c(3,6)) +
  facet_wrap(vars(geo_value), scales = "free_y", ncol = 1) +
  scale_x_date(minor_breaks = "month", date_labels = "%b %y") +
  labs(x = "Date", y = "Growth rate", col = "Method") 
```

Comparing the `rel_change_log` curves with their `rel_change` counterparts 
(shown in earlier figures), we see that the former curves appear less volatile 
and match the linear regression estimates much more closely. In particular, when
`rel_change` has upward spikes, `rel_change_log` has less pronounced spikes.
Why does this occur? The estimate of $g'(t)$ here can be expressed as $\mathbb
E[\log(B)-\log(A)]/h = \mathbb E[\log(1+hR)]/h$, where $R = ((B-A)/h) / A$, and 
the expectation refers to averaging over the $h$ observations in each window.
Consider the following two relevant inequalities, both due to concavity of the
logarithm function: 

$$
\mathbb E[\log(1+hR)]/h \leq \log(1+h\mathbb E[R])/h \leq \mathbb E[R].
$$

The first inequality is Jensen's; the second inequality is because the tangent 
line of a concave function lies above it. Finally, we observe that $\mathbb
E[R] \approx ((\bar B-\bar A)/h) / \bar A$, which the `rel_change` estimate.  
This explains why the `rel_change_log` curve often lies below the `rel_change`
curve.

<!-- Whether or not a log transform is used should likely be informed by the 
working model for noise. *todo: write something intelligent here?* -->

## Attribution
This document contains a dataset that is a modified part of the [COVID-19 Data Repository by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University](https://github.com/CSSEGISandData/COVID-19) as [republished in the COVIDcast Epidata API](https://cmu-delphi.github.io/delphi-epidata/api/covidcast-signals/jhu-csse.html). This data set is licensed under the terms of the [Creative Commons Attribution 4.0 International license](https://creativecommons.org/licenses/by/4.0/) by the Johns Hopkins University on behalf of its Center for Systems Science in Engineering. Copyright Johns Hopkins University 2020.

[From the COVIDcast Epidata API](https://cmu-delphi.github.io/delphi-epidata/api/covidcast-signals/jhu-csse.html): 
 These signals are taken directly from the JHU CSSE [COVID-19 GitHub repository](https://github.com/CSSEGISandData/COVID-19) without changes.
