---
title: 6. Detect and correct outliers in signals
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{6. Detect and correct outliers in signals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette describes functionality for detecting and correcting outliers in
signals in the `epi_detect_outlr()` function provided in the `epiprocess`
package. This is designed to be modular and extendable, so that you can provide
your own outlier detection and correction functions and use them with `epi_df`
objects. We'll work state-level daily reported COVID-19 case counts in Florida
and New Jersey.

```{r, message = FALSE, fig.width = 8, fig.height = 7}
library(covidcast)
library(epiprocess)
library(dplyr)
library(tidyr)
library(ggplot2)
theme_set(theme_bw())

x <- covidcast_signal(data_source = "jhu-csse",
                      signal = "confirmed_incidence_num",
                      start_day = "2020-06-01",
                      end_day = "2021-05-31",
                      geo_type = "state",
                      geo_values = c("fl", "nj"),
                      as_of = "2021-10-28") %>%
  select(geo_value, time_value, cases = value) %>%
  as_epi_df()

ggplot(x, aes(x = time_value, y = cases)) +
  geom_line() +
  geom_hline(yintercept = 0, linetype = 3) +
  facet_wrap(vars(geo_value), scales = "free_y", ncol = 1) +
  scale_x_date(minor_breaks = "month", date_labels = "%b %y") +
  labs(x = "Date", y = "Reported COVID-19 counts")
```

There are multiple outliers in these data that a modeler may want to detect and
correct. We'll discuss those two tasks in turn.

## Outlier detection

The `epi_detect_outlr()` function allows us to run multiple outlier detection
methods, and then (optionally) combine the results from those methods. Here,
we'll investigate outlier detection results from each of following methods.

1. Detection based on a rolling median, using `epi_detect_outlr_rm()`, which 
  computes a rolling median on with a default window size of 21 time points 
  centered at the time point under consideration, and then computes thresholds 
  based on a multiplier times a rolling IQR computed on the residuals. 
2. Detection based on a seasonal-trend decomposition using LOESS (STL), using
  `epi_detect_outlr_stl()`, which is similar to the rolling median method but 
  replaces the rolling median with fitted values from STL. 
3. Detection based on an STL decomposition, but without seasonality term, which
  amounts to smoothing using LOESS.

The outlier detection methods are specified using a `tibble` that is passed to
`epi_detect_outlr()`, with one row per method, and whose columms specify the
outlier detection function, any input arguments (only nondefault values need to
be supplied), and an abbreviated name for the method used in tracking results.
Abbreviations "rm" and "stl" can be used for the built-in detection functions 
`epi_detect_outlr_rm()` and `epi_detect_outlr_stl()`, respectively.

```{r}
detection_methods = bind_rows(
  tibble(method = "rm",
         args = list(list(detect_negatives = TRUE,
                          detection_multiplier = 2.5)),
         abbr = "rm"),
  tibble(method = "stl",
         args = list(list(detect_negatives = TRUE,
                          detection_multiplier = 2.5)),
         abbr = "stl_seasonal"),
  tibble(method = "stl",
         args = list(list(detect_negatives = TRUE,
                          detection_multiplier = 2.5,
                          seasonal_period = NULL)),
         abbr = "stl_nonseasonal"))

detection_methods
```

Additionally, we'll form combined lower and upper thresholds, calculated as the
median of the lower and upper thresholds from the methods at each time point.
Note that using this combined median threshold is equivalent to using a majority
vote across the base methods to determine whether a value is an outlier.

```{r}
x <- x %>%
  group_by(geo_value) %>%
  epi_detect_outlr(
    var = cases,
    methods = detection_methods,
    combiner = "median",
    new_col_name = "outlier_info") 
```

We can see that `x` has now been updated to include a new column `outlier_info` 
with information about the detection thresholds and possible replacement values 
from each method.

```{r}
head(x)
head(x %>% unnest(outlier_info))
```

To visualize the results, we first define a convenience function for plotting.

```{r fig.width = 9, fig.height = 7}
# Plot outlier detection bands and/or points identified as outliers
plot_outliers = function(x, var, outliers_col = outlier_info, 
                         bands = FALSE, points = TRUE, combined_only = TRUE, 
                         facet_vars = vars(geo_value), nrow = NULL, ncol = NULL, 
                         scales = "fixed") {
  var <- rlang::enquo(var)
  outliers_col <- rlang::enquo(outliers_col)
  
  # Convert outlier detection results to long format 
  x_long <- x %>%
    unnest(!!outliers_col) %>%
    pivot_longer(
      cols = x %>% pull(!!outliers_col) %>% `[[`(1) %>% colnames(),
      names_to = c("method", ".value"),
      names_pattern = "(.+)_(.+)")

  # If requested, filter to only combined method
  if (combined_only) x_long <- x_long %>% filter(method == "combined")
  
  # Start of plot with observed data
  p <- ggplot() +
    geom_line(data = x, mapping = aes(x = time_value, y = !!var))

  # If requested, add bands
  if (bands) 
    p <- p + geom_ribbon(data = x_long, 
                         aes(x = time_value, ymin = lower, ymax = upper, 
                             color = method), fill = NA)

  # If requested, add points
  if (points) {
    x_detected <- x_long %>% filter((!!var < lower) | (!!var > upper))
    p <- p + geom_point(data = x_detected, 
                        aes(x = time_value, y = !!var, color = method, 
                            shape = method))
  }

  # If requested, add faceting
  if (!is.null(facet_vars)) 
    p <- p + facet_wrap(facet_vars, nrow = nrow, ncol = ncol, scales = scales)

  return(p)
}
```

Now we produce plots for each state at a time, faceting by the detection method.

```{r fig.width = 9, fig.height = 14}
plot_outliers(x %>% filter(geo_value == "fl"), cases, 
              bands = TRUE, combined_only = FALSE,
              facet_vars = vars(method), scales = "free_y", ncol = 1) +
  scale_x_date(minor_breaks = "month", date_labels = "%b %y") +
  labs(x = "Date", y = "Reported COVID-19 counts")

plot_outliers(x %>% filter(geo_value == "nj"), cases, 
              bands = TRUE, combined_only = FALSE,
              facet_vars = vars(method), scales = "free_y", ncol = 1) +
  scale_x_date(minor_breaks = "month", date_labels = "%b %y") +
  labs(x = "Date", y = "Reported COVID-19 counts")
```

## Outlier correction

Finally, in order to correct outliers, we can use the posited replacement values
returned by each outlier detection method. Below we use the replacement value
from the combined method, which is defined by the median of replacement values 
from the base methods at each time point.

```{r fig.width = 8, fig.height = 7}
y <- x %>% 
  unnest(outlier_info) %>%
  mutate(cases_corrected = combined_replacement) %>%
  select(geo_value, time_value, cases, cases_corrected) 

y %>% filter(cases != cases_corrected)

ggplot(y, aes(x = time_value)) +
  geom_line(aes(y = cases), linetype = 2) +
  geom_line(aes(y = cases_corrected), col = "red") +
  geom_hline(yintercept = 0, linetype = 3) +
  facet_wrap(vars(geo_value), scales = "free_y", ncol = 1) +
  scale_x_date(minor_breaks = "month", date_labels = "%b %y") +
  labs(x = "Date", y = "Reported COVID-19 counts")
```

More advanced correction functionality will be coming at some point in the 
future. 