---
title: 2. Compute percentage change over time
description: Compute percentage change of signal values over time.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{2. Compute percentage change over time}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

A basic way of assessing growth in a signal is to look at its percentage change
over two neighboring time windows. We investigate this in the current vignette,
using the `pct_change()` function in the `epitools` package. As before, we'll 
look at state-level daily reported COVID-19 cases.

```{r, message = FALSE}
library(covidcast)
library(epitools)
library(dplyr)

x <- covidcast_signal(data_source = "jhu-csse", 
                      signal = "confirmed_incidence_num",
                      start_day = "2020-06-01", 
                      end_day = "2020-12-31",
                      geo_type = "state", 
                      geo_values = "fl")  %>%
  as.epi_signal(name = "covid19_cases") %>%
  select(value, geo_value, time_value, issue)
```

## Percentage change

The `pct_change()` function operates on a `epi_signal` data frame, and takes an 
argument `n`, indicating the size of the local window (in days) to use in order
to compute the percentage change. For example, if `n = 10`, then (for each 
`geo_value`) the percentage change on November 10 is calculated as 
100 * (B - A) / A, where A is the sum of the values between November 6 and 
November 10, and A is the sum of the values between November 1 and November 5. 
The default is `n = 14`, giving the percentage change between adjacent weeks.

```{r}
pct_change(x, n = 14) %>%
  head(21)
```

We can see that a column `pct_change` column has been appended to the output 
data frame, which contains the percentage change values estimates. Next we plot 
these values alongside the signal itself. 

```{r, message = FALSE, warning = FALSE, fig.width = 8, fig.height = 4}
library(ggplot2)
library(gridExtra)
theme_set(theme_bw())

p1 <- x %>% 
  ggplot(aes(x = time_value, y = value)) + geom_line() + 
  scale_x_date(minor_breaks = "month", date_labels = "%b %y") +
  labs(x = "Date", y = "Reported COVID-19 cases")

p2 <- x %>% 
  pct_change(n = 14) %>%
  ggplot(aes(x = time_value, y = pct_change)) + geom_line() + 
  scale_x_date(minor_breaks = "month", date_labels = "%b %y") +
  labs(x = "Date", y = "Weekly percentage change") 

grid.arrange(p1, p2, nrow = 1)
```

## Smoothing via 7-day averaging

The percentage change in the last figure appears pretty volatile in some parts, 
which is not too surprising, looking at how wiggly the signal is itself. With a
suitable call to `slide_to_geo()`, we can either pre-smooth or post-smooth using
a 7-day trailing average. The differences here (between these two approaches) do 
not end up being very noticeable at all.

```{r, message = FALSE, warning = FALSE, fig.width = 8, fig.height = 4}
p1 <- x %>% 
  slide_by_geo(slide_fun = ~ Mean(.x$value), n = 7, col_name = "value") %>%
  pct_change(n = 14) %>%
  ggplot(aes(x = time_value, y = pct_change)) + geom_line() + 
  scale_x_date(minor_breaks = "month", date_labels = "%b %y") +
  labs(x = "Date", y = "Weekly percentage change, pre-smoothed")

p2 <- x %>% 
  pct_change(n = 14) %>%
  slide_by_geo(slide_fun = ~ Mean(.x$pct_change), n = 7) %>%
  ggplot(aes(x = time_value, y = slide_value)) + geom_line() + 
  scale_x_date(minor_breaks = "month", date_labels = "%b %y") +
  labs(x = "Date", y = "Weekly percentage change, post-smoothed") 

grid.arrange(p1, p2, nrow = 1)
```