---
title: 2. Compute percentage change over time
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{2. Compute percentage change over time}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

A basic way of assessing growth in a signal is to look at its percentage change
over two neighboring time windows. We investigate this in the current vignette,
using the `pct_change()` function in the `epitools` package. As before, we'll 
look at state-level daily reported COVID-19 cases.

```{r, message = FALSE}
library(covidcast)
library(epitools)
library(dplyr)

x <- covidcast_signal(data_source = "jhu-csse", 
                      signal = "confirmed_incidence_num",
                      start_day = "2020-06-01", 
                      end_day = "2021-05-31",
                      geo_type = "state", 
                      geo_values = "fl")  %>%
  as.epi_tibble() %>%
  rename(cases = value) %>%
  select(geo_value, time_value, cases)
```

## Percentage change

The `pct_change()` function operates on a `epi_tibble` object, and aside from 
this, takes two main arguments: `var`, the variable whose percentage change 
values are to be computed; and `n`, the length of the local window (here, in 
days) to use to compute the percentage change values. For example, if `n = 10`,
then the percentage change on November 10 is computed as 100 * (B - A) / A, 
where A is the sum of the signal in between November 6 and 10, and A is the sum 
in between November 1 and 5. The default is `n = 14` (here, corresponding to the
percentage change between adjacent weeks).

```{r}
x <- pct_change(x, cases, n = 14)
head(x, 21) %>% print(n = Inf)
```

We can see that a column `pct_change` column has been appended in the returned
object, which contains the percentage change values estimates. Next we plot 
these values alongside the signal itself. We highlight the periods in time for 
which the percentage change is above 10% (in red) and below -10% (in blue).

```{r, message = FALSE, warning = FALSE, fig.width = 9, fig.height = 4}
library(ggplot2)
theme_set(theme_bw())

threshold_upper = 10
threshold_lower = -10

p1 <- ggplot(x, aes(x = time_value, y = cases)) + 
  geom_tile(data = x %>% filter(pct_change >= threshold_upper),
            aes(x = time_value, y = 0, width = 7, height = Inf), 
            fill = 2, alpha = 0.1) + 
  geom_tile(data = x %>% filter(pct_change <= threshold_lower),
            aes(x = time_value, y = 0, width = 7, height = Inf), 
            fill = 4, alpha = 0.1) +  
  geom_line() + 
  scale_x_date(minor_breaks = "month", date_labels = "%b %y") +
  labs(x = "Date", y = "Reported COVID-19 cases")

p2 <- ggplot(x, aes(x = time_value, y = pct_change)) + 
  geom_line() + 
  geom_hline(yintercept = threshold_upper, linetype = 2, col = 2) +
  geom_hline(yintercept = threshold_lower, linetype = 2, col = 4) +
  scale_x_date(minor_breaks = "month", date_labels = "%b %y") +
  labs(x = "Date", y = "Weekly percentage change") 

gridExtra::grid.arrange(p1, p2, nrow = 1)
```

## Smoothing via 7-day averaging

The percentage change in the last figure appears pretty volatile in some parts, 
which is not too surprising, looking at how wiggly the signal is itself. With a
suitable call to `epi_slide()`, we can either pre-smooth or post-smooth using a
7-day trailing average. The differences here (between these two approaches) do 
not end up being very noticeable at all (which is not unexpected, due to the way
in which percentage change is defined; it is already defined using a sum over a 
7-day period). We just show results for one of these approaches, pre-smoothing.

```{r, message = FALSE, warning = FALSE, fig.width = 9, fig.height = 4}
x <- x %>%
  epi_slide(slide_fun = ~ Mean(.x$cases), n = 7, 
            new_col_name = "cases_7dav") %>%
  pct_change("cases_7dav", n = 14)

p1 <- ggplot(x, aes(x = time_value, y = cases_7dav)) + 
  geom_tile(data = x %>% filter(pct_change >= threshold_upper),
            aes(x = time_value, y = 0, width = 7, height = Inf), 
            fill = 2, alpha = 0.1) + 
  geom_tile(data = x %>% filter(pct_change <= threshold_lower),
            aes(x = time_value, y = 0, width = 7, height = Inf), 
            fill = 4, alpha = 0.1) + 
  geom_line() + 
  scale_x_date(minor_breaks = "month", date_labels = "%b %y") +
  labs(x = "Date", y = "Reported COVID-19 cases (7-day average)")

p2 <- ggplot(x, aes(x = time_value, y = pct_change)) + 
  geom_line() + 
  geom_hline(yintercept = threshold_upper, linetype = 2, col = 2) +
  geom_hline(yintercept = threshold_lower, linetype = 2, col = 4) +
  scale_x_date(minor_breaks = "month", date_labels = "%b %y") +
  labs(x = "Date", y = "Weekly percentage change") 

gridExtra::grid.arrange(p1, p2, nrow = 1)
```