This notebook contains sections removed from other notebooks. It currently doesn't compile.

## Some details on metadata

In general, an `epi_df` object has the following fields in its metadata:

* `geo_type`: the type for the geo values.
* `as_of`: the time value at which the given data were available.

Metadata for an `epi_df` object `x` can be accessed (and altered) via
`attributes(x)$metadata`. The field, `geo_type`,is not currently used by any
downstream functions in the `epiprocess` package, and serve only as useful bits
of information to convey about the data set at hand. The last field here,
`as_of`, is one of the most unique aspects of an `epi_df` object.

If `geo_type` or `as_of` arguments are missing in a call to `as_epi_df()`, then
this function will try to infer them from the passed object. Usually, `geo_type`
can be inferred from the `geo_value` columns, respectively, but inferring the
`as_of` field is not as easy. See the documentation for `as_epi_df()` more
details.

## Using additional key columns in `epi_df`

In the following examples we will show how to create an `epi_df` with additional keys.

### Converting a `tsibble` that has county code as an extra key

```{r}
ex1 <- tibble(
  geo_value = rep(c("ca", "fl", "pa"), each = 3),
  county_code = c(
    "06059", "06061", "06067",
    "12111", "12113", "12117",
    "42101", "42103", "42105"
  ),
  time_value = rep(seq(as.Date("2020-06-01"), as.Date("2020-06-03"), by = "day"), length.out = length(geo_value)),
  value = seq_along(geo_value) + 0.01 * withr::with_rng_version("3.0.0", withr::with_seed(42, length(geo_value)))
) %>%
  as_tsibble(index = time_value, key = c(geo_value, county_code))

ex1 <- as_epi_df(x = ex1, as_of = "2020-06-03")
```

The metadata now includes `county_code` as an extra key.
```{r}
attr(ex1, "metadata")
```

### Dealing with misspecified column names

`epi_df` requires there to be columns `geo_value` and `time_value`, if they do not exist then `as_epi_df()` throws an error.

```{r, error = TRUE}
data.frame(
  # misnamed
  state = rep(c("ca", "fl", "pa"), each = 3),
  # extra key
  pol = rep(c("blue", "swing", "swing"), each = 3),
  # misnamed
  reported_date = rep(seq(as.Date("2020-06-01"), as.Date("2020-06-03"), by = "day"), length.out = 9),
  value = 1:9 + 0.01 * withr::with_rng_version("3.0.0", withr::with_seed(42, 9))
) %>% as_epi_df(as_of = as.Date("2024-03-20"))
```

The columns can be renamed to match `epi_df` format. In the example below, notice there is also an additional key `pol`.

```{r}
ex2 <- tibble(
  # misnamed
  state = rep(c("ca", "fl", "pa"), each = 3),
  # extra key
  pol = rep(c("blue", "swing", "swing"), each = 3),
  # misnamed
  reported_date = rep(seq(as.Date("2020-06-01"), as.Date("2020-06-03"), by = "day"), length.out = length(state)),
  value = seq_along(state) + 0.01 * withr::with_rng_version("3.0.0", withr::with_seed(42, length(state)))
) %>% data.frame()

head(ex2)

ex2 <- ex2 %>%
  rename(geo_value = state, time_value = reported_date) %>%
  as_epi_df(
    as_of = "2020-06-03",
    other_keys = "pol"
  )

attr(ex2, "metadata")
```

### Adding additional keys to an `epi_df` object

In the above examples, all the keys are added to objects that are not `epi_df` objects. We illustrate how to add keys to an `epi_df` object.

We use a toy data set included in `epiprocess` prepared using the `covidcast` library and are filtering to a single state for simplicity.

```{r}
ex3 <- jhu_csse_county_level_subset %>%
  filter(time_value > "2021-12-01", state_name == "Massachusetts") %>%
  slice_tail(n = 6)

attr(ex3, "metadata") # geo_type is county currently
```

Now we add `state` (MA) and `pol` as new columns to the data and as new keys to the metadata. Reminder that lower case state name abbreviations are what we would expect if this were a `geo_value` column.

```{r}
ex3 <- ex3 %>%
  as_tibble() %>% # needed to add the additional metadata
  mutate(
    state = rep(tolower("MA"), 6),
    pol = rep(c("blue", "swing", "swing"), each = 2)
  ) %>%
  as_epi_df(other_keys = c("state", "pol"), as_of = as.Date("2024-03-20"))

attr(ex3, "metadata")
```

Note that the two additional keys we added, `state` and `pol`, are specified as a character vector in the `other_keys` argument. They must be specified in this manner so that downstream actions on the `epi_df`, like model fitting and prediction, can recognize and use these keys.

Currently `other_keys` metadata in `epi_df` doesn't impact `epi_slide()`, contrary to `other_keys` in `as_epi_archive` which affects how the update data is interpreted.
