---
title: Slide a computation over signal values
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Slide a computation over signal values}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

A central tool in the `epiprocess` package is `epi_slide()`, which is based on
the powerful functionality provided in the
[`slider`](https://cran.r-project.org/web/packages/slider) package. In
`epiprocess`, to "slide" means to apply a computation---represented as a
function or formula---over a sliding/rolling data window. Suitable
groupings can always be achieved by a preliminary call to `group_by()`.

By default, the meaning of one time step is inferred from the `time_value`
column of the `epi_df` object under consideration, based on the way this column
understands addition and subtraction. For example, if the time values are coded
as `Date` objects, then one time step is one day, since `as.Date("2022-01-01") +
1` equals `as.Date("2022-01-02")`. Alternatively, the time step can be specified
manually in the call to `epi_slide()`; you can read the documentation for more
details. Furthermore, the alignment of the running window used in `epi_slide()`
is specified by `before` and `after`.

As in getting started guide, we'll fetch daily reported COVID-19 cases from CA,
FL, NY, and TX (note: here we're using new, not cumulative cases) using the
[`epidatr`](https://github.com/cmu-delphi/epidatr) package,
and then convert this to `epi_df` format.

```{r, message = FALSE, warning=FALSE}
library(epidatr)
library(epiprocess)
library(dplyr)
```

The data is fetched with the following query:
```{r, message = FALSE, eval=F}
x <- pub_covidcast(
  source = "jhu-csse",
  signals = "confirmed_incidence_num",
  geo_type = "state",
  time_type = "day",
  geo_values = "ca,fl,ny,tx,ga,pa",
  time_values = epirange(20200301, 20211231),
) %>%
  select(geo_value, time_value, cases = value) %>%
  arrange(geo_value, time_value) %>%
  as_epi_df()
```

The data has 2,684 rows and 3 columns.

```{r, echo=FALSE}
data(jhu_csse_daily_subset)
x <- jhu_csse_daily_subset %>%
  select(geo_value, time_value, cases) %>%
  arrange(geo_value, time_value) %>%
  as_epi_df()
```

## Optimized rolling mean

We first demonstrate how to apply a 7-day trailing average to the daily cases
in order to smooth the signal, by passing in the name of the column(s) we
want to average for the first argument of `epi_slide_mean()`. `epi_slide_mean
()` can only be used for averaging. To do this computation per state, we
first call `group_by()`.

```{r}
x %>%
  group_by(geo_value) %>%
  epi_slide_mean("cases", .window_size = 7) %>%
  ungroup() %>%
  head(10)
```

The calculation is done using `data.table::frollmean`, whose behavior can be
adjusted by passing relevant arguments via `...`.

## Slide with a formula

The previous computation can also be performed using `epi_slide()`, which is
more flexible but quite a bit slower than `epi_slide_mean()`. It is
recommended to use `epi_slide_mean()` when possible.

The same 7-day trailing average of daily cases can be computed by passing in a
formula for the first argument of `epi_slide()`. To do this per state, we
first call `group_by()`.

```{r}
x %>%
  group_by(geo_value) %>%
  epi_slide(~ mean(.x$cases), .window_size = 7) %>%
  ungroup() %>%
  head(10)
```

The formula specified has access to all non-grouping columns present in the
original `epi_df` object (and must refer to them with the prefix `.x$`). As we
can see, the function `epi_slide()` returns an `epi_df` object with a new column
appended that contains the results (from sliding), named `slide_value` as the
default. We can of course change this post hoc, or we can instead specify a new
name up front using the `new_col_name` argument:

```{r}
x <- x %>%
  group_by(geo_value) %>%
  epi_slide(~ mean(.x$cases), .window_size = 7, new_col_name = "cases_7dav") %>%
  ungroup()

head(x, 10)
```

Some other information is available in additional variables:

* `.group_key` is a one-row tibble containing the values of the grouping
  variables for the associated group
* `.ref_time_value` is the reference time value the time window was based on

Like in `group_modify()`, there are alternative names for these variables as
well: `.` can be used instead of `.x`, `.y` instead of `.group_key`, and `.z`
instead of `.ref_time_value`.

## Slide with a function

We can also pass a function for the first argument in `epi_slide()`. In this
case, the passed function must accept the following arguments:

In this case, the passed function `f` must accept the following arguments: a
data frame with the same column names as the original object, minus any grouping
variables, containing the time window data for one group-`ref_time_value`
combination; followed by a one-row tibble containing the values of the grouping
variables for the associated group; followed by the associated `ref_time_value`.
It can accept additional arguments; `epi_slide()` will forward any `...` args it
receives to `f`.

Recreating the last example of a 7-day trailing average:

```{r}
x <- x %>%
  group_by(geo_value) %>%
  epi_slide(function(x, gk, rtv) mean(x$cases), .window_size = 7, new_col_name = "cases_7dav") %>%
  ungroup()

head(x, 10)
```

## Slide the tidy way

Perhaps the most convenient way to setup a computation in `epi_slide()` is to
pass in an expression for tidy evaluation. In this case, we can simply define
the name of the new column directly as part of the expression, setting it equal
to a computation in which we can access any columns of `x` by name, just as we
would in a call to `dplyr::mutate()`, or any of the `dplyr` verbs. For example:

```{r}
x <- x %>%
  group_by(geo_value) %>%
  epi_slide(cases_7dav = mean(cases), .window_size = 7) %>%
  ungroup()

head(x, 10)
```
In addition to referring to individual columns by name, you can refer to the
time window data as an `epi_df` or `tibble` using `.x`.  Similarly, the other arguments of the function format are available through the magic names `.group_key` and `.ref_time_value`, and the tidyverse "pronouns" `.data` and `.env` can also be used.

As a simple sanity check, we visualize the 7-day trailing averages computed on
top of the original counts:

```{r, message = FALSE, warning = FALSE, fig.width = 9, fig.height = 6}
library(ggplot2)
theme_set(theme_bw())

ggplot(x, aes(x = time_value)) +
  geom_col(aes(y = cases, fill = geo_value), alpha = 0.5, show.legend = FALSE) +
  geom_line(aes(y = cases_7dav, col = geo_value), show.legend = FALSE) +
  facet_wrap(~geo_value, scales = "free_y") +
  scale_x_date(minor_breaks = "month", date_labels = "%b %y") +
  labs(x = "Date", y = "Reported COVID-19 cases")
```

As we can see from the top right panel, it looks like Texas moved to weekly
reporting of COVID-19 cases in summer of 2021.

## Running a local forecaster

As a more complex example, we create a forecaster based on a local (in time)
autoregression or AR model. AR models can be fit in numerous ways (using base R
functions and various packages), but here we define it "by hand" both because it
provides a more advanced example of sliding a function over an `epi_df` object,
and because it allows us to be a bit more flexible in defining a *probabilistic*
forecaster: one that outputs not just a point prediction, but a notion of
uncertainty around this. In particular, our forecaster will output a point
prediction along with an 90\% uncertainty band, represented by a predictive
quantiles at the 5\% and 95\% levels (lower and upper endpoints of the
uncertainty band).

The function defined below, `prob_ar()`, is our probabilistic AR forecaster. The
`lags`argument indicates which lags to use in the model, and `ahead` indicates
how far ahead in the future to make forecasts (both are encoded in terms of the
units of the `time_value` column; so, days, in the working `epi_df` being
considered in this vignette).

```{r}
prob_ar <- function(y, lags = c(0, 7, 14), ahead = 6, min_train_window = 20,
                    lower_level = 0.05, upper_level = 0.95, symmetrize = TRUE,
                    intercept = FALSE, nonneg = TRUE) {
  # Return NA if insufficient training data
  if (length(y) < min_train_window + max(lags) + ahead) {
    return(data.frame(point = NA, lower = NA, upper = NA))
  }

  # Build features and response for the AR model
  dat <- do.call(
    data.frame,
    purrr::map(lags, function(j) lag(y, n = j))
  )
  names(dat) <- paste0("x", seq_len(ncol(dat)))
  if (intercept) dat$x0 <- rep(1, nrow(dat))
  dat$y <- lead(y, n = ahead)

  # Now fit the AR model and make a prediction
  obj <- lm(y ~ . + 0, data = dat)
  point <- predict(obj, newdata = tail(dat, 1))

  # Compute a band
  r <- residuals(obj)
  s <- ifelse(symmetrize, -1, NA) # Should the residuals be symmetrized?
  q <- quantile(c(r, s * r), probs = c(lower_level, upper_level), na.rm = TRUE)
  lower <- point + q[1]
  upper <- point + q[2]

  # Clip at zero if we need to, then return
  if (nonneg) {
    point <- max(point, 0)
    lower <- max(lower, 0)
    upper <- max(upper, 0)
  }
  return(data.frame(point = point, lower = lower, upper = upper))
}
```

We go ahead and slide this AR forecaster over the working `epi_df` of COVID-19
cases. Note that we actually model the `cases_7dav` column, to operate on the
scale of smoothed COVID-19 cases. This is clearly equivalent, up to a constant,
to modeling weekly sums of COVID-19 cases.

```{r}
fc_time_values <- seq(as.Date("2020-06-01"),
  as.Date("2021-12-01"),
  by = "1 months"
)
x %>%
  group_by(geo_value) %>%
  epi_slide(
    fc = prob_ar(cases_7dav), before = 119,
    ref_time_values = fc_time_values
  ) %>%
  ungroup() %>%
  head(10)
```

Note that here we have utilized an argument `ref_time_values` to perform the
sliding computation (here, compute a forecast) at a specific subset of reference
time values. We get out a ["packed"][tidyr::pack] data frame column `fc`
containing `fc$point`, `fc$lower`, and `fc$upper` that correspond to the point
forecast, and the lower and upper endpoints of the 95\% prediction band,
respectively. (We could also have used `, prob_ar(cases_7dav)` to get three
separate columns `point`, `lower`, and `upper`, or used `fc =
list(prob_ar(cases_7dav))` to make an `fc` column with a ["nested"][tidyr::nest]
format (list of data frames) instead.)

To finish off, we plot the forecasts at some times (spaced out by a few months)
over the last year, at multiple horizons: 7, 14, 21, and 28 days ahead. To do
so, we encapsulate the process of generating forecasts into a simple function,
so that we can call it a few times.

```{r, message = FALSE, warning = FALSE, fig.width = 9, fig.height = 6}
# Note the use of all_rows = TRUE (keeps all original rows in the output)
k_week_ahead <- function(x, ahead = 7) {
  x %>%
    group_by(.data$geo_value) %>%
    epi_slide(
      fc = prob_ar(.data$cases_7dav, ahead = ahead), before = 119,
      ref_time_values = fc_time_values, all_rows = TRUE
    ) %>%
    ungroup() %>%
    mutate(target_date = .data$time_value + ahead)
}

# First generate the forecasts, and bind them together
z <- bind_rows(
  k_week_ahead(x, ahead = 7),
  k_week_ahead(x, ahead = 14),
  k_week_ahead(x, ahead = 21),
  k_week_ahead(x, ahead = 28)
)

# Now plot them, on top of actual COVID-19 case counts
ggplot(z) +
  geom_line(aes(x = time_value, y = cases_7dav), color = "gray50") +
  geom_ribbon(aes(
    x = target_date, ymin = fc$lower, ymax = fc$upper,
    group = time_value
  ), fill = 6, alpha = 0.4) +
  geom_line(aes(x = target_date, y = fc$point, group = time_value)) +
  geom_point(aes(x = target_date, y = fc$point, group = time_value),
    size = 0.5
  ) +
  geom_vline(
    data = tibble(x = fc_time_values), aes(xintercept = x),
    linetype = 2, alpha = 0.5
  ) +
  facet_wrap(vars(geo_value), scales = "free_y") +
  scale_x_date(minor_breaks = "month", date_labels = "%b %y") +
  labs(x = "Date", y = "Reported COVID-19 cases")
```

Two points are worth making. First, the AR model's performance here is pretty
spotty. At various points in time, we can see that its forecasts are volatile
(its point predictions are all over the place), or overconfident (its bands are
too narrow), or both at the same time. This is only meant as a simple demo and
not entirely unexpected given the way the AR model is set up. The
[`epipredict`](https://cmu-delphi.github.io/epipredict) package, which is a
companion package to `epiprocess`, offers a suite of predictive modeling tools
that can improve on some of the shortcomings of the above simple AR model.

Second, the AR forecaster here is using finalized data, meaning, it uses the
latest versions of signal values (reported COVID-19 cases) available, for both
training models and making predictions historically. However, this is not
reflective of the provisional nature of the data that it must cope with in a
true forecast task. Training and making predictions on finalized data can lead
to an overly optimistic sense of accuracy; see, for example, [McDonald et al.
(2021)](https://www.pnas.org/content/118/51/e2111453118/), and references
therein. Fortunately, the `epiprocess` package provides a data structure called
`epi_archive` that can be used to store all data revisions, and furthermore, an
`epi_archive` object knows how to slide computations in the correct
version-aware sense (for the computation at each reference time $t$, it uses
only data that would have been available as of $t$). We will revisit this
example in the [archive
vignette](https://cmu-delphi.github.io/epiprocess/articles/archive.html).

## Attribution
This document contains a dataset that is a modified part of the [COVID-19 Data Repository by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University](https://github.com/CSSEGISandData/COVID-19) as [republished in the COVIDcast Epidata API](https://cmu-delphi.github.io/delphi-epidata/api/covidcast-signals/jhu-csse.html). This data set is licensed under the terms of the [Creative Commons Attribution 4.0 International license](https://creativecommons.org/licenses/by/4.0/) by the Johns Hopkins University on behalf of its Center for Systems Science in Engineering. Copyright Johns Hopkins University 2020.

[From the COVIDcast Epidata API](https://cmu-delphi.github.io/delphi-epidata/api/covidcast-signals/jhu-csse.html):
 These signals are taken directly from the JHU CSSE [COVID-19 GitHub repository](https://github.com/CSSEGISandData/COVID-19) without changes.

