---
title: 1. Slide a computation over signal values
description: Slide a computation signal values over time.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{1. Slide a computation over signal values}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

One of the most basic tools in the `epitools` package is `slide_by_geo()`,
which is based on the family of functions provided by the 
[`slider`](https://cran.r-project.org/web/packages/slider/) package. In
`epitools`, to "slide" means to apply a computation---represented as a function 
or formula---over a trailing window of `n` days of data, grouped by `geo_value`. 
Several other functions in the `epitools` package, such as `pct_change()` and 
`estimate_deriv()`, use `slide_by_geo()` as their workhorse. 

Similar to the getting started guide, we'll fetch daily reported COVID-19 cases 
for a few U.S. states (note: new, not cumulative cases) using the
[`covidcast`](https://cmu-delphi.github.io/covidcast/covidcastR/) package, and 
then convert this to `epi_signal` format.

```{r, message = FALSE}
library(covidcast)
library(epitools)
library(dplyr)

x <- covidcast_signal(data_source = "jhu-csse", 
                      signal = "confirmed_incidence_num",
                      start_day = "2020-06-01", 
                      end_day = "2021-05-31",
                      geo_type = "state", 
                      geo_values = c("ca", "fl", "ny", "tx"))  %>%
  as.epi_signal(name = "covid19_cases") %>%
  select(value, geo_value, time_value, issue)
```

## Slide with a formula

We first demonstrate how to apply a 7-day trailing average to the daily counts
in order to smooth the signal, by using a formula in the `slide_fun` argument of
`slide_by_geo()`.

```{r}
slide_by_geo(x, slide_fun = ~ Mean(.x$value), n = 7) %>%
  head(10)
```

The formula specified via `slide_fun` has access to all columns present in the
original `epi_signal` data frame, and must refer to them with the prefix `.x$`. 
Here the function `Mean()` is a simple wrapper around `mean()` that omits `NA` 
values by default (provided by the `epitools` package). 

Notice that `slide_by_geo()` returns a data frame with a new column appended 
that contains the results (from sliding the formula), named `slide_value` by 
default. We can instead specify a name up front using the `col_name` argument: 

```{r}
slide_by_geo(x, slide_fun = ~ Mean(.x$value), n = 7, col_name = "7dav") %>%
  head(10)
```

As a simple sanity check, we visualize the 7-day trailing averages computed on
top of the original counts:

```{r, message = FALSE, warning = FALSE, fig.width = 9, fig.height = 6}
library(ggplot2)
theme_set(theme_bw())

slide_by_geo(x, slide_fun = ~ Mean(.x$value), n = 7, col_name = "7dav") %>%
  ggplot(aes(x = time_value)) +
  geom_col(aes(y = value, fill = geo_value), alpha = 0.5) +
  geom_line(aes(y = `7dav`, col = geo_value), show.legend = FALSE) +
  facet_wrap(~ geo_value, scales = "free_y") +
  scale_x_date(minor_breaks = "month", date_labels = "%b %y") +
  labs(x = "Date", y = "Reported COVID-19 cases", fill = "State")
```

## Slide with a function 

We can also pass a function for the `slide_fun` argument in `slide_by_geo()`. In 
this case, the passed function must have the following argument structure: `x`, 
a data frame the same column names as the original data frame; followed by any
number of named additional arguments; and ending with `...`, to capture general 
additional arguments. Recreating the last example of a 7-day trailing average: 

```{r}
slide_by_geo(x, slide_fun = function(x, ...) Mean(x$value), n = 7,
             col_name = "7dav") %>%
  head(10)
```

As a more complicated example, we recreate two "local" forecasters. Simple 
versions of these can be built using the [`{tsibble}` package](https://tsibble.tidyverts.org), but here we also generate prediction 
intervals in addition to the point forecasts. The first is a simple "baseline"
forecaster which propagates a flat line forward, and uses the residuals to
create quantiles. The result will be a data frame with one column containing
the desired quantiles, and the other containing the predicted values at those
quantiles.

```{r}
local_baseline <- function(
  x, ahead = 1, min_train_window = 25,
  quantiles = c(0.025, 0.100, 0.250, 0.500, 0.750, 0.900, 0.975),
  symmetrize = TRUE) 
{
  ahead <- round(ahead) # how many units of time ahead do we predict
  if (length(x) < min_train_window) # return NA forecasts if insufficient data
    return(data.frame(value = NA, quantile = c(NA, quantiles)))
  stopifnot(ahead > 0)
  s <- ifelse(symmetrize, -1, NA) # should the PIs be symmetric?
  point <- tail(x, 1) # the point forecast is the most recent value
  resids <- x - lag(x, ahead)
  resids <- c(resids, s * resids)
  q <- c(0, quantile(resids, probs = quantiles, na.rm = TRUE))
  data.frame(value = q + point, quantile = c(NA, quantiles))
}

baseline_fcasts <- sx %>% 
  slide_by_geo(slide_fun = ~ local_baseline(.x$value), 
               n = 100, col_type = "list", col_name = "value") %>%
  rename(forecast_date = time_value) %>%
  mutate(target_date = forecast_date + 1) %>% # forecast 1 day ahead
  select(-issue)

head(baseline_fcasts, 10)
```

Note the `col_type = "list"` argument is required here. To convert to a "long"
`tibble`, we can unnest the result.

```{r}
baseline_fcasts %>% rename(actual = value) %>% unnest(forecasts) %>% tail(20)
```

We can do a more complicated "local autoregressive" forecast by using trailing 
windows. Here, by default, we use 0-, 1-, and 2- week lags to forecast 1 day
ahead.

```{r}
local_ar <- function(
  x, lags = c(0, 7, 14), ahead = 1, min_train_window = 100,
  quantiles = c(0.025, 0.100, 0.250, 0.500, 0.750, 0.900, 0.975),
  symmetrize = TRUE) 
{
  ahead <- round(ahead)
  if (length(x) < min_train_window)  # return NA forecasts if insufficient data
    return(data.frame(value = NA, quantile = c(NA, quantiles)))
  stopifnot(all(lags > -1), ahead > 0, max(lags) + ahead < length(x))
  lags <- sort(lags)
  s <- ifelse(symmetrize, -1, NA)
  ma <- ahead + 1
  ml <- max(lags) + 1
  mat <- embed(x, ml + ma - 1)
  y <- mat[, ma - ahead]
  X <- mat[, lags + ma]
  fit <- lm(y ~ X)
  newx <- rev(tail(x, n = ml))[lags + 1]
  point <- drop(c(1, newx) %*% coef(fit))
  resids <- c(residuals(fit), s * residuals(fit))
  q <- c(0, quantile(resids, probs = quantiles, na.rm = TRUE))
  data.frame(value = q + point, quantile = c(NA, quantiles), row.names = NULL)
}

ar_fcasts <- x %>% 
  slide_by_geo(slide_fun = ~ local_ar(.x$value), 
               n = 100, col_type = "list", col_name = "value") %>%
  rename(forecast_date = time_value) %>%
  mutate(target_date = forecast_date + 1) %>%
  select(-issue)

ar_fcasts %>% unnest(value) %>% tail(20)
```

If we want to make forecasts using weekly data, we can apply `slide_by_geo()`
twice: first to create weekly sums, then to do the forecast.

```{r}
weekly_baseline_fcast <- x %>% 
  # sums but with imputed NAs
  slide_by_geo( ~ Mean(.x$value) * 7, n = 7, col_name = "value") %>% 
  slide_by_geo(x, slide_fun = ~ local_baseline(.x$value), 
               n = 100, col_name = "value", col_type = "list") %>%
  rename(forecast_date = time_value) %>%
  mutate(target_date = forecast_date + 1) %>%
  select(-issue) %>%
  unnest(value)

tail(weekly_baseline_fcast, 20)
```

Note that with both of these forecasting tasks, we are unfairly using "finalized"
data, the data that is currently available to us. This may not be the same as 
the data that would have been available at the `forecast_date`. See 
`?covidcast::covidcast_signal()` for some discuss on "Issue dates and revisions".
