---
title: Slide a computation over signal values
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Slide a computation over signal values}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

A central tool in the `epiprocess` package is `epi_slide()`, which is based on
the powerful functionality provided in the
[`slider`](https://cran.r-project.org/web/packages/slider) package. In
`epiprocess`, to "slide" means to apply a computation---represented as a
function or formula---over a sliding/rolling data window. The function always
applies the slide inside each group and the grouping is assumed to be across all
group keys of the `epi_df` (this is the grouping used by default if you do not
group the `epi_df` with a `group_by()`).

By default, the `.window_size` units depend on the `time_type` of the `epi_df`,
which is determined from the types in the `time_value` column of the `epi_df`.
See the "Details" in `epi_slide()` for more.

As in getting started guide, we'll fetch daily reported COVID-19 cases from CA,
FL, NY, and TX (note: here we're using new, not cumulative cases) using the
[`epidatr`](https://github.com/cmu-delphi/epidatr) package, and then convert
this to `epi_df` format.

```{r, message = FALSE, warning=FALSE}
library(epidatr)
library(epiprocess)
library(dplyr)
```

The data is fetched with the following query:

```{r, message = FALSE, eval=F}
edf <- pub_covidcast(
  source = "jhu-csse",
  signals = "confirmed_incidence_num",
  geo_type = "state",
  time_type = "day",
  geo_values = "ca,fl,ny,tx,ga,pa",
  time_values = epirange(20200301, 20211231),
) %>%
  select(geo_value, time_value, cases = value) %>%
  arrange(geo_value, time_value) %>%
  as_epi_df()
```

The data has 2,684 rows and 3 columns.

```{r, echo=FALSE}
data(jhu_csse_daily_subset)
edf <- jhu_csse_daily_subset %>%
  select(geo_value, time_value, cases) %>%
  arrange(geo_value, time_value) %>%
  as_epi_df()
```

## Optimized rolling mean and sums

For the two most common sliding operations, we offer two optimized versions:
`epi_slide_mean()` and `epi_slide_sum()`. This example gets the 7-day trailing
average of the daily cases. Note that the name of the column(s) that we want to
average is specified as the first argument of `epi_slide_mean()`.

```{r}
edf %>%
  group_by(geo_value) %>%
  epi_slide_mean("cases", .window_size = 7, na.rm = TRUE) %>%
  ungroup() %>%
  head(10)
```

Note that we passed `na.rm = TRUE` to `data.table::frollmean()` via `...` to
`epi_slide_mean`.

The following computes the 7-day trailing sum of daily cases (and passed `na.rm`
to `data.table::frollsum()` similarly):

```{r}
edf %>%
  group_by(geo_value) %>%
  epi_slide_sum("cases", .window_size = 7, na.rm = TRUE) %>%
  ungroup() %>%
  head(10)
```

## General sliding with a formula

The previous computations can also be performed using `epi_slide()`, which can
be used for more general sliding computations (but is much slower for the
specific cases of mean and sum).

The same 7-day trailing average of daily cases can be computed by passing in a
formula for the first argument of `epi_slide()`:

```{r}
edf %>%
  group_by(geo_value) %>%
  epi_slide(~ mean(.x$cases, na.rm = TRUE), .window_size = 7) %>%
  ungroup() %>%
  head(10)
```

If your formula returns a data.frame, then the columns of the data.frame
will be unpacked into the resulting `epi_df`. For example, the following
computes the 7-day trailing average of daily cases and the 7-day trailing sum of
daily cases:

```{r}
edf %>%
  group_by(geo_value) %>%
  epi_slide(
    ~ data.frame(cases_mean = mean(.x$cases, na.rm = TRUE), cases_sum = sum(.x$cases, na.rm = TRUE)),
    .window_size = 7
  ) %>%
  ungroup() %>%
  head(10)
```

Note that this formula has access to all non-grouping columns present in the
original `epi_df` object and must refer to them with the prefix `.x$...`. As we
can see, the function `epi_slide()` returns an `epi_df` object with a new column
appended that contains the results (from sliding), named `slide_value` as the
default.

Some other information is available in additional variables:

* `.group_key` is a one-row tibble containing the values of the grouping
  variables for the associated group
* `.ref_time_value` is the reference time value the time window was based on

```{r}
# Returning geo_value in the formula
edf %>%
  group_by(geo_value) %>%
  epi_slide(~ .x$geo_value[[1]], .window_size = 7) %>%
  ungroup() %>%
  head(10)

# Returning time_value in the formula
edf %>%
  group_by(geo_value) %>%
  epi_slide(~ .x$time_value[[1]], .window_size = 7) %>%
  ungroup() %>%
  head(10)
```

While the computations above do not look very useful, these can be used as
building blocks for computations that do something different depending on the
geo_value or ref_time_value.

## Slide the tidy way

Perhaps the most convenient way to setup a computation in `epi_slide()` is to
pass in an expression for tidy evaluation. In this case, we can simply define
the name of the new column directly as part of the expression, setting it equal
to a computation in which we can access any columns of `.x` by name, just as we
would in a call to `dplyr::mutate()`, or any of the `dplyr` verbs. For example:

```{r}
slide_output <- edf %>%
  group_by(geo_value) %>%
  epi_slide(cases_7dav = mean(cases, na.rm = TRUE), .window_size = 7) %>%
  ungroup() %>%
  head(10)
```

In addition to referring to individual columns by name, you can refer to
`epi_df` time window as `.x` (`.group_key` and `.ref_time_value` are still
available). Also, the tidyverse "pronouns" `.data` and `.env` can also be used
if you need distinguish between the data and environment.

As a simple sanity check, we visualize the 7-day trailing averages computed on
top of the original counts:

```{r, message = FALSE, warning = FALSE, fig.width = 9, fig.height = 6}
library(ggplot2)
theme_set(theme_bw())

ggplot(slide_output, aes(x = time_value)) +
  geom_col(aes(y = cases, fill = geo_value), alpha = 0.5, show.legend = FALSE) +
  geom_line(aes(y = cases_7dav, col = geo_value), show.legend = FALSE) +
  facet_wrap(~geo_value, scales = "free_y") +
  scale_x_date(minor_breaks = "month", date_labels = "%b %y") +
  labs(x = "Date", y = "Reported COVID-19 cases")
```

As we can see from the top right panel, it looks like Texas moved to weekly
reporting of COVID-19 cases in summer of 2021.

## Slide with a function

We can also pass a function to the second argument in `epi_slide()`. In this
case, the passed function `.f` must have the form `function(x, g, t, ...)`,
where

- "x" is an epi_df with the same column names as the archive's `DT`, minus
  the `version` column
- "g" is a one-row tibble containing the values of the grouping variables
for the associated group
- "t" is the ref_time_value for the current window
- "..." are additional arguments

Recreating the last example of a 7-day trailing average:

```{r}
edf %>%
  group_by(geo_value) %>%
  epi_slide(function(x, g, t) mean(x$cases, na.rm = TRUE), .window_size = 7) %>%
  ungroup() %>%
  head(10)
```

## Running a forecaster on a sliding window of data

The natural next step is to use the sliding window to forecast future values.
However to do this correctly, we should make sure that our data is historically
accurate. The data structure we use for that is the `epi_archive` and the
analogous slide function is `epix_slide()`. To read further along this train of
thought, see the ["Working with archives"]() vignette.

## Attribution

The `percent_cli` data is a modified part of the [COVIDcast Epidata API Doctor Visits data](https://cmu-delphi.github.io/delphi-epidata/api/covidcast-signals/doctor-visits.html). This dataset is licensed under the terms of the [Creative Commons Attribution 4.0 International license](https://creativecommons.org/licenses/by/4.0/). Copyright Delphi Research Group at Carnegie Mellon University 2020.

This document contains a dataset that is a modified part of the [COVID-19 Data Repository by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University](https://github.com/CSSEGISandData/COVID-19) as [republished in the COVIDcast Epidata API](https://cmu-delphi.github.io/delphi-epidata/api/covidcast-signals/jhu-csse.html). This data set is licensed under the terms of the [Creative Commons Attribution 4.0 International license](https://creativecommons.org/licenses/by/4.0/) by the Johns Hopkins University on behalf of its Center for Systems Science in Engineering. Copyright Johns Hopkins University 2020.

[From the COVIDcast Epidata API](https://cmu-delphi.github.io/delphi-epidata/api/covidcast-signals/jhu-csse.html):
These signals are taken directly from the JHU CSSE [COVID-19 GitHub repository](https://github.com/CSSEGISandData/COVID-19) without changes.
