---
title: Slide a computation over signal values
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Slide a computation over signal values}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

A central tool in the `epiprocess` package is `epi_slide()`, which is based on
the functionality provided in the
[`slider`](https://cran.r-project.org/web/packages/slider/) package. In
`epiprocess`, to "slide" means to apply a computation---represented as a
function or formula---over a running window of `n` time steps. Suitable
groupings can always be achieved by a preliminary call to `group_by()`.

By default, the meaning of one time step is inferred from the `time_value`
column of the `epi_df` object under consideration, based on the way this column
understands addition and subtraction. For example, if the time values are coded
as `Date` objects, then one time step is one day, since `as.Date("2022-01-01") +
1` equals `as.Date("2022-01-02")`. Alternatively, the time step can be specified
manually in the call to `epi_slide()`; you can read the documentation for more
details. Furthermore, the alignment of the running window used in `epi_slide()`
can be "right", "center", or "left"; the default is "right", and is what we use
in this vignette.

As in getting started guide, we'll fetch daily reported COVID-19 cases from CA,
FL, NY, and TX (note: here we're using new, not cumulative cases) using the
[`covidcast`](https://cmu-delphi.github.io/covidcast/covidcastR/) package, and
then convert this to `epi_df` format.

```{r, message = FALSE}
library(covidcast)
library(epiprocess)
library(dplyr)

x <- covidcast_signal(data_source = "jhu-csse", 
                      signal = "confirmed_incidence_num",
                      start_day = "2020-03-01", 
                      end_day = "2021-12-31",
                      geo_type = "state", 
                      geo_values = c("ca", "fl", "ny", "tx")) %>%
  select(geo_value, time_value, cases = value) %>%
  arrange(geo_value, time_value) %>%
  as_epi_df() 
```

## Slide with a formula

We first demonstrate how to apply a 7-day trailing average to the daily cases in
order to smooth the signal, by passing in a formula for the first argument of
`epi_slide()`. To do this computation per state, we first call `group_by()`.

```{r}
x %>% 
  group_by(geo_value) %>% 
  epi_slide(~ mean(.x$cases), n = 7) %>%
  head(10)
```

The formula specified has access to all columns present in the original `epi_df`
object (and must refer to them with the prefix `.x$`). As we can see, the
function `epi_slide()` returns an `epi_df` object with a new column appended
that contains the results (from sliding), named `slide_value` as the default. We
can of course change this post hoc, or we can instead specify a new name up
front using the `new_col_name` argument:

```{r}
x <- x %>% 
  group_by(geo_value) %>% 
  epi_slide(~ mean(.x$cases), n = 7, new_col_name = "cases_7dav")

head(x, 10)
```

## Slide with a function 

We can also pass a function for the first argument in `epi_slide()`. In this
case, the passed function must have the following argument structure: `x`, a
data frame with the same column names as the original object; followed by any
number of named arguments; and ending with `...` to capture additional
arguments. Recreating the last example of a 7-day trailing average:

```{r}
x <- x %>% 
  group_by(geo_value) %>% 
  epi_slide(function(x, ...) mean(x$cases), n = 7, new_col_name = "cases_7dav")

head(x, 10)
```

## Slide the tidy way

Perhaps the most convenient way to setup a computation in `epi_slide()` is to
pass in an expression for tidy evaluation. In this case, we can simply define
the name of the new column directly as part of the expression, setting it equal
to a computation in which we can access any columns of `x` by name, just as we
would in a call to `dplyr::mutate()`, or any of the `dplyr` verbs. For example:

```{r}
x <- x %>% 
  group_by(geo_value) %>% 
  epi_slide(cases_7dav = mean(cases), n = 7)

head(x, 10)
```

As a simple sanity check, we visualize the 7-day trailing averages computed on
top of the original counts:

```{r, message = FALSE, warning = FALSE, fig.width = 9, fig.height = 6}
library(ggplot2)
theme_set(theme_bw())

ggplot(x, aes(x = time_value)) +
  geom_col(aes(y = cases, fill = geo_value), alpha = 0.5, show.legend = FALSE) +
  geom_line(aes(y = cases_7dav, col = geo_value), show.legend = FALSE) +
  facet_wrap(~ geo_value, scales = "free_y") +
  scale_x_date(minor_breaks = "month", date_labels = "%b %y") +
  labs(x = "Date", y = "Reported COVID-19 cases")
```

As we can see from the top right panel, it looks like Texas moved to weekly 
reporting of COVID-19 cases in summer of 2021.  

## Running a local forecaster

As a more complex example, we create a forecaster based on a local (in time)
autoregression or AR model. AR models can be fit in numerous ways (using base R
functions and various packages), but here we define it "by hand" both because it
provides a more advanced example of sliding a function over an `epi_df` object,
and because it allows us to be a bit more flexible in defining a *probabilistic*
forecaster: one that outputs not just a point prediction, but a notion of
uncertainty around this. In particular, our forecaster will output a point
prediction along with an 90\% uncertainty band, represented by a predictive
quantiles at the 5\% and 95\% levels (lower and upper endpoints of the
uncertainty band).

The function defined below, `prob_ar()`, is our probabilistic AR forecaster. The
`lags`argument indicates which lags to use in the model, and `ahead` indicates
how far ahead in the future to make forecasts (both are encoded in terms of the
units of the `time_value` column; so, days, in the working `epi_df` being
considered in this vignette).

```{r}
prob_ar <- function(v, lags = c(0, 7, 14), ahead = 7, min_train_window = 20, 
                    lower_level = 0.05, upper_level = 0.95, symmetrize = TRUE,
                    nonneg = TRUE) {   
  # Return NA if insufficient training data
  if (length(v) < min_train_window + max(lags) + ahead) {
    return(data.frame(point = NA, lower = NA, upper = NA))
  }
  
  # Perform other simple checks
  stopifnot(all(lags >= 0), ahead > 0, max(lags) + ahead < length(v))
  
  # Go and fit the AR model
  y <- dplyr::lead(v, n = ahead)
  x <- do.call(cbind, purrr::map(lags, ~ dplyr::lag(v, n = .x)))
  fit <- lm(y ~ x, na.action = na.omit)

  # Make a prediction
  newx <- tail(x, 1)
  point <- drop(c(1, newx) %*% coef(fit))
  
  # Compute a band 
  r <- residuals(fit)
  s <- ifelse(symmetrize, -1, NA) # Should the residuals be symmetrized?
  q <- quantile(c(r, s * r), probs = c(lower_level, upper_level), na.rm = TRUE)
  lower <- point + q[1]
  upper <- point + q[2]
  
  # Clip at zero if we need to, then return
  if (nonneg) { 
    point = max(point, 0) 
    lower = max(lower, 0) 
    upper = max(upper, 0) 
  }
  return(data.frame(point = point, lower = lower, upper = upper))
}
```

We go ahead and slide this AR forecaster over the working `epi_df` of COVID-19 
cases. Note that we actually model the `cases_7dav` column, to operate on the 
scale of smoothed COVID-19 cases. This is clearly equivalent, up to a constant,
to modeling weekly sums of COVID-19 cases.

```{r}
x %>% 
  group_by(geo_value) %>%
  epi_slide(fc = prob_ar(cases_7dav), n = 120) %>%
  tail(10)
```

Now we get three columns `fc_point`, `fc_lower`, and `fc_upper` that correspond
to the point forecast, and the lower and upper endpoints of the 95\% prediction 
band, respectively. (If instead we had set `as_list_col = TRUE` in the call to 
`epi_slide()`, then we would have gotten a list column `fc`, where each element
is a data frame with named components `point`, `lower`, and `upper`.)

To finish off, we plot the forecasts at some times (spaced out by a few months)
over the last year, at multiple horizons: 7, 14, 21, and 28 days ahead. To do 
so, we encapsulate the process of generating forecasts into a simple function, 
so that we can call it a few times.

```{r, message = FALSE, warning = FALSE, fig.width = 9, fig.height = 6}
k_week_ahead <- function(x, ahead = 7) {
  x %>% 
    group_by(geo_value) %>%
    epi_slide(fc = prob_ar(cases_7dav, ahead = ahead), n = 120) %>%
    mutate(target_date = time_value + ahead) 
}

# First generate the forecasts, and bind them together
z <- bind_rows(k_week_ahead(x, ahead = 7),
               k_week_ahead(x, ahead = 14),
               k_week_ahead(x, ahead = 21),
               k_week_ahead(x, ahead = 28))

# Now plot some of them, spaced apart by 2 months
z_to_show <- z %>%
  filter(time_value %in% 
           seq(as.Date("2020-06-01"), as.Date("2021-12-31"), by = "2 months"))

p <- ggplot(z_to_show, aes(x = target_date, group = time_value)) +
  geom_ribbon(aes(ymin = fc_lower, ymax = fc_upper), fill = 6, alpha = 0.4) +
  geom_line(aes(y = fc_point)) + geom_point(aes(y = fc_point), size = 0.5) + 
  geom_vline(aes(xintercept = time_value), linetype = 2, alpha = 0.5) +
  facet_wrap(vars(geo_value), scales = "free_y") +
  scale_x_date(minor_breaks = "month", date_labels = "%b %y") +
  labs(x = "Date", y = "Reported COVID-19 cases")
  
gginnards::append_layers(
  p, geom_line(data = z, aes(x = time_value, y = cases_7dav), 
               inherit.aes = FALSE, color = "gray50"), pos = "bottom")
```

Two points are worth making. First, the AR model's performance here is pretty
spotty. At various points in time, we can see that its forecasts are volatile
(its point predictions are all over the place), or overconfident (its bands are
too narrow), or both at the same time. This is only meant as a simple demo and
not entirely unexpected given the way the AR model is set up. For example, the
overconfidence in this example is due to the fact that the bands are based on
quantiles of training residuals, so if the AR model fits well to the training
set, then these can clearly be too small in magnitude. The
[`epipredict`](https://cmu-delphi.github.io/epipredict/) package, which is a
companion package to `epiprocess`, offers a suite of predictive modeling tools 
that can improve on some of the shortcomings of the above simple AR model.

Second, the AR forecaster here is using finalized data, meaning, it uses the
latest versions of signal values (reported COVID-19 cases) available, for both
training models and making predictions historically. However, this is not
reflective of the provisional nature of the data that it must cope with in a
true forecast task. Training and making predictions on finalized data can lead
to an overly optimistic sense of accuracy; see, for example, [McDonald et al.
(2021)](https://www.pnas.org/content/118/51/e2111453118/), and references
therein. Fortunately, the `epiprocess` package provides a data structure called
`epi_archive` that can be used to store all data revisions, and furthermore, an
`epi_archive` object knows how to slide computations in the correct
version-aware sense (for the computation at each reference time $t$, it uses
only data that would have been available as of $t$). See the [archive
vignette](https://cmu-delphi.github.io/epiprocess/articles/archive.html) for
more details.
