---
title: "linelist_to_archive Benchmarks"
format:
  gfm:
    toc: true
    code-fold: true
---

```{r setup, include=FALSE}

#' To install mytidyfunctions, you need 
#' remotes::install_github("JavierMtzRdz/mytidyfunctions")
if (!require("pacman")) install.packages("pacman")
if (!require("devtools")) install.packages("devtools")
if (require("mytidyfunctions")) mytidyfunctions::set_mytheme() 

Sys.setlocale("LC_ALL", "en.US.UTF-8")

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  comment =  "#>"
)
```


This benchmark evaluates `linelist_to_archive()` performance.

```{r}
# Load packages
devtools::load_all("..")

pacman::p_load(tidyverse, janitor, bench, epidatr)
```

## Synthetic Linelist Data

Create linelist data of different sizes to benchmark scaling behavior.

```{r generate-data}
generate_linelist <- function(n_events, n_geos = 5, date_range_days = 365) {

  # Scenario: constant flow of cases over 50 days
  days <- seq(as.Date("2023-01-01"), by = "day", length.out = date_range_days)
  geos <- paste0("state_", sprintf("%02d", 1:n_geos))

  # Generate cases
  cases <- tibble(
    id = 1:n_events,
    geo_value = sample(geos, n_events, replace = TRUE),
    # random day in the range
    time_value = sample(days, n_events, replace = TRUE),
    # most reported quickly, some late
    lag_rec = rgeom(n_events, 0.2),
    # cases later deleted
    is_deletion = runif(n_events) < 0.02,
    lag_del = rgeom(n_events, 0.1)
  ) %>%
    mutate(
      version_recorded = time_value + lag_rec,
      # Deleted some time after recording
      version_deleted = if_else(is_deletion, version_recorded + lag_del + 1, as.Date(NA))
    )

  linelist <- cases %>%
    select(id, geo_value, time_value, version_recorded, version_deleted) %>%
    arrange(version_recorded)
  
  linelist
}
```

```{r}
# One example
set.seed(42)

linelist <- generate_linelist(1000, 1, 30)

ea <- linelist_to_archive(
    linelist,
    geo_value = geo_value,
    time_value = time_value,
    version_recorded = version_recorded,
    version_deleted = version_deleted,
    id = id,
    value = "cases"
  )

autoplot(ea)
```


## Benchmark Results

### Scaling with Dataset Size

It is tested across 1K, 10K, and 50K datasets.

```{r bench-scaling}
# Dataset
linelist_small <- generate_linelist(1000)

linelist_medium <- generate_linelist(10000)

linelist_large <- generate_linelist(50000)

inputs <- list(
  small = linelist_small,
  medium = linelist_medium,
  large = linelist_large
)

# Run benchmark
scaling_bench <- bench::press(
  size = names(inputs),
  {
    dat <- inputs[[size]]
    bench::mark(
      linelist_to_archive(
        dat,
        geo_value = geo_value,
        time_value = time_value,
        version_recorded = version_recorded,
        version_deleted = version_deleted,
        id = id
      ),
      iterations = 5,
      check = FALSE
    )
  }
)

# Summary
scaling_bench %>%
  select(size, median, mem_alloc, n_itr) %>%
  arrange(median)

```

### Cost of ID Validation

We compare `linelist_to_archive()` with different validation modes using a 50K rows dataset.

```{r bench-validation}
# Helper to convert to chart-style
convert_to_chart_style <- function(df) {
  # Rows that are recordings
  recs <- df %>%
    select(id, geo_value, time_value, version_recorded) %>%
    mutate(is_deletion = FALSE)
  
  # Rows that are deletions
  dels <- df %>%
    filter(!is.na(version_deleted)) %>%
    select(id, geo_value, time_value, version_recorded = version_deleted) %>%
    mutate(is_deletion = TRUE)
  
  bind_rows(recs, dels) %>%
    arrange(version_recorded)
}

linelist_chart <- convert_to_chart_style(linelist_large)

validation_bench <- bench::mark(
  "No Validation" = linelist_to_archive(
    linelist_large,
    geo_value = geo_value,
    time_value = time_value,
    version_recorded = version_recorded,
    version_deleted = version_deleted,
    id = NULL
  ),
  "Interval Validation" = linelist_to_archive(
    linelist_large,
    geo_value = geo_value,
    time_value = time_value,
    version_recorded = version_recorded,
    version_deleted = version_deleted,
    id = id # validation on
  ),
  "Chart-Style Validation" = linelist_to_archive(
    linelist_chart,
    geo_value = geo_value,
    time_value = time_value,
    version_recorded = version_recorded,
    is_deletion = is_deletion, # Chart-style
    id = id # validation on
  ),
  iterations = 10,
  check = FALSE
)

# Summary
validation_bench %>%
  select(expression, median, mem_alloc, n_itr)

# Plot
plot(validation_bench, type = "boxplot")
```


## Real-World Data (NHSN)

This section benchmarks using versioned NHSN data for California (CA) and Florida (FL) in 2023.

**Key Insight**: NHSN data from `pub_covidcast(issues = "*")` is already in "interval-style" format—each row is a (geo, time, version) snapshot of aggregate counts. We can pass this directly to `linelist_to_archive` treating `issue` as `version_recorded`.

```{r nhsn-bench}
states <- c("ca", "fl")
start_week <- 202301
end_week   <- 202352

nhsn_data <- pub_covidcast(
  source = "nhsn",
  signals = "confirmed_admissions_covid_ew",
  geo_type = "state",
  time_type = "week",
  geo_values = states,
  time_values = epirange(start_week, end_week),
  issues = "*" 
)

message("Fetched ", nrow(nhsn_data), " rows of versioned history.")

# Baseline: using as_epi_archive directly
archive_direct <- as_epi_archive(nhsn_data)

# Using linelist_to_archive (treating each row as interval update)
archive_linelist <- linelist_to_archive(
  nhsn_data,
  geo_value = geo_value,
  time_value = time_value,
  version_recorded = issue,
  value = "count"
)

# Compare the two approaches
cat("\n--- Direct as_epi_archive ---\n")
print(archive_direct)

cat("\n--- Via linelist_to_archive ---\n")
print(archive_linelist)

# Validation: Compare key metrics
cat("\n--- Validation ---\n")
cat("Direct archive rows:", nrow(archive_direct$DT), "\n")
cat("Linelist archive rows:", nrow(archive_linelist$DT), "\n")
cat("Direct version range:", 
    as.character(min(archive_direct$DT$version)), "to", 
    as.character(max(archive_direct$DT$version)), "\n")
cat("Linelist version range:", 
    as.character(min(archive_linelist$DT$version)), "to", 
    as.character(max(archive_linelist$DT$version)), "\n")
```

### Benchmark

```{r nhsn-bench-time}
real_bench <- bench::mark(
  `as_epi_archive` = as_epi_archive(nhsn_data),
  `linelist_to_archive` = linelist_to_archive(
    nhsn_data,
    geo_value = geo_value,
    time_value = time_value,
    version_recorded = issue,
    value = "count"
  ),
  iterations = 5,
  check = FALSE
)

# Summary
real_bench %>%
  select(expression, median, mem_alloc, n_itr)

# Plot
plot(real_bench, type = "boxplot")
```

## Session Info

```{r}
sessionInfo()
```
